---
title: "The Shape of Things to Come: The Standardized Length Index for Quantifying Relative Appendage Length"
number-sections: true
execute:
  echo: false
  message: false
  warning: false
format: docx
bibliography: ../../Suppfiles/Morphology.bib
link-citations: true
csl: ../../_extensions/quarto-journals/elsevier/bib/elsevier-harvard.csl
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
editor_options:
  chunk_output_type: console
metadata-files: 
  - ../../Suppfiles/abstract_keywords.yml
  - ../../Suppfiles/authors.yml
---

{{< pagebreak >}}

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(png)
library(grid)
library(ggpubr)
library(cowplot)
library(scales)
library(smatr)
library(MASS)
library(ggpmisc)
ggplot2::theme_set(theme_cowplot())
```

```{r}
#| label: load-lit-review-results
#| include: false

# Literature review is processed separately by Scripts/02_Lit_review.R (re-run it any time the underlying spreadsheet or classification logic changes) and saved to Derived/Rds/lit_review_results.rds. This brings Lit_review, Unclear_exclude, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/lit_review_results.rds"), envir = .GlobalEnv)
```

```{r}
#| label: Percents-in-line

Per_class <- Lr %>%
  summarize(Percent = n() / nrow(Lr) * 100, .by = class) %>%
  mutate(Percent = paste0(round(Percent, 0), "%"))

per_aves <- Per_class %>% 
  filter(class == "Aves") %>% 
  pull(Percent)

# Pooled (both classes combined) approach-usage counts/percentages, cited
# inline below and matching the combined @fig-lit-review panel.
pull_pooled <- function(approach_name, stat) {
  Lr_per_pooled %>% filter(approach == approach_name) %>% pull({{ stat }})
}

tot_approach_uses <- unique(Lr_per_pooled$Tot_studies)
n_multi_approach  <- sum(duplicated(Lr_pivot$study))
pct_mv_reg <- pull_pooled("multiple_regression", Per)
pct_raw    <- pull_pooled("raw_values", Per)
pct_other  <- pull_pooled("other", Per)
pct_ratio  <- pull_pooled("ratio", Per)
pct_resid  <- pull_pooled("residuals", Per)

# How many studies had multiple approaches? 
Num_mult_approach <- Lr_pivot %>% 
  count(study, sort = T) %>% 
  filter(n > 1) %>% 
  nrow()

# Studies examining relative (not absolute) appendage length, by whether mass
# was the body-size-control variable used.
n_relative_studies <- sum(Mass_control_tbl$n)
n_mass_control      <- Mass_control_tbl %>% filter(Mass_control == TRUE) %>% pull(n)
pct_mass_control    <- round(100 * n_mass_control / n_relative_studies, 0)

# Studies using an OLS-based approach (multiple regression or residual regression),
# as a percentage of studies examining relative (not just absolute) appendage length --
# same n_relative_studies denominator as the mass-as-body-size-control stat above.
n_ols_studies <- Lr_pivot %>%
  filter(approach %in% c("multiple_regression", "residuals")) %>%
  distinct(study) %>%
  nrow()
pct_ols_relative <- round(100 * n_ols_studies / n_relative_studies, 0)
```

```{r}
#| label: source-key-fns
#| include: false
source("Scripts/00_Key_allometry_fns.R")
```

```{r}
#| label: load-simulation-results
#| include: false

# Simulation is run separately by Scripts/01_Run_simulation.R (re-run it any time the parameter grid, sample size, or evaluation logic changes) and saved to Derived/Rds/simulation_results.rds
# Both this manuscript and supplementary_info.qmd just read the result. This brings N_spp, Parms_mat3, Parms_tbl4, Eval_tbl, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/simulation_results.rds"), envir = .GlobalEnv)
```

# Introduction

Wildlife around the world are responding to contemporary climate change with shifts in morphology. Studies have documented declines in body size consistent with Bergmann's rule across diverse taxa, and growing evidence suggests that body shape is also changing in a pattern consistent with Allen's rule [@rydingShapeshiftingChangingAnimal2021; @rydingLongShorttermResponses2024]. Because Allen's rule concerns appendage length relative to overall body size, testing it requires methods that can distinguish variation in appendage size from variation in overall body size [@symondsItsAllRelative2025]. Yet the statistical methods used to quantify relative appendage length have received far less attention than the biological hypothesis itself.

There have been three primary approaches used in empirical studies to test Allen's rule (Supplementary Figure 1). The simplest uses absolute appendage length as the response variable, regressed against temperature or spatial or temporal proxies (e.g., latitude or year), hereafter referred to as the 'gradient of interest'. Throughout this paper I focus on temperature, although the methods apply equally to any of these gradients. Using absolute appendage length has been compared with methods that explicitly account for body size elsewhere [e.g., @symondsItsAllRelative2025], so I do not consider it further. A second approach, the most common in this review, includes body size as a covariate in a multiple regression, interpreting the slope of temperature (hereafeter, $\hat{\beta}_T$) as temperature's conditional effect on appendage length after accounting for body size. A third approach takes the ratio of appendage length divided by body size as the response variable.

Despite their widespread use, both the multiple regression and ratio approaches have important statistical limitations [@glazierComplicationsBodysizeCorrection2022]. Ratio-based metrics are frequently correlated with body size itself, limiting their utility as measures of relative shape [@jakobEstimatingFitnessComparison1996; @greenMassLengthResiduals2001]. Multiple regression, on the other hand, makes a key assumption – it assigns all unexplained variation around the fitted allometric line to the $Y$ variable (appendage length). Consequently, every deviation from the fitted line is interpreted as evidence of an unexpectedly long or short appendage, rather than allowing deviations to be shared between both size variables (@nte-sma-intuition). This assumption is often difficult to justify in studies of allometric scaling [@greenMassLengthResiduals2001], and alternative approaches have been recommended [@wartonSmatr3Package2012].

Standardized major axis (SMA) regression provides an alternative line-fitting approach by treating variation in body size and appendage length symmetrically rather than assigning all unexplained variation to the response variable [@wartonBivariateLinefittingMethods2006; @nte-sma-intuition]. Consequently, SMA estimates steeper allometric slopes whenever the correlation between $X$ and $Y$ is $< 1$. SMA has long been recommended for studies of allometric scaling [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009; @greenMassLengthResiduals2001] and has been used in studies examining Allen's rule [e.g., @adamczakBiggerBodiesRequire2020], yet its performance has never been systematically compared with the classical methods used to quantify relative appendage length. Understanding the statistical behavior of these approaches, and the conditions under which they diverge, is therefore critical for understanding when and to what extent the methodological choice may influence conclusions about climate-associated morphological change.

Here I introduce the Standardized Length Index (SLI), an analogue of the Standardized Mass Index [@peigNewPerspectivesEstimating2009] but adapted to measure appendage length relative to body size. The SLI uses the SMA slope to standardize each individual's appendage length to a common reference body size, producing a size-corrected shape index in the original units of the appendage. I present two parameterisations, corresponding to two distinct questions about relative appendage change: SLI-isometry fixes the scaling exponent at 0.33 (the isometric expectation for a linear versus volumetric measurement like mass) and asks whether appendage size departs from that proportional, geometric expectation. SLI-estimated instead estimates the exponent empirically from the data and asks whether appendage size departs from the species' own observed allometric relationship. Using simulated data from `r N_spp` hypothetical species spanning a realistic range of allometric relationships and temperature sensitivities, I characterize how these six approaches diverge from one another under different combinations of allometric scaling and temperature-associated changes in body size and appendage length. I also examine whether the rank-order patterns that emerge during simulation are reproduced across three independent empirical datasets. Finally, I provide recommendations for future studies examining Allen’s rule.

::: {#nte-sma-intuition .callout-note icon="false"}
## Building intuition for SMA regression

SMA and OLS regression differ in two closely related ways. First, SMA regression estimates a steeper allometric slope whenever the correlation between variables is \< 1. Second, because it treats variation in both variables symmetrically, relative appendage length and relative body size become reciprocal descriptions of the same deviation from the allometric relationship. The equations below show why these properties arise.

*SMA estimates a steeper allometric slope*

There is a direct mathematical link between the slopes of OLS and SMA regression.\
The slope of a predictor variable $X$ (e.g., body size) is related to the response variable $Y$ (e.g., appendage length) in OLS regression via the following equation:

- $\beta_{ols} = r * \frac{Sy}{Sx}$ where $r$ is the correlation between $X$ and $Y$, and $S$ is the standard deviation

The slope of SMA regression is estimated by:

- $\beta_{sma} = sign(r) * \frac{Sy}{Sx}$ , i.e., if $r$ is negative then $\beta_{sma}$ is also negative.

A simple substitution of $\frac{Sy}{Sx}$ in the first equation shows that

$$\beta_{ols} = r * \beta_{sma}$$

This makes it evident that as $r$ increases, $\beta_{\text{OLS}} \rightarrow \beta_{\text{SMA}}$, and that the difference in estimated slopes will be highest when $X$ and $Y$ are only weakly correlated.

*SMA gives a symmetric geometric interpretation*

Unlike OLS, the SMA line of best fit is unchanged when the $X$ and $Y$ axes are flipped (@fig-sma_flip_axes), leading to a useful geometric property: swapping the axes simply reflects the points across the fitted line. Consequently, the individual highlighted in green can be described either as having relatively short appendages for its body size (panel A) or as having a relatively large body for its appendage length (panel B). These are reciprocal descriptions of the same deviation from the allometric relationship (i.e., residuals retain the same magnitude but reverse sign). In contrast, because the OLS line changes when the axes are exchanged, this reciprocity does not hold.

![In SMA regression the X and Y axes can be flipped and the line of best fit will not change. Here I flip the axes and plot the OLS (red) & SMA (blue) regression lines of best fit, as well as the same individual from my hypothetical species (highlighted in green). SMA residuals are equally correlated with both $X$ and $Y$ ($|r_{sma}| = 0.59$, whereas OLS residuals remain uncorrelated with the predictor but strongly correlated with the response ($r_{ols} = 0.96$).](../../Figures/sma_flip_axes.png){#fig-sma_flip_axes width="100%"}
:::

# Literature review

I reviewed empirical studies published between 2021 and 2025 that quantified appendage length in the context of Allen's rule in endotherms. Studies were identified using the list of studies published in @skinnerBergmannAllenUnified2026 and subset to include only those that examined Allen's rule, leaving `r nrow(Lit_review)` studies. Of these, `r nrow(Unclear_exclude)` were further excluded due to unclear methodologies, leaving `r nrow(Lr)` studies. `r n_multi_approach` of these studies used multiple approaches to examine Allen's rule, resulting in `r tot_approach_uses` analytical approaches that were classified into five categories: ratio-based methods, two-step residual regression, multiple regression with body size as a covariate, use of absolute appendage length only, or other. 'Other' included shape axes from principal component analyses, path analyses, and geometric morphometric approaches, among other creative solutions [e.g., @youngfleshAbioticConditionsShape2022]. Multiple regression was the most common approach (`r pct_mv_reg`%), followed by absolute appendage length (`r pct_raw`%), other approaches (`r pct_other`%), ratio-based methods (`r pct_ratio`%), and residual regression (`r pct_resid`%; Supplementary Figure 1). Of the studies examining appendage length *relative* to body size, `r n_mass_control` / `r n_relative_studies` (`r pct_mass_control`%) used mass as the variable representing body size. The majority of studies (`r per_aves`) examined Allen's rule in birds. The remainder used mammal focal taxa.

# Materials and Methods

## Data simulation

I simulated appendage length, mass, and a temperature increase for `r N_spp` hypothetical species, each represented by `r N_ind` individuals. I simulated mass measurements as a proxy for body size as it was the most frequent variable used to control for body size in studies of Allen’s rule. Data were drawn from a multivariate normal distribution:

$$\begin{bmatrix}
\log(\text{Appendage}) \\
\log(\text{Mass}) \\
\text{Temperature}
\end{bmatrix}
\sim \mathrm{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

with $\boldsymbol{\mu} = [\log(180),\ \log(80),\ 1.0]^\top$ representing a species with a mean appendage length of 180 mm and mean mass of 80 g, experiencing an average 1°C temperature increase. I simulated morphological and environmental data from a joint multivariate normal distribution rather than from a directional model (e.g., $Appendage \sim Mass + Temperature + \epsilon$). This approach does not designate any variable as a response or predictor in the data-generating process, allowing characterization of each statistical approach against the same underlying joint distribution without presupposing a particular regression structure. The covariance matrix $\boldsymbol{\Sigma}$ was constructed from three Pearson correlations ($\rho_{12}$: appendage–mass; $\rho_{13}$: appendage–temperature; $\rho_{23}$: mass–temperature) and the allometric scaling slope. Hypothetical species were constructed by crossing the following parameter values: appendage–mass correlation $\rho_{12} \in {`r paste(r_12, collapse = ", ")`}$, temperature–morphology correlations $\rho_{13}, \rho_{23} \in {`r paste(r_13, collapse = ", ")`}$, crossed factorially to produce species with different relative temperature associations of mass and appendage, and allometric scaling slope $b \in {`r paste(b_avg_12, collapse = ", ")`}$, representing hypoallometry, isometry, and hyperallometry respectively. Standard deviations of log-mass and log-appendage were set so that the average of the OLS and SMA slopes equalled the specified allometric slope, allowing me to vary allometric scaling independently of the correlation structure.

## Effect classification

Because species were simulated from a covariance structure rather than from a causal model, there is no single fitted regression coefficient representing a universal "true" effect against which every method's estimate is either correct or incorrect. Instead, I define an isometrically-defined reference quantity, $\beta_T$, that captures departure from proportional (isometric) scaling between appendage length and body size:

$$\beta_T = \rho_{13} \cdot \beta_{SMA} - 0.33 \cdot \rho_{23}, \qquad \beta_{SMA} = \frac{2b}{\rho_{12}+1}$$ {#eq-b-true}

where $b$ is the simulated allometric scaling slope, and $\beta_{SMA}$ is a species' population SMA slope implied by its simulated parameters. Importantly, this is only one theoretically motivated definition of relative appendage change, and alternative definitions are possible (discussed further in Section 'Statistical approach classification'). Defining $\beta_T$ in terms of both allometric scaling and the strength of temperature-morphology relationships makes it clear that there are two ways a species could depart from $\beta_T = 0$, the isometric reference: (1) temperature having a differential association with mass vs. appendage ($\rho_{13} \neq \rho_{23}$), or (2) the species exhibiting non-isometric scaling ($\beta_{SMA} \neq 0.33$). Rewriting $\rho_{13} = \rho_{23} + \Delta\rho$, where $\Delta\rho$ is the difference between $\rho_{13}$ and $\rho_{23}$, decomposes @eq-b-true into these same two distinct components, expressed additively within the isometric-relative reference quantity itself:

$$\beta_T \;\propto\; \underbrace{\rho_{23}\left(\beta_{SMA} - 0.33\right)}_{\text{allometry-vs-isometry component}} \;+\; \underbrace{\beta_{SMA} \cdot \Delta\rho}_{\substack{\text{differential temperature-} \\ \text{association component}}}$$

As simulated temperature increased, species with $\beta_T > 0$ were classified as getting relatively longer and $\beta_T < 0$ species were classified as getting relatively fatter. Fatter and Longer are both realizations of Bergmann's rule, as warming always decreased overall body size in these two categories.

I additionally simulated a set of species with $\beta_T = 0$ exactly, classified as proportionally decreasing in size ("Proportionally smaller"). I solved for $\rho_{13}$ algebraically based on a species $\rho_{23}$ and $\beta_{SMA}$, following directly from @eq-b-true:

$$\rho_{13} = \frac{0.33\ \cdot \rho_{23}}{\beta_{SMA}}$$

guaranteeing $\beta_T = 0$ exactly, regardless of allometric scaling category.

Finally, I simulated species that increased in size with warmer temperatures, in line with several recent studies that have documented inverse Bergmannian clines [@stapelfeldtLongtermFieldStudy2023; @castillo-figueroaDoesBergmannsRule2022; @jumanRecentRapidEcogeographical2022]. The 'Proportionally larger' scenario was the mirror image of the 'Proportionally smaller' category, solving $\rho_{13}$ the same way from positive $\rho_{23} \in \{0.15, 0.3, 0.5, 0.7\}$, $b \in \{0.22, 0.33, 0.44\}$, and $\rho_{12} \in \{0.3, 0.45, 0.6\}$ (as above), representing species that grow larger with warming while maintaining the isometric-relative reference quantity at zero.

## Statistical approaches

I applied six methods from three families to each simulated species (@tbl-methods-estimands).

**Ratio approaches: (**1) The appendage-to-mass ratio $log(A / S)$ was regressed on temperature. (2) The squared appendage-to-mass ratio $log(A^2 / S)$ was regressed on temperature, reflecting a surface-area-to-volume formulation used in some empirical studies [e.g., @nwaoguTemperatureAridityDetermine2018; @dubinerWidespreadRecentChanges2022].

**Ordinary least squares (OLS) regression approaches: (**1) A multiple regression approach where appendage was modelled as a linear function of mass and temperature simultaneously. The temperature coefficient $\beta_T$ was interpreted as temperature's conditional effect on appendage length, holding mass constant. (2) A two-step residual regression approach was conducted where appendage was regressed on mass using OLS, and the resulting residuals were regressed on temperature.

**Standardized Length Index — estimated (SLI-estimated):** Following @peigNewPerspectivesEstimating2009, I computed:

$$SLI_i = A_i \cdot \left(\frac{M_0}{M_i}\right)^{b_{SLI}}$$

where $A_i$ is the appendage length of individual $i$, $M_i$ is the mass of individual $i$, $M_0$ is the mean mass of the sampled population, and $b_{SLI}$ comes from the empirical SMA slope of $\log(A)$ on $\log(S)$ estimated from the data. The SLI reprojects each individual's appendage length to the value it would take at the population mean body size (@fig-sli-concept, panel a). $SLI_i$ is then regressed on temperature.

**Standardized Length Index — isometry (SLI-isometry):** Identical to SLI-estimated but replacing the estimated $b_{SLI}$ by $b_{SLI} = 0.33$, the isometric scaling exponent for a linear versus volumetric measurement.

Although the present study focuses on appendage length relative to body mass, the SLI framework can also be applied when both the body-size measure and the appendage measure are linear dimensions (e.g., body length and wing length). In this case, the isometric reference exponent is $b_{SLI} = 1$, because both measurements scale linearly with length rather than volumetrically. This formulation may be particularly useful when body condition is expected to vary systematically with temperature or another environmental gradient.

Because SLI-isometry standardizes against the fixed isometric exponent while SLI-estimated standardizes against each species' own SMA slope, the two parameterizations are structurally capable of detecting different components of the isometric-relative reference quantity defined above (@fig-sli-concept, panel b). SLI-isometry's exponent ($b_{SLI} = 0.33$) generally differs from a species' true SMA slope, so it remains sensitive to both the allometry-vs-isometry component and the differential temperature-association component. SLI-estimated's exponent is by definition the species' own SMA slope, so the allometry-vs-isometry component vanishes, leaving only the differential temperature-association component. This decomposition is algebraic rather than causal - the two components describe distinct contributions to the isometric-relative quantity but do not imply distinct causal pathways.

```{r}
#| label: fig-sli-concept
#| fig-cap: "The Standardized Length Index (SLI), adapted from the Scaled Mass Index [@peigNewPerspectivesEstimating2009]. (a) The SLI reprojects each individual's appendage length to the value it would take at the population's mean mass, $M_0$. Two example individuals (filled black circles) are shown sliding along their own standardization curves (dashed black line) to the population’s mean mass to their standardized values (open points). The standardization curves are parallel to the population-level slope (solid grey curve) once linearized. (b) SLI-estimated and SLI-isometry estimate two different quantities. Points are a representative subset of individuals (log-log scale) for a simulated hyperallometric species whose appendage length increases with warming, coloured by simulated temperature increase. The white-filled point marks the model-predicted phenotype for an individual experiencing a 0.5°C temperature increase. The solid blue line ('Estimated allometry') is the empirical SMA relationship ($b_{SMA} = 0.68$), which provides the reference relationship for SLI-estimated; the dashed grey line ('Isometric reference') is the isometric reference ($b_{SMA} = 0.33$), which provides the reference relationship for SLI-isometry. The green arrow shows the allometry-isometry component: the gap between the two reference lines at this individual's predicted body size. The red arrow shows the temperature-association component: the additional displacement associated with temperature, measured from the isometric line to this individual's predicted appendage length."
#| fig-pos: 'H'
#| fig-width: 13
#| fig-height: 5.7

# fig-width matches the ggsave() dimensions in Scripts/04a_SLI_concept_figure.R exactly, so
# the R graphics device's aspect ratio matches the source PNG's and draw_grob() below doesn't
# stretch it -- LaTeX still scales the final image down to fit the text column width.
# Both panels are built in Scripts/04a_SLI_concept_figure.R (re-run it any time either panel's
# code changes) and saved to Figures/SLI_concept.png; that script also exports the panel-(b)
# example species to Derived/Rds/Ex_df_hypo_hyper.rds for fig-ols-sma (Discussion) to reuse.
sli_concept_img  <- png::readPNG("Figures/SLI_concept.png")
sli_concept_grob <- grid::rasterGrob(sli_concept_img, width = grid::unit(1, "npc"),
                                      height = grid::unit(1, "npc"))
cowplot::ggdraw() + cowplot::draw_grob(sli_concept_grob)
```

I log-transformed all variables and then standardised to unit variance before fitting regression models so that coefficients were comparable across species and methods.

```{r}
#| label: tbl-methods-estimands
#| tbl-cap: "The six statistical approaches compared in this study, how each accounts for body size, the assumed relationship between mass and appendage length, and the quantity represented by the resulting measure. 'Directional' denotes the direction assumed by the statistical model, not a causal claim; a causal interpretation of that direction requires additional assumptions (main text)."

Methods_estimands_tbl <- tibble::tribble(
  ~Method,                    ~`Body-size relationship`,             ~`Mass–appendage relationship`,          ~`Quantity represented`,
  "Ratio ($A/S$)",            "$b = 1$ (fixed)",                      "Proportional relationship",        "Ratio-defined relative size",
  "Squared ratio ($A^2/S$)",  "$b = 0.5$ (fixed)",                    "Geometric relationship",           "Surface-to-volume-inspired relative size",
  "OLS multiple regression",  "$Appendage \\sim Mass$",       "Directional (statistical model): $Mass \\rightarrow Appendage$", "Temperature effect on appendage, conditional on mass",
  "OLS residuals",            "$Appendage \\sim Mass$",       "Directional (statistical model): $Mass \\rightarrow Appendage$", "Residual appendage size after OLS adjustment",
  "SLI-isometry",              "$b = 0.33$ (fixed)",                   "Relative to isometry",            "Appendage size relative to isometric scaling",
  "SLI-estimated",             "Species-specific SMA slope",          "Symmetric: $Mass \\leftrightarrow Appendage$", "Appendage size relative to species-specific allometry"
)

knitr::kable(Methods_estimands_tbl, format = "latex", align = "c", booktabs = TRUE, escape = FALSE, linesep = "") %>%
  kableExtra::column_spec(1, width = "2.6cm", latex_valign = "m") %>%
  kableExtra::column_spec(2, width = "3.1cm", latex_valign = "m") %>%
  kableExtra::column_spec(3, width = "3.4cm", latex_valign = "m") %>%
  kableExtra::column_spec(4, width = "4.2cm", latex_valign = "m") %>%
  kableExtra::kable_styling(font_size = 10)
```

## Statistical approach classification

I evaluated the six statistical approaches using two criteria relative to the isometrically defined benchmark quantity $\beta_T$. First, I assessed whether each approach agreed with the sign of $\beta_T$: a positive $\hat{\beta}_T$ when $\beta_T > 0$, a negative $\hat{\beta}_T$ when $\beta_T < 0$, and $\beta_T \approx 0$ when $\beta_T = 0$. I considered $\hat{\beta}_T$ to agree with the benchmark when its 95% confidence interval excluded zero in cases where $\beta_T \neq 0$, or included zero when $\beta_T = 0$. Second, I assessed how closely $\hat{\beta}_T$ tracked $\beta_T$ in magnitude using a bivariate regression of $\hat{\beta}_T$ on $\beta_T$ across species. An approach that reproduced the benchmark exactly would yield no sign misclassifications, $r = 1$, and estimates falling on the 1:1 line.

Importantly, $\beta_T$ is not a universal ground truth against which the six approaches can be ranked. I defined $\beta_T$ using the SMA scaling relationship anchored to an exponent of 0.33, reflecting the theoretical expectation that $A \propto M^{1/3}$ [@shingletonAllometryStudyBiological2010] and the same isometric anchor used by SLI-isometry. A benchmark defined using the OLS scaling relationship, or another scaling exponent, would instead define a different quantity and could consequently favour different approaches (Supplementary Figure 7). The purpose of this comparison is therefore not to identify the universally most accurate method, but to characterize how the six approaches behave relative to a clearly defined isometric quantity.

Because SLI-isometry uses the same isometric anchor as $\beta_T$, close agreement between the two is expected by construction. Indeed, the SLI-isometry coefficient and $\beta_T$ differ only by a known scaling factor, consisting of a variance-correction term derived from the same three simulation parameters used to define $\beta_T$. Dividing the SLI-isometry coefficient by this factor produces an exact correspondence with $\beta_T$ (Supplementary Figure 8), demonstrating that the two quantities are algebraically equivalent up to rescaling. This correspondence provides an algebraic check on the implementation of the isometric benchmark and SLI-isometry calculation, but does not demonstrate that SLI-isometry is superior to the other approaches. Rather, the informative comparisons concern how the remaining approaches systematically diverge from the isometrically defined benchmark.

## Empirical case studies

I applied all six methods to three independent empirical datasets to assess whether the rank-order patterns observed in simulation were also observed in real data. First, I used morphological data from three migratory nightjar species, European nightjars (*Caprimulgus europaeus*), eastern whip-poor-wills (*Antrostomus vociferus*), and common nighthawks (*Chordeiles minor*), spanning a spatial temperature gradient [@skinnerEnvironmentalGeographicConditions2025]. Second, I applied the six methods to a dataset comprising 43 North American migrant species (1979–2016), using capture year as the gradient of interest [@weeksSharedMorphologicalConsequences2020]. Third, I applied the methods to a dataset of 49 Atlantic Forest bird species using mean annual temperature as the gradient of interest [@rodriguesATLANTICBIRDTRAITS2019]. Together, these datasets span different taxonomic groups, ecological systems, gradients of interest, and putative selective pressures, providing a broad test of whether the same rank-order patterns are observed across empirical contexts.

Age and sex can influence body size independently of the gradient of interest (e.g., through sexual size dimorphism or age-related growth). For each species, I therefore tested whether age or sex predicted mass or wing length while controlling for the gradient variable (i.e., temperature or year), retaining a covariate for further consideration only when it significantly predicted body size ($p < 0.10$).

Retained covariates were incorporated differently by method. For the ratio ($A/S$) and squared-ratio ($A^2/S$) I assume that age and sex influence wing and mass roughly proportionally, and thus would not importantly modify the ratios. I do not include age and sex in the SLI-isometry approach because $b_{SLI}$ is fixed at 0.33 by definition. For the OLS-residual approach, age and sex were included in the first-stage regression of wing on mass; for multiple regression, they were included alongside mass and the gradient variable. In both cases, a covariate entered the model only when it had at least two groups that each met a minimum sample size (100 individuals). Individuals with unrecorded age or sex were retained as their own category in both regressions rather than dropped. All six approaches were fit to the same individuals within a species, differing only in how age and sex information was incorporated.

For SLI-estimated, each individual's SMA slope was estimated using the most specific reliable grouping available. Importantly, SMA regression is unreliable when the correlation between appendage and body mass is weak [@smithUseMisuseReduced2009]. Thus, SLI-estimated was calculated only for species with reliable mass–wing correlation (i.e., $r \geq 0.3$ and $p < 0.05$). Where possible, I estimated per-group SMA slopes, starting with the: (1) age-by-sex combination, when it contained at least 50 individuals and had a reliable mass–wing correlation; (2) age and/or sex class considered separately (each pooling across the other covariate, requiring at least 100 individuals per group as well as a reliable correlation); or (3) otherwise, the species-wide SMA slope. In (2), if age and sex were both able to be estimated they were averaged together (i.e., a hatch-year male would receive $(b_{HY} + b_M)/2$). Individuals with unrecorded age or sex were treated as their own class and passed through this same hierarchical slope-estimation procedure.

To evaluate whether the same rank-order patterns were observed in empirical data, I averaged the gradient coefficients ($\hat\beta_T$) within each method family (Ratio: $A / S$ and $A^2 / S$; OLS: multiple regression and OLS-residual; SLI: SLI-isometry and SLI-estimated) for each species and compared their rank order with the simulation results. While this provides only a coarse assessment of rank-order correspondence, my objective was to identify broad patterns across method families to assess qualitative alignment between simulations and empirical data. Because the expected ordering depends on the direction of body size change (Supplementary Figure 6), species were first classified according to whether they exhibited Bergmannian, inverse Bergmannian, stable, or mixed size trends (@tbl-classification). Bergmannian species exhibited declines in wing length, body mass, or both along the gradient, whereas inverse Bergmannian species exhibited increases in one or both traits ($p < 0.05$). Stable species showed no significant change in either trait, while mixed species exhibited opposing trends in wing length and body mass. Only Bergmannian and inverse Bergmannian species were used to evaluate agreement with the simulation results because these represent the situations where relative appendage length is particularly sensitive to the method used to account for body size. Species with stable or mixed size trends are reported in Supplementary Figures 11 and 12.

Past authors have noted that ratio methods can remain strongly correlated with body size when the exponent imposed by the ratio does not match the underlying scaling relationship [@greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996]. To test this, and to assess whether SLI's use of an explicit allometric exponent reduces this dependence relative to an implicit ratio, I calculated the Pearson correlation between body mass and each of the ratio metrics, SLI-isometry, and SLI-estimated for every species in the three empirical datasets.

```{r}
#| label: tbl-classification
#| tbl-cap: "Classification of species according to significant ($p < 0.05$) changes in wing length and body mass along the gradient of interest (temperature or year)."

Classification_tbl <- tibble::tribble(
  ~Category,               ~`Classification criterion`,
  "Bergmann's",            "Wing length, body mass, or both decrease",
  "Inverse Bergmann's",    "Wing length, body mass, or both increase",
  "Mixed - Wingyier",      "Wing length increases while body mass decreases",
  "Mixed - Fatter",        "Wing length decreases while body mass increases",
  "Stable",                "No significant change in either trait"
)
knitr::kable(Classification_tbl)
```

# Results

## Simulation performance

```{r}
#| label: plot-approaches-fun
#| include: false

plot_approaches <- function(df, x_txt_size = 9, legend.pos = "top") {
  df <- df %>%
    filter(Temp_eff != "Proportionally larger") %>%
    mutate(Model = str_replace_all(Model, x_labs))

  ggplot(df, aes(x = Model, y = b_temp_inc)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, height = 0, alpha = .5, size = 1.9,
                aes(color = Strength, shape = Scaling)) +
    facet_wrap(vars(Temp_eff)) +
    labs(x = NULL, y = expression(hat(beta)[T] ~ "on relative appendage length"),
         color = "Strength", shape = "Scaling") +
    guides(shape = guide_legend(nrow = 1)) +
    theme(
      axis.text.x = element_text(size = x_txt_size, vjust = .58, angle = 55),
      legend.position = legend.pos, legend.box = "vertical"
    )
}
```

The six methods produced strikingly different estimates of temperature's effect on relative appendage length, and the pattern was consistent across allometric scaling categories and correlation structures (@fig-compare-approaches).

```{r}
#| label: fig-compare-approaches
#| fig-cap: "Estimated effect of temperature on relative appendage length ($\\hat{\\beta}_T$) for each method across three different realizations of Bergmann's rule (Fatter, Proportionally smaller, and Longer species). Positive values of $\\hat{\\beta}_T$ indicate that species were estimated to become relatively longer with increasing temperature. Negative values indicate they become relatively fatter. Each point is a simulated species. Point shape indicates allometric scaling category and colour indicates the strength of the difference between temperature's effect on appendage length and mass."
#| fig-width: 9
#| fig-height: 4.5
#| fig-pos: 'H'

plot_approaches(Parms_tbl4, x_txt_size = 11)
```

**Ratio approaches** tended to diverge from the isometric-relative reference quantity in the positive direction, although this was much more prominent when $A/S$ was used as the response variable. When simulated to become relatively fatter, $A^2/S$ disagreed with the reference for only `r Ratio2_fatter_wrong_n` of `r Fatter_n` hypothetical species, whereas $A/S$ disagreed for roughly 1/4 of species (@fig-eval). Among species simulated as Proportionally smaller, where $\beta_T = 0$, no species had 95% confidence intervals containing $0$ for either $A/S$ or $A^2/S$.

**Multiple regression and OLS residuals** approaches consistently diverged from the isometric-relative reference quantity in the negative direction (@fig-eval). Both methods returned negative $\hat{\beta}_T$ in the majority of species simulated to become relatively longer, agreed with the reference for Fatter species, and overcorrected in Proportionally smaller species: only `r Ryding_proportional_null_correct`% of these species had a 95% confidence interval that included $0$ for either method, showing statistically significant departures from the null isometric-relative reference quantity in the large majority of null-effect species. The two approaches behaved nearly identically, consistent with their mathematical equivalence when predictor variables are uncorrelated.

**SLI-isometry** tracked the isometric-relative reference quantity closely, as expected since both are anchored to the same isometric scaling exponent (@fig-eval). Its 95% confidence interval excluded zero on the correct side (matching the reference's sign) for `r Sli.iso_longer_right`% of species simulated to become relatively longer, and `r Sli.iso_fatter_right`% of species simulated to become relatively fatter. Among species simulated as Proportionally smaller, `r Sli.iso_proportional_null_correct`% had a 95% confidence interval that included zero. **SLI-estimated** agreed with the isometric reference quantity in the Longer and Fatter categories (`r Sli.est_longer_right`% and `r Sli.est_fatter_right`% agreement, respectively; @fig-eval), but diverged sharply among Proportionally smaller species, where only `r Sli.est_proportional_null_correct`% of species had a 95% confidence interval that included zero.

Results for an additional **Proportionally larger** scenario are presented in the Supplementary Information, but in short these hypothetical species showed the inverse pattern as the Proportionally smaller species.

```{r}
#| label: fig-eval
#| fig-cap: "Each method's estimated coefficient against each species' own analytically-derived isometric-relative reference quantity, one point per simulated species. Colour indicates shape-response category and point shape indicates allometric scaling category. Grey lines mark zero on each axis, the classification boundary. Proportionally smaller species sit exactly on the vertical zero line by construction. r (Pearson correlation with the reference quantity) and the percentage of species in agreement, pooled across all three categories shown in each panel, are annotated in each panel. A species counts as in agreement if its 95% confidence interval falls on the correct side of zero (Longer, Fatter), or if that interval includes zero (Proportionally smaller)."
#| fig-width: 9
#| fig-height: 6.5
#| fig-pos: 'H'

Eval_scatter_df <- Parms_tbl4 %>%
  filter(Temp_eff != "Proportionally larger") %>%
  mutate(Model = str_replace_all(Model, x_labs))

# Includes Longer, Fatter, and Proportionally smaller -- "correct" uses Est_correct (01_Run_simulation.R,
# Evaluation section) throughout: whole-CI-on-the-correct-side for Longer/Fatter, CI-includes-zero
# (reversed) for Proportionally smaller, since its true effect is exactly 0 by construction.
Eval_scatter_stats <- Eval_scatter_df %>%
  group_by(Model) %>%
  summarize(r = round(cor(beta_iso_true, b_temp_inc), 2),
            pct_correct = round(mean(Est_correct) * 100, 1),
            .groups = "drop")

ggplot(Eval_scatter_df, aes(x = beta_iso_true, y = b_temp_inc, color = Temp_eff, shape = Scaling)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_point(alpha = .55, size = 1.2) +
  geom_text(data = Eval_scatter_stats, aes(label = paste0("r=", r, ", ", pct_correct, "% in agreement")),
            x = -Inf, y = Inf, hjust = -0.05, vjust = 1.3, inherit.aes = FALSE, size = 2.6) +
  facet_wrap(vars(Model), scales = "free_y") +
  labs(x = expression(beta[T] ~ "(isometric-relative reference, per species)"),
       y = expression(hat(beta)[T] ~ "(estimated)"),
       color = "Category", shape = "Scaling") +
  guides(color = guide_legend(nrow = 1), shape = guide_legend(nrow = 1)) +
  theme(legend.position = "top", legend.box = "vertical")
```

## Empirical case studies

```{r}
#| label: tbl-species-accounting
#| tbl-cap: "Number of empirical species classified into each size / shape-shifting category (@tbl-classification), by dataset. Species exhibiting Bergmann's and Inverse Bergmann's rules are shown in the main text (@fig-empirical)."

dir_order <- c("Bergmann's", "Inverse Bergmann's", "Mixed - Wingier", "Mixed - Fatter", "Stable")

Species_accounting_tbl <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(Study, species, Direction) %>%
  count(Study, Direction) %>%
  pivot_wider(names_from = Direction, values_from = n, values_fill = 0) %>%
  mutate(Study = case_match(Study,
                             "Nightjar" ~ "Skinner et al (2025)",
                             "Weeks (2020)" ~ "Weeks et al (2020)",
                             "Atlantic birds" ~ "Rodrigues et al (2019)"),
         Study = factor(Study, levels = c("Skinner et al (2025)", "Weeks et al (2020)", "Rodrigues et al (2019)"))) %>%
  arrange(Study) %>%
  mutate(Study = as.character(Study)) %>%
  dplyr::select(Study, any_of(dir_order)) %>%
  mutate(Total = rowSums(across(any_of(dir_order))))

n_species_total <- sum(Species_accounting_tbl$Total)

Species_accounting_tbl <- bind_rows(
  Species_accounting_tbl,
  Species_accounting_tbl %>%
    summarise(Study = "Total", across(where(is.numeric), sum))
)

Dir_totals_row        <- Species_accounting_tbl %>% filter(Study == "Total")
pct_bergmanns_all     <- round(100 * Dir_totals_row$`Bergmann's` / n_species_total)
pct_inv_bergmanns_all <- round(100 * Dir_totals_row$`Inverse Bergmann's` / n_species_total)
pct_mixed_wingier_all <- round(100 * Dir_totals_row$`Mixed - Wingier` / n_species_total)
n_mixed_fatter_all    <- Dir_totals_row$`Mixed - Fatter`
pct_stable_all        <- round(100 * Dir_totals_row$Stable / n_species_total)

knitr::kable(Species_accounting_tbl,
             col.names = c("Study", "Bergmann's", "Inverse Bergmann's",
                           "Mixed - Wingier", "Mixed - Fatter", "Stable", "Total"),
             align = "c")
```

Across the `r n_species_total` species examined here, most species (`r pct_bergmanns_all`%) adhered to Bergmann's rule in at least one body size metric, whereas `r pct_inv_bergmanns_all`% exhibited a pattern more consistent with Inverse Bergmann's rule. `r pct_mixed_wingier_all`% of species increased in wing length while simultaneously getting lighter, while only `r n_mixed_fatter_all` species got heavier and simultaneously decreased wing length. `r pct_stable_all`% of species showed no significant response in either mass or wing (@tbl-classification; @tbl-species-accounting).

```{r}
#| label: rank-consistency-empirical
#| include: false

Rank_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(species, Study, Direction, rank_consistent) %>%
  filter(Direction %in% c("Bergmann's", "Inverse Bergmann's"))

Rank_summary_dir <- Rank_all %>%
  group_by(Direction) %>%
  summarize(n = n(), n_consistent = sum(rank_consistent),
            pct = round(100 * mean(rank_consistent)), .groups = "drop")

n_rank_total     <- nrow(Rank_all)
pct_rank_overall <- round(100 * mean(Rank_all$rank_consistent))

pull_rank_stat <- function(dir, col) {
  Rank_summary_dir %>% filter(Direction == dir) %>% pull({{ col }})
}

pct_rank_bergmanns     <- pull_rank_stat("Bergmann's", pct)
n_rank_bergmanns       <- pull_rank_stat("Bergmann's", n)
pct_rank_inv_bergmanns <- pull_rank_stat("Inverse Bergmann's", pct)
n_rank_inv_bergmanns   <- pull_rank_stat("Inverse Bergmann's", n)

fig_cap_empirical <- glue::glue("Estimated effect ($\\hat{{\\beta}}$) of the gradient of interest on relative wing length for each of the six methods, across three empirical avian datasets (Nightjars: red points; North American migrants: blue points; Atlantic Forest birds: green points), grouped by direction of body size change: (a) Bergmann's ($n$ = {n_rank_bergmanns} species) and (b) Inverse Bergmann's ($n$ = {n_rank_inv_bergmanns} species). Only species with a significant ($p < 0.05$) mass or wing response are shown. Each point is one species' estimate ($\\hat{{\\beta}}$) for a given method; boxplots summarize the distribution of these estimates across species within each method. The y-axis label matches Figure 3's to indicate that both figures show the same quantity, temperature's estimated effect on relative appendage length. A positive value indicates birds become relatively longer with increasing temperature (or through time); a negative value indicates they become relatively fatter. The percentage in the top-right corner of each panel is the proportion of species whose rank order across methods matched the simulation-predicted order (Bergmann's: Ratio > SLI > OLS; Inverse Bergmann's: Ratio < SLI < OLS).")
```

The rank-order pattern from simulations was generally reproduced in all three datasets: ratio methods consistently produced the most positive estimates and multiple regression / OLS residuals the most negative, with SLI falling between these extremes. `r pct_rank_overall`% of the `r n_rank_total` species with a directional (Bergmann's or Inverse Bergmann's) classification showed the rank order predicted by the simulations (`r pct_rank_bergmanns`% of `r n_rank_bergmanns` Bergmann's species and `r pct_rank_inv_bergmanns`% of `r n_rank_inv_bergmanns` Inverse Bergmann's species, @fig-empirical). This pattern also held at the level of individual methods with a fixed implicit scaling exponent, rather than only after averaging within method families (Supplementary Information).

```{r}
#| label: fig-empirical
#| fig-cap: !expr fig_cap_empirical
#| fig-pos: 'H'
#| fig-height: 6

emp_img  <- png::readPNG("Figures/Empirical_combined.png")
emp_grob <- grid::rasterGrob(emp_img, width = grid::unit(1, "npc"),
                              height = grid::unit(1, "npc"))
cowplot::ggdraw() + cowplot::draw_grob(emp_grob)
```

### Correlation between the wing/mass ratio and body mass

```{r}
#| label: ratio-mass-correlation-empirical
#| include: false

Ratio_mass_cor_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_mass_cor.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_mass_cor.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_mass_cor.csv",  show_col_types = FALSE)
) %>%
  dplyr::select(Study, species, Metric, n, r, p_value)

# One row per Study x Metric (Ratio, Ratio2, Sli_iso, Sli_est), median r only --
# pivoted to one column per metric for display (tbl-ratio-mass-summary).
Ratio_mass_summary_tbl <- Ratio_mass_cor_all %>%
  summarize(N_species = n(), Median_r = median(r), .by = c(Study, Metric)) %>%
  mutate(Median_r = round(Median_r, 2),
         Metric = factor(Metric, levels = c("Ratio", "Ratio2", "Sli_iso", "Sli_est"))) %>%
  arrange(Study, Metric)

Ratio_mass_species_tbl <- Ratio_mass_cor_all %>%
  arrange(Study, species, Metric) %>%
  mutate(r = round(r, 2),
         p_value = ifelse(p_value < 0.001, "< 0.001", as.character(round(p_value, 3))))

# Simple ratio (A/S) -- the metric the surrounding prose already refers to
Ratio_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio")
n_rmc_total        <- nrow(Ratio_cor_only)
mean_rmc_all       <- round(mean(Ratio_cor_only$r), 2)
range_rmc_all      <- paste0(round(min(Ratio_cor_only$r), 2), " to ",
                             round(max(Ratio_cor_only$r), 2))
pct_rmc_strong_neg <- round(100 * mean(Ratio_cor_only$r < -0.7))

# Squared ratio (A^2/S) -- quantifies how much squaring the appendage term
# weakens (but does not eliminate) the correlation with mass.
Ratio2_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio2")
mean_rmc2_all       <- round(mean(Ratio2_cor_only$r), 2)
range_rmc2_all      <- paste0(round(min(Ratio2_cor_only$r), 2), " to ",
                              round(max(Ratio2_cor_only$r), 2))
pct_rmc2_strong_neg <- round(100 * mean(Ratio2_cor_only$r < -0.7))

# SLI-isometry / SLI-estimated species counts, for the table caption's sample-size
# caveat -- SLI-estimated is undefined (dropped, not NA-padded) for species that
# failed the per-group allometric-correlation filter, so it always has n <= SLI-isometry's.
n_sli_iso_total <- Ratio_mass_cor_all %>% filter(Metric == "Sli_iso") %>% nrow()
n_sli_est_total <- Ratio_mass_cor_all %>% filter(Metric == "Sli_est") %>% nrow()

# SLI-isometry / SLI-estimated mass correlations -- the SLI counterpart to the
# Ratio/Ratio2 stats above, for the Results paragraph's brief SLI comparison.
Sli_iso_cor_only <- Ratio_mass_cor_all %>% filter(Metric == "Sli_iso")
mean_rmc_sli_iso  <- round(mean(Sli_iso_cor_only$r), 2)
range_rmc_sli_iso <- paste0(round(min(Sli_iso_cor_only$r), 2), " to ",
                            round(max(Sli_iso_cor_only$r), 2))

Sli_est_cor_only <- Ratio_mass_cor_all %>% filter(Metric == "Sli_est")
mean_rmc_sli_est  <- round(mean(Sli_est_cor_only$r), 2)
range_rmc_sli_est <- paste0(round(min(Sli_est_cor_only$r), 2), " to ",
                            round(max(Sli_est_cor_only$r), 2))

tbl_cap_ratio_mass_summary <- glue::glue("Median per-species Pearson correlation with body mass, by empirical dataset, for the wing/mass ratio ($\\log(A/S)$), its squared counterpart ($\\log(A^2/S)$), SLI-isometry, and SLI-estimated. SLI-estimated is reported for fewer species ({n_sli_est_total} of {n_sli_iso_total} across all three datasets) than the other three metrics, because it is undefined for species that failed the per-group allometric-correlation filter (Methods).")
```

Across `r n_rmc_total` species, the ratio metric was consistently and strongly negatively correlated with mass (mean $r$ = `r mean_rmc_all`, range `r range_rmc_all`). `r pct_rmc_strong_neg`% of species had a correlation more negative than $-0.7$. This pattern held regardless of taxon, gradient type (spatial or temporal), or geographic region, confirming that the ratio's dependence on body size is a general mechanical property of the metric rather than a quirk of any one dataset (@tbl-ratio-mass-summary). I calculated the same correlation for the squared-ratio metric - the correlation between $\log(A^2/S)$ and body mass was still strongly negative, but consistently weaker in magnitude than its unsquared counterpart (mean $r$ = `r mean_rmc2_all`, range `r range_rmc2_all`, `r pct_rmc2_strong_neg`% of species below $-0.7$, @tbl-ratio-mass-summary). Both SLI parameterizations were less correlated with mass than either ratio metric (SLI-isometry: mean $r$ = `r mean_rmc_sli_iso`, range `r range_rmc_sli_iso`; SLI-estimated: mean $r$ = `r mean_rmc_sli_est`, range `r range_rmc_sli_est`) (@tbl-ratio-mass-summary; Supplementary Figure 4).

```{r}
#| label: tbl-ratio-mass-summary
#| tbl-cap: !expr tbl_cap_ratio_mass_summary

Ratio_mass_summary_tbl %>%
  dplyr::select(Study, Metric, Median_r) %>%
  pivot_wider(names_from = Metric, values_from = Median_r) %>%
  knitr::kable(
    col.names = c("Study", "Wing / Mass", "Wing² / Mass", "SLI-isometry", "SLI-estimated")
  )
```

# Discussion

My results demonstrate that body-size adjustment is not a neutral preprocessing step in studies of Allen's rule. The six approaches examined here produced systematically different estimates of relative appendage morphology in simulations, and this rank ordering was reproduced across three empirical datasets spanning `r n_species_total` species and substantially different taxa, ecosystems, and geographic contexts. The important implication is not simply that one method performed better than another, but that different approaches can quantify different aspects of morphological change. This raises a more fundamental question for studies of Allen's rule: what biological quantity should be represented as a change in relative appendage morphology? Addressing this question requires distinguishing among the morphological quantities that different statistical approaches estimate, rather than treating body-size adjustment as an interchangeable preprocessing decision.

## What constitutes an Allen's-rule response?

Allen's rule is commonly evaluated using relative appendage size, but studies do not use a single, consistent definition of the morphological response. My literature review illustrates this directly, with studies using absolute appendage measurements, ratios, regression-based residuals or conditional effects, among other approaches to quantify appendage morphology. These approaches define relative appendage morphology in different ways, because they incorporate body size through different scaling and statistical relationships. Indeed, controlling for body size is not a neutral technical preprocessing step, as traditional methods frequently fail to separate the potentially interactive, context-dependent effects of body size and other ecological factors on phenotypic variation [@glazierComplicationsBodysizeCorrection2022]. Relative appendage size should therefore be understood as an inferential quantity, with the choice of body-size adjustment determining how this biological quantity is operationalized.

## Ratios and implicit scaling relationships

Ratios are attractive because they are simple, familiar, and directly interpretable. However, their algebraic form implicitly specifies a scaling relationship between appendage length and body size. A ratio therefore does not simply "control for" body size; it defines relative morphology according to a particular scaling assumption. The simple ratios observed in this literature review ($A/S$ and $A^2/S$) were strongly correlated with body size itself in these data -- a mathematical artifact also documented across condition-index studies [@greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996]. Rather than isolating shape change, simple ratios often propagate body-size variation directly into the response variable, making it impossible to separate genuine relative shape-shifting from overall body size trends.

Importantly, the SLI metrics did a comparatively better job of removing the body size signal than ratios, which follows from the scaling relationships they use. The SLI metrics and ratios can both be thought of in the general form $A / S^\beta$. Under log-log transformation, the simple ratio $A/S$ is equivalent to $A / S^1$ (where $\beta = 1$), and the squared-ratio metric $A^2/S$ is linearized as $\sqrt{A^2/S} = A / S^{0.5}$ (where $\beta = 0.5$). The geometric appeal of dividing an area-like quantity ($A^2$) by mass is understandable, but allometric scaling theory and empirical data [@youngfleshAbioticConditionsShape2022] suggest that $1/3$ is likely a more appropriate scaling reference for appendage length than $1/2$. When using a volumetric proxy for body size (such as body mass, $M$), SLI-isometry uses the geometrically expected isometric exponent of $1/3$ and is therefore algebraically equivalent to $A/M^{1/3}$ up to a constant multiplier determined by the reference mass $M_0$. More generally, the appropriate exponent should be justified by the biological question rather than assumed a priori.

This also highlights that SLI-isometry is not fundamentally distinct from ratio approaches; rather, it represents a mathematically explicit, geometrically justified ratio. I argue that the primary advantages of the SLI framework come from transparency, interpretability, and flexibility. Rather than embedding scaling assumptions implicitly in an algebraic division, the SLI framework makes the scaling assumption ($\beta$), the reference body size ($M_0$), and the resulting units explicit. The same framework can then be used with an empirically estimated scaling exponent, as in SLI-estimated, rather than requiring a separate ratio formulation for each scaling assumption.

This dimensional bridge exposes a fundamental scaling mismatch in past studies of climate-driven morphological change. Some past work has utilized a simple ratio of an appendage to body mass ($A/M$) to represent relative appendage size [e.g. @jirinecMorphologicalConsequencesClimate2021; @ballingerContributionGeneticEnvironmental2022], implicitly forcing a scaling exponent of $\beta = 1.0$. Because a linear appendage scales with a volumetric mass proxy to the $1/3$ power ($\beta \approx 0.33$) under isometry, forcing $\beta = 1.0$ does not correspond to the geometrically expected scaling relationship for relative appendage length. When overall body mass declines in response to warming (adhering to Bergmann's rule), this scaling mismatch can create a strong negative correlation between the ratio $A/M$ and body size, meaning that $A/M$ will increase even if the relationship between appendage length and body mass remains strictly proportional [@rydingMultipleRegressionNot2025]. Consequently, the ratio can fail to isolate relative morphology because it does not appropriately remove the influence of body size, leading to spurious temperature associations that could be misclassified as adaptive shape-shifting [@glazierComplicationsBodysizeCorrection2022].

```{r}
#| label: fig-ols-sma
#| fig-cap: "The allometric relationship between log-mass and log-appendage for a simulated hyperallometric species in which temperature increases produce relatively longer appendages. The SMA regression line (blue) is steeper than the OLS line (red). Individual points are coloured by temperature. Sections labelled I and II highlight regions where OLS and SMA assign opposite signs to residuals: small-bodied individuals (Section I) are classified as having relatively short appendages by OLS but relatively long appendages by SMA. Large-bodied individuals (Section II) are classified in the opposite direction. As temperature increases select for smaller body size (Bergmann's rule), individuals are displaced toward Section I, exactly where OLS assigns more negative residuals than SMA, producing a spurious negative $\\hat{\\beta}_T$."
#| fig-pos: 'H'
#| fig-height: 4.5

# Reuses Ex_df_hypo_hyper, the same example species already generated for @fig-sli-concept
# panel (b) by Scripts/04a_SLI_concept_figure.R, which exports it here for reuse.
Ex_df_hypo_hyper <- readRDS("Derived/Rds/Ex_df_hypo_hyper.rds")
xy_lims <- Ex_df_hypo_hyper %>%
  summarize(max.append = max(Append_log), min.append = min(Append_log),
            max.mass   = max(Mass_log),   min.mass   = min(Mass_log))

ggplot(Ex_df_hypo_hyper, aes(x = Mass_log, y = Append_log)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ggpmisc::stat_ma_line(method = sma_or_ma, color = "blue") +
  geom_point(alpha = 0.4, aes(color = Temp_inc)) +
  annotate("text", x = 4.23, y = 5.13, label = "I",  size = 6, color = "black") +
  annotate("text", x = 4.52, y = 5.25, label = "II", size = 6, color = "black") +
  scale_color_viridis_c(option = "plasma") +
  lims(y = c(xy_lims$min.append, xy_lims$max.append),
       x = c(xy_lims$min.mass,   xy_lims$max.mass)) +
  labs(x = "Log(mass)", y = "Log(appendage)", color = "Temperature\nincrease")
```

## Multiple regression for conditional associations

Multiple regression takes a different approach to body size adjustment, estimating the relationship between appendage size and temperature conditional on body size. This approach was consistently more negative than the isometric reference quantity due to foundational differences between SMA and OLS regression. The SMA regression produces a steeper slope than OLS whenever $|r| < 1$, because it minimizes deviations in both dimensions rather than only vertically [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009]. The consequence for OLS-based measures of relative appendage length is illustrated in @fig-ols-sma. Because the OLS slope is shallower, OLS and SMA assign different relative appendage values to individuals at different body sizes. When temperature is associated with declining body size, as under Bergmann's rule, the temperature gradient therefore shifts observations systematically toward region I, where OLS assigns more negative residuals than SMA.

Several authors have argued that multiple regression is the most appropriate way to adjust for body size in questions of Allen’s rule [@nakagawaDivideConquerSize2017; @rydingResponseAllometryEvaluate2022; @rydingMultipleRegressionNot2025]. I agree that multiple regression is generally appropriate given that Allen's rule describes a pattern of covariation between temperature and appendage morphology [@symondsItsAllRelative2025; also see @wattBergmannsRuleConcept2010]. However, it is important to note that multiple regression does not necessarily isolate departures from isometry or any other geometric null. I generally agree with recent simulation studies that argue that a proportional change in linear dimensions is a reasonable theoretical expectation for "no shape change" when distinguishing genuine shape-shifting from size change alone [@symondsItsAllRelative2025; @rydingMultipleRegressionNot2025]. A null hypothesis of isometry is harder to justify when non-thermoregulatory forces are known to produce systematic departures from isometry. For example, bird wingspan scales with positive allometry due to the biomechanics of flight [@raynerFormFunctionAvian1988], and some ornaments used in display can experience positive allometry due to strong sexual selection [e.g., @shingletonAllometryStudyBiological2010]. Most critically, work is needed to understand the thermoregulatory relevance of isometric scaling versus departures from this null [@tabhLimitedEvidenceThat2025].

Null models aside, the statistical behavior of these OLS-based approaches has important implications for our understanding of morphological response to climate change. OLS-based approaches were consistently more negative than the isometric reference quantity, even classifying individuals as becoming relatively shorter despite the simulated population becoming longer (relative to the isometric null). OLS-based approaches were used in `r pct_ols_relative`% of studies examining relative appendage length, making this the most common approach used to quantify relative appendage length. Thus, if the conditions represented in our simulations are representative of empirical systems more broadly, the magnitude of appendage elongation [reviewed in @rydingShapeshiftingChangingAnimal2021] reported in the literature may be underestimated.

## SLI-isometry versus SLI-estimated: different inferential targets

I hope that the SLI metrics presented here will encourage biologists to think critically about allometric scaling and, more importantly, about whether the inferential quantity being estimated matches the biological question at hand. Indeed, SLI-isometry and SLI-estimated should not be viewed as competing estimators of a single universal "true" Allen's-rule effect. They instead define relative appendage morphology against different reference relationships.

One important next step is to examine the relationship between SLI-derived measures and biologically meaningful aspects of morphology or function. For example, @peigNewPerspectivesEstimating2009 found that the Scaled Mass Index predicted variation in fat and protein reserves better than several conventional condition indices across multiple vertebrate taxa. Analogous validation against independent morphological measures, such as geometric morphometric or three-dimensional measurements of appendage surface area and body volume [e.g., @rydingMeasuringAvianBill2025; @adamczakUsing3DModels2019], could similarly test whether particular SLI parameterizations capture biologically meaningful variation in relative appendage size. Such validation should span taxa near the center and the periphery of morphospace [@pigotMacroevolutionaryConvergenceConnects2020]. For example, SLI indices may capture relative appendage length well in taxa with ‘standard’ body plans (e.g., sparrows), but struggle to capture more unusual body plans (i.e., hummingbirds, penguins).

## Empirical patterns

Importantly, the divergence among these approaches was not restricted to the simulated species. The same rank ordering was reproduced across the three empirical datasets, suggesting that the methodological differences identified here are not simply artifacts of the particular covariance structures used in the simulations. Independent evidence also points to similar methodological divergence. @rydingMultipleRegressionNot2025, for example, reported substantial differences between ratio-based and multiple-regression approaches, including cases in which ratios indicated increasing relative appendage size while multiple regression indicated decreasing relative appendage size. Although their study used a different analytical framework and does not provide a direct benchmark against which the methods evaluated here can be judged, the agreement in qualitative behavior provides evidence that divergence among body-size adjustment approaches is a broader empirical issue rather than a feature unique to my simulations.

## Limitations of the SLI approaches

```{r}
#| label: sli-limitations-stats
#| include: false

Allometric_cor_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_allometric_cor.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_allometric_cor.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_allometric_cor.csv",  show_col_types = FALSE)
)

pct_sli_est_total <- round(100 * n_sli_est_total / n_sli_iso_total)

# Fraction of species per study whose mass-wing correlation falls below the
# cor_min = 0.3 threshold used to gate SLI-estimated (Methods).
pct_atlantic_weak_cor <- Allometric_cor_all %>%
  filter(Study == "Atlantic birds") %>%
  summarise(pct = round(100 * mean(cor_mw < 0.3))) %>%
  pull(pct)

pct_weeks_weak_cor <- Allometric_cor_all %>%
  filter(Study == "Weeks (2020)") %>%
  summarise(pct = round(100 * mean(cor_mw < 0.3))) %>%
  pull(pct)
```

Despite their advantages, the SLI approaches are not without flaws. Unfortunately, SMA regression is inflexible in its ability to customize models because it can only model the two scaling variables $X$ and $Y$. This inflexibility can make it more challenging to meet SMA regression assumptions relative to OLS assumptions [@wartonBivariateLinefittingMethods2006]. For example, SMA regression assumes homoskedasticity, but missing covariates that influence appendage measurements (e.g., feather wear) cannot be included in SMA regression, which could lead to heteroskedasticity. Similarly, if there is still a non-linear relationship after log-log transforming the linear and volumetric measurement, the SMA assumption of linearity is violated. This non-linearity can be easily accommodated in classical regression [e.g., $\mathit{Body\ size} + \mathit{Body\ size^2}$, @baldwinComplementarityAllensBergmanns2023], but not in SMA regression. In these cases, regression approaches that accommodate additional covariates may be preferable [but also see @glazierComplicationsBodysizeCorrection2022].

One limitation particular to SLI-estimated is that the SMA slope is estimated empirically, meaning it is sample-dependent and requires a minimal sample size to be considered reliable. Consequently, its interpretation can depend on sample composition, including sex and age structure, geographic or population structure, and the range of body sizes sampled. This is not necessarily a limitation, as SLI-estimated can accommodate group-specific allometric relationships by estimating separate scaling exponents for demographic or ecological groups. Each individual's appendage length can then be standardized to a common reference body size using the allometric relationship appropriate to its group, ultimately improving comparability among groups [@peigParadigmBodyCondition2010].

Furthermore, SMA regression is unreliable when there is little allometric signal [i.e., mass–appendage correlation is weak; @smithUseMisuseReduced2009]. In the empirical datasets examined here the SMA slope could be estimated for only `r pct_sli_est_total`% of species with the sample size and minimum correlation filters employed. Interestingly, there was a general morphological decoupling of mass and wing in Atlantic Forest birds (`r pct_atlantic_weak_cor`% below r \< 0.3) that was less prominent in North American migrants \[`r pct_weeks_weak_cor`%; @skinnerDataEnvironmentalGeographic2025, @weeksSharedMorphologicalConsequences2020\]. While this dependence on a minimum correlation to use SLI-estimated can be seen as a limitation, researchers should examine whether there is a need to control for body size in the first place if correlation between size and appendage is minimal.

## Future work: From association to causal inference

For those who seek to move beyond documenting patterns, causal inference provides a framework for asking how environmental changes would alter biological responses under specified interventions [@franksEcologyNeedsCausal2025]. For example, predicting the effects of climate change on species or populations, separately from other environmental changes such as land-use change, requires asking how morphology would change under an intervention on temperature, precipitation, or any other environmental gradient, relative to an appropriate counterfactual in which that intervention did not occur [@dudneyCausalInferenceFramework2024]. Thus, conservation in the face of global change will increasingly require approaches that distinguish causal effects from statistical associations to distinguish the consequences of distinct, but often correlated, environmental drivers [@franksEcologyNeedsCausal2025].

Another benefit of a causal inference framework is the ability to decompose a total effect into direct and indirect effects. In the context of Allen's rule, the effect of temperature on appendage length could be decomposed into an indirect effect mediated through body size (temperature $\rightarrow$ body size $\rightarrow$ appendage length) and a direct effect representing the temperature $\rightarrow$ appendage pathway. Path analysis or structural equation modeling within an explicit causal framework are particularly well suited to estimating such decompositions [e.g., @xuEcologicalPredictorsInterspecific2023], which could help clarify the combined effects of Bergmann's and Allen's rules [@skinnerBergmannAllenUnified2026]. Importantly, the SLI decomposition should not itself be interpreted as a causal decomposition. The components describing allometric and temperature-associated variation are statistical quantities defined by the chosen standardization framework, but are not causal pathways that cleanly separate into direct and indirect effects of temperature on appendage length.

## Recommendations for empirical studies

These results suggest several considerations for studies of Allen's rule and other questions involving relative morphology. First, researchers should define the biological question and estimand of interest before selecting a body-size adjustment technique. Absolute appendage change, departure from theoretical scaling, and deviation from empirical allometry are distinct quantities and should be described explicitly. SLI-isometry and SLI-estimated provide transparent, interpretable measures of relative appendage length that can be selected according to whether the research question concerns theoretical isometry or empirical allometry. Together, they provide a flexible framework for quantifying relative appendage morphology while keeping the underlying scaling assumption explicit.

Second, the scaling assumptions underlying body-size adjustment should be made explicit and justified in relation to the biological question. Ratios, regression approaches, and SLI encode or estimate different scaling relationships, and differences among them should not be dismissed as merely statistical artifacts. Ratios may be appropriate when their implied scaling relationship corresponds to the intended biological quantity, but given the improved interpretability and transparency of the SLI metrics, I recommend SLI over simple ratios for quantifying relative appendage length.

Third, descriptive association should be distinguished from causal interpretation. A body-size-adjusted temperature coefficient can describe a temperature-associated pattern without representing the causal effect of temperature on morphology. Where causal interpretation is the objective, researchers should articulate the relevant causal assumptions and consider whether an explicit causal framework is warranted [@siegelFoundationsFutureDirections2025].

# Conclusions

Relative appendage length is not a uniquely defined quantity. Different approaches encode different assumptions about how body size and appendage morphology should be related, and these assumptions lead to systematically different estimates of climate-associated morphological change. The SLI metrics presented here provide an explicit and interpretable framework for scalar relative appendage size, with two parameterizations that correspond to distinct biological questions. More broadly, identifying an Allen's-rule pattern is distinct from establishing its causal basis. Careful definition of the morphological quantity being analyzed is therefore a necessary first step toward interpreting climate-associated changes in relative morphology. Indeed, adjustment for body size should be treated as part of the inferential design of a study rather than as a routine preprocessing step.

# Data Availability

All analyses can be reproduced using the code and data in the zipped 'for reviewers' file. Nightjar morphological data are available from @skinnerDataEnvironmentalGeographic2025. North American migrants data are available from @weeksSharedMorphologicalConsequences2020. Atlantic Forest bird data are available from @rodriguesATLANTICBIRDTRAITS2019.

# Acknowledgements

I thank Andy Green for helpful comments on the development of the Standardized Length Index and its relationship to the Standardized Mass Index, and David Warton for clarifying the assumptions and properties of SMA regression.

{{< pagebreak >}}

# References
