---
title: "The Shape of Things to Come: Simulations Show Promise in a New Metric, the Standardized Length Index"
number-sections: true
execute:
  echo: false
  message: false
  warning: false
format: #docx
  elsevier-pdf: 
    toc: false
    header-includes:
      - |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhead[R]{SLI to examine Allen's rule}
        \usepackage{makecell}
        \usepackage{lineno}
        \linenumbers
        \usepackage[margin=1in]{geometry}
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
        \usepackage{tabularray}
    journal:
      name: Methods in Ecology and Evolution
      formatting: preprint #doubleblind suppresses authors!, return to preprint
      cite-style: authoryear
bibliography: ../../Suppfiles/Morphology.bib
link-citations: true
csl: ../../_extensions/quarto-journals/elsevier/bib/elsevier-harvard.csl
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
editor_options:
  chunk_output_type: console
metadata-files: 
  - ../../Suppfiles/abstract_keywords.yml
  - ../../Suppfiles/authors.yml
---

{{< pagebreak >}}

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(png)
library(grid)
library(ggpubr)
library(cowplot)
library(scales)
library(smatr)
library(MASS)
library(ggpmisc)
ggplot2::theme_set(theme_cowplot())
```

```{r}
#| label: load-lit-review-results
#| include: false

# Literature review is processed separately by Scripts/Lit_review.R (re-run it any time the underlying spreadsheet or classification logic changes) and saved to Derived/Rds/lit_review_results.rds. This brings Lit_review, Unclear_exclude, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/lit_review_results.rds"), envir = .GlobalEnv)
```

```{r}
#| label: Percents-in-line

Per_class <- Lr %>%
  summarize(Percent = n() / nrow(Lr) * 100, .by = class) %>%
  mutate(Percent = paste0(round(Percent, 0), "%"))

per_aves <- Per_class %>% 
  filter(class == "Aves") %>% 
  pull(Percent)

# Pooled (both classes combined) approach-usage counts/percentages, cited
# inline below and matching the combined @fig-lit-review panel.
pull_pooled <- function(approach_name, stat) {
  Lr_per_pooled %>% filter(approach == approach_name) %>% pull({{ stat }})
}

tot_approach_uses <- unique(Lr_per_pooled$Tot_studies)
pct_mv_reg <- pull_pooled("multiple_regression", Per)
pct_raw    <- pull_pooled("raw_values", Per)
pct_other  <- pull_pooled("other", Per)
pct_ratio  <- pull_pooled("ratio", Per)
pct_resid  <- pull_pooled("residuals", Per)

# How many studies had multiple approaches? 
Num_mult_approach <- Lr_pivot %>% 
  count(study, sort = T) %>% 
  filter(n > 1) %>% 
  nrow()

# Studies examining relative (not absolute) appendage length, by whether mass
# was the body-size-control variable used.
n_relative_studies <- sum(Mass_control_tbl$n)
n_mass_control      <- Mass_control_tbl %>% filter(Mass_control == TRUE) %>% pull(n)
pct_mass_control    <- round(100 * n_mass_control / n_relative_studies, 0)
```

```{r}
#| label: source-key-fns
#| include: false
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: load-simulation-results
#| include: false

# Simulation is run separately by Scripts/Run_simulation.R (re-run it any time the parameter grid, sample size, or evaluation logic changes) and saved to Derived/Rds/simulation_results.rds
# Both this manuscript and supplementary_info.qmd just read the result. This brings N_spp, Parms_mat3, Parms_tbl4, Eval_tbl, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/simulation_results.rds"), envir = .GlobalEnv)
```

# Introduction

Wildlife around the world are responding to contemporary climate change with shifts in morphology. Studies have documented declines in body size consistent with Bergmann's rule across diverse taxa, and growing evidence suggests that body shape is also changing in a pattern consistent with Allen's rule [@rydingShapeshiftingChangingAnimal2021; @rydingLongShorttermResponses2024]. Because Allen's rule concerns appendage length relative to overall body size, testing it requires methods that can distinguish variation in appendage size from variation in overall body size [@symondsItsAllRelative2025]. Here I distinguish an isometric definition of relative morphology from one based on the species' observed allometry. Yet the statistical methods used to quantify relative appendage length have received far less attention than the biological hypothesis itself.

There have been three primary approaches used in empirical studies to test Allen's rule (@fig-lit-review). The simplest uses absolute appendage length as the response variable, regressed against temperature or spatial or temporal proxies (e.g., latitude or year), hereafter referred to as the 'gradient of interest'. Throughout this paper I focus on temperature, although the methods apply equally to any of these gradients. Using absolute appendage length has been compared with methods that explicitly account for body size elsewhere [e.g., @symondsItsAllRelative2025], so I do not consider it further. A second approach, the most common in this review, includes body size as a covariate in a multiple regression, interpreting the slope of temperature (hereafeter, $\hat{\beta}_T$) as temperature's conditional effect on appendage length after accounting for body size. A third approach takes the ratio of appendage length divided by body size as the response variable.

Despite their widespread use, both the multiple regression and ratio approaches have important statistical limitations. Ratio-based metrics are frequently correlated with body size itself, limiting their utility as measures of relative shape [@jakobEstimatingFitnessComparison1996; @greenMassLengthResiduals2001]. Multiple regression, on the other hand, makes a key assumption – it assigns all unexplained variation around the fitted allometric line to the $Y$ variable (appendage length). Consequently, every deviation from the fitted line is interpreted as evidence of an unexpectedly long or short appendage, rather than allowing deviations to be shared between both variables (@nte-sma-intuition). This assumption is often difficult to justify in studies of allometric scaling [@greenMassLengthResiduals2001], and alternative approaches have been recommended [@wartonSmatr3Package2012].

Standardized major axis (SMA) regression provides an alternative line-fitting approach by treating variation in body size and appendage length symmetrically rather than assigning all unexplained variation to the response variable (@wartonBivariateLinefittingMethods2006; @nte-sma-intuition). Consequently, SMA estimates steeper allometric slopes whenever the correlation between $X$ and $Y$ is $< 1$. SMA has long been recommended for studies of allometric scaling [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009; @greenMassLengthResiduals2001] and has been used in studies examining Allen's rule \[e.g., CITE Adamczack 2020\], yet its performance has never been systematically compared with the classical methods used to quantify relative appendage length. Understanding these different approaches to addressing Allen's rule, and the conditions under which the methodologies diverge, is necessary for determining which approach is appropriate for the biological question at hand.

Here I introduce the Standardized Length Index (SLI), an analogue of the Standardized Mass Index [@peigNewPerspectivesEstimating2009] but adapted to measure appendage length relative to body size. The SLI uses the SMA slope to standardize each individual's appendage length to a common reference body size, producing a size-corrected shape index in the original units of the appendage. I present two parameterisations, corresponding to two distinct questions about relative appendage change: SLI-isometry fixes the scaling exponent at 0.33 (the isometric expectation for a linear versus volumetric measurement like mass) and asks whether appendage size departs from that proportional, geometric expectation. SLI-estimated instead estimates the exponent empirically from the data and asks whether appendage size departs from the species' own observed allometric relationship. Using simulated data from `r N_spp` hypothetical species spanning a realistic range of allometric relationships and temperature sensitivities, I characterize how these six approaches diverge from one another under different combinations of allometric scaling and temperature-associated changes in body size and appendage length. I also examine whether the rank-order patterns that emerge during simulation are reproduced across three independent empirical datasets. Finally, I provide recommendations for future studies examining Allen’s rule.

::: {#nte-sma-intuition .callout-note icon="false"}
## Building intuition for SMA regression

SMA and OLS regression differ in two closely related ways. First, SMA regression estimates a steeper allometric slope whenever the correlation between variables is \< 1. Second, because it treats variation in both variables symmetrically, relative appendage length and relative body size become reciprocal descriptions of the same deviation from the allometric relationship. The equations below show why these properties arise.

*SMA estimates a steeper allometric slope*

There is a direct mathematical link between the slopes of OLS and SMA regression.\
The slope of a predictor variable $X$ (e.g., body size) is related to the response variable $Y$ (e.g., appendage length) in OLS regression via the following equation:

- $\beta_{ols} = r * \frac{Sy}{Sx}$ where $r$ is the correlation between $X$ and $Y$, and $S$ is the standard deviation

The slope of SMA regression is estimated by:

- $\beta_{sma} = sign(r) * \frac{Sy}{Sx}$ , i.e., if $r$ is negative then $\beta_{sma}$ is also negative.

A simple substitution of $\frac{Sy}{Sx}$ in the first equation shows that

$$\beta_{ols} = r * \beta_{sma}$$

This makes it evident that as $r$ increases, $\beta_{\text{OLS}} \rightarrow \beta_{\text{SMA}}$, and that the difference in estimated slopes will be highest when $X$ and $Y$ are only weakly correlated.

*SMA gives a symmetric geometric interpretation*

Unlike OLS, the SMA line of best fit is unchanged when the $X$ and $Y$ axes are flipped (@fig-sma_flip_axes), leading to a useful geometric property: swapping the axes simply reflects the points across the fitted line. Consequently, the individual highlighted in green can be described either as having relatively short appendages for its body size (panel A) or as having a relatively large body for its appendage length (panel B). These are reciprocal descriptions of the same deviation from the allometric relationship (i.e., residuals retain the same magnitude but reverse sign). In contrast, because the OLS line changes when the axes are exchanged, this reciprocity does not hold.

![In SMA regression the X and Y axes can be flipped and the line of best fit will not change. Here I flip the axes and plot the OLS (red) & SMA (blue) regression lines of best fit, as well as the same individual from my hypothetical species (highlighted in green). SMA residuals are equally correlated with both $X$ and $Y$ ($|r_{sma}| = 0.59$, whereas OLS residuals remain uncorrelated with the predictor but strongly correlated with the response ($r_{ols} = 0.96$).](../../Figures/sma_flip_axes.png){#fig-sma_flip_axes width="100%"}
:::

# Literature review

I reviewed empirical studies published between 2021 and 2025 that quantified appendage length in the context of Allen's rule in endotherms. Studies were identified using the list of studies published in @skinnerBergmannAllenUnified2026 and subset to include only those that examined Allen's rule. I scored `r nrow(Lit_review)` papers of which `r nrow(Unclear_exclude)` were excluded for not meeting inclusion criteria. Five additional studies were excluded due to unclear methodologies, leaving 45 studies. Sixteen of these studies used multiple approaches to examine Allen's rule, resulting in `r tot_approach_uses` analytical approaches that were classified into five categories: ratio-based methods, two-step residual regression, multiple regression with body size as a covariate, use of absolute appendage length only, or other. 'Other' included shape axes from principal component analyses, path analyses, and geometric morphometric approaches, among other creative solutions [e.g., @youngfleshAbioticConditionsShape2022]. Multiple regression was the most common approach (`r pct_mv_reg`%), followed by absolute appendage length (`r pct_raw`%), other approaches (`r pct_other`%), ratio-based methods (`r pct_ratio`%), and residual regression (`r pct_resid`%; @fig-lit-review). Of the studies examining appendage length *relative* to body size, `r n_mass_control` / `r n_relative_studies` (`r pct_mass_control`%) used mass as the variable representing body size. The majority of studies (`r per_aves`) examined Allen's rule in birds. The remainder used mammal focal taxa.

```{r}
#| label: fig-lit-review
#| fig-cap: !expr fig_cap_lit_review
#| fig-height: 6.5
#| fig-pos: 'H'

fig_cap_lit_review <- glue::glue("Frequency of the primary methodological approaches used across 45 studies testing Allen's rule since 2021. Each point represents one analytical approach (n = {tot_approach_uses}) to classify relative appendage length, and colored lines connect a single study with multiple approaches. Percentages are computed as the proportion of total approaches rather than the proportion of studies.")

lit_review_img  <- png::readPNG("Figures/lit_review.png")
lit_review_grob <- grid::rasterGrob(lit_review_img, width = grid::unit(1, "npc"),
                                     height = grid::unit(1, "npc"))
cowplot::ggdraw() + cowplot::draw_grob(lit_review_grob)
```

# Materials and Methods

## Data simulation

I simulated appendage length, mass, and a temperature increase for `r N_spp` hypothetical species, each represented by approximately `r N_ind` individuals. I simulated mass measurements as a proxy for body size as it was the most frequent variable used to control for body size in studies of Allen’s rule. Data were drawn from a multivariate log-normal distribution:

$$\begin{bmatrix}
\log(\text{Appendage}) \\
\log(\text{Mass}) \\
\text{Temperature}
\end{bmatrix}
\sim \mathrm{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

with $\boldsymbol{\mu} = [\log(180),\ \log(80),\ 1.0]^\top$ representing a species with a mean appendage length of 180 mm and mean mass of 80 g, experiencing an average 1°C temperature increase. I simulated morphological and environmental data from a joint multivariate normal distribution rather than from a directional model (e.g., $Appendage \sim Mass + Temperature + \epsilon$). This approach does not designate any variable as a response or predictor in the data-generating process, allowing characterization of each statistical approach against the same underlying joint distribution without presupposing a particular regression structure. The covariance matrix $\boldsymbol{\Sigma}$ was constructed from three Pearson correlations ($\rho_{12}$: appendage–mass; $\rho_{13}$: appendage–temperature; $\rho_{23}$: mass–temperature) and the allometric scaling slope. Hypothetical species were constructed by crossing the following parameter values: appendage–mass correlation $\rho_{12} \in {`r paste(r_12, collapse = ", ")`}$, temperature–morphology correlations $\rho_{13}, \rho_{23} \in {`r paste(r_13, collapse = ", ")`}$, crossed factorially to produce species with different relative temperature associations of mass and appendage, and allometric scaling slope $b \in {`r paste(b_avg_12, collapse = ", ")`}$, representing hypoallometry, isometry, and hyperallometry respectively. Standard deviations of log-mass and log-appendage were set so that the average of the OLS and SMA slopes equalled the specified allometric slope, allowing me to vary allometric scaling independently of the correlation structure.

## Effect classification

Because species were simulated from a covariance structure rather than from a causal model, there is no single fitted regression coefficient representing a universal "true" effect against which every method's estimate is either correct or incorrect. Instead, I define an isometrically-defined reference quantity, $\beta_T$, that captures departure from proportional (isometric) scaling between appendage length and body size. This is one specific, theoretically motivated definition of relative appendage change, not a privileged ground truth that every method should be expected to recover:

$$\beta_T = \rho_{13} \cdot \beta_{SMA} - 0.33 \cdot \rho_{23}, \qquad \beta_{SMA} = \frac{2b}{\rho_{12}+1}$$ {#eq-b-true}

where $b$ is the simulated allometric scaling slope, and $\beta_{SMA}$ is a species' population SMA slope implied by its simulated parameters. $\beta_T = 0$ denotes no departure from this isometric reference. Defining $\beta_T$ in terms of both allometric scaling and the strength of temperature-morphology relationships makes it clear that there are two ways a species could depart from $\beta_T = 0$: (1) temperature having a differential association with mass vs. appendage ($\rho_{13} \neq \rho_{23}$), or (2) the species exhibiting non-isometric scaling ($\beta_{SMA} \neq 0.33$). Rewriting $\rho_{13} = \rho_{23} + \Delta\rho$, where $\Delta\rho$ is the difference between $\rho_{13}$ and $\rho_{23}$, decomposes @eq-b-true into these same two distinct components, expressed additively within the isometric-relative reference quantity itself:

$$\beta_T \;\propto\; \underbrace{\rho_{23}\left(\beta_{SMA} - 0.33\right)}_{\text{allometry-vs-isometry component}} \;+\; \underbrace{\beta_{SMA} \cdot \Delta\rho}_{\substack{\text{differential temperature-} \\ \text{association component}}}$$

As simulated temperature increased, species with $\beta_T > 0$ were classified as getting relatively longer and $\beta_T < 0$ species were classified as getting relatively fatter. Fatter and Longer are both realizations of Bergmann's rule, as warming always decreased overall body size in these two categories.

I additionally simulated a set of species with $\beta_T = 0$ exactly, classified as proportionally decreasing in size ("Proportionally smaller"). I solved for $\rho_{13}$ algebraically based on a species $\rho_{23}$ and $\beta_{SMA}$, following directly from @eq-b-true:

$$\rho_{13} = \frac{0.33\ \cdot \rho_{23}}{\beta_{SMA}}$$

guaranteeing $\beta_T = 0$ exactly, regardless of allometric scaling category.

Finally, I simulated species that increased in size with warmer temperatures, in line with several recent studies that have documented inverse Bergmannian clines [@stapelfeldtLongtermFieldStudy2023; @castillo-figueroaDoesBergmannsRule2022; @jumanRecentRapidEcogeographical2022]. The 'Proportionally larger' scenario was the mirror image of the 'Proportionally smaller' category, solving $\rho_{13}$ the same way from positive $\rho_{23} \in \{0.15, 0.3, 0.5, 0.7\}$, $b \in \{0.22, 0.33, 0.44\}$, and $\rho_{12} \in \{0.3, 0.45, 0.6\}$ (as above), representing species that grow larger with warming while maintaining the isometric-relative reference quantity at zero.

## Statistical approaches

I applied six methods from three families to each simulated species.

**Ratio approaches: (**1) The appendage-to-mass ratio $log(A / S)$ was regressed on temperature. (2) The squared appendage-to-mass ratio $log(A^2 / S)$ was regressed on temperature, reflecting a surface-area-to-volume formulation used in some empirical studies [e.g., @nwaoguTemperatureAridityDetermine2018; @dubinerWidespreadRecentChanges2022].

**Ordinary least squares (OLS) regression approaches: (**1) A multiple regression approach where appendage was modelled as a linear function of mass and temperature simultaneously. The temperature coefficient $\beta_T$ was interpreted as temperature's conditional effect on appendage length, holding mass constant. (2) A two-step residual regression approach was conducted where appendage was regressed on mass using OLS, and the resulting residuals were regressed on temperature.

**Standardized Length Index — estimated (SLI-estimated):** Following @peigNewPerspectivesEstimating2009, I computed:

$$SLI_i = A_i \cdot \left(\frac{M_0}{M_i}\right)^{b_{SLI}}$$

where $A_i$ is the appendage length of individual $i$, $M_i$ is the mass of individual $i$, $M_0$ is the mean mass of the sampled population, and $b_{SLI}$ comes from the empirical SMA slope of $\log(A)$ on $\log(S)$ estimated from the data. The SLI reprojects each individual's appendage length to the value it would take at the population mean body size. $SLI_i$ is then regressed on temperature.

**Standardized Length Index — isometry (SLI-isometry):** Identical to SLI-estimated but replacing the estimated $b_{SLI}$ by $b_{SLI} = 0.33$, the isometric scaling exponent for a linear versus volumetric measurement.

Although the present study focuses on appendage length relative to body mass, the SLI framework can also be applied when both the body-size measure and the appendage measure are linear dimensions (e.g., body length and wing length). In this case, the isometric reference exponent is $b_{SLI} = 1$, because both measurements scale linearly with length rather than volumetrically. This formulation may be particularly useful when body condition is expected to vary systematically with temperature or another environmental gradient.

I log-transformed all variables and then standardised to unit variance before fitting regression models so that coefficients were comparable across species and methods.

These six methods differ in how they account for body size and, for the OLS approaches, in how they treat the relationship between mass and appendage length (@tbl-methods-estimands). The OLS approaches and SLI-estimated share a similar goal of separating variation associated with body size from variation in appendage size, but differ in how they define the mass–appendage relationship: OLS treats mass as the conditioning variable and appendage length as the response, whereas SLI-estimated uses a symmetric SMA relationship.

```{r}
#| label: tbl-methods-estimands
#| tbl-cap: "The six statistical approaches compared in this study, how each accounts for body size, the assumed relationship between mass and appendage length, and the quantity represented by the resulting measure. 'Directional' denotes the direction assumed by the statistical model, not a causal claim; a causal interpretation of that direction requires additional assumptions (main text)."

Methods_estimands_tbl <- tibble::tribble(
  ~Method,                    ~`Body-size relationship`,             ~`Mass–appendage relationship`,          ~`Quantity represented`,
  "Ratio ($A/S$)",            "$b = 1$ (fixed)",                      "Proportional relationship",        "Ratio-defined relative size",
  "Squared ratio ($A^2/S$)",  "$b = 0.5$ (fixed)",                    "Geometric relationship",           "Surface-to-volume-inspired relative size",
  "OLS multiple regression",  "$Appendage \\sim Mass$",       "Directional (statistical model): $Mass \\rightarrow Appendage$", "Temperature effect on appendage, conditional on mass",
  "OLS residuals",            "$Appendage \\sim Mass$",       "Directional (statistical model): $Mass \\rightarrow Appendage$", "Residual appendage size after OLS adjustment",
  "SLI-isometry",              "$b = 0.33$ (fixed)",                   "Relative to isometry",            "Appendage size relative to isometric scaling",
  "SLI-estimated",             "Species-specific SMA slope",          "Symmetric: $Mass \\leftrightarrow Appendage$", "Appendage size relative to species-specific allometry"
)

knitr::kable(Methods_estimands_tbl, format = "latex", align = "c", booktabs = TRUE, escape = FALSE, linesep = "") %>%
  kableExtra::column_spec(1, width = "2.6cm", latex_valign = "m") %>%
  kableExtra::column_spec(2, width = "3.1cm", latex_valign = "m") %>%
  kableExtra::column_spec(3, width = "3.4cm", latex_valign = "m") %>%
  kableExtra::column_spec(4, width = "4.2cm", latex_valign = "m") %>%
  kableExtra::kable_styling(font_size = 10)
```

## Statistical approach classification

I evaluated the six statistical approaches using two criteria relative to the isometrically defined benchmark quantity $\beta_T$. First, I assessed whether each approach agreed with the sign of $\beta_T$: a positive $\hat{\beta}_T$ when $\beta_T > 0$, a negative $\hat{\beta}_T$ when $\beta_T < 0$, and $\beta_T \approx 0$ when $\beta_T = 0$. I considered $\hat{\beta}_T$ to agree with the benchmark when its 95% confidence interval excluded zero in cases where $\beta_T \neq 0$, or included zero when $\beta_T = 0$. Second, I assessed how closely $\hat{\beta}_T$ tracked $\beta_T$ in magnitude using a bivariate regression of $\hat{\beta}_T$ on $\beta_T$ across species. An approach that reproduced the benchmark exactly would yield no sign misclassifications, $r = 1$, and estimates falling on the 1:1 line.

Importantly, $\beta_T$ is not a universal ground truth against which the six approaches can be ranked. I defined $\beta_T$ using the SMA scaling relationship anchored to an exponent of 0.33, reflecting the theoretical expectation that $A \propto M^{1/3}$ [@shingletonAllometryStudyBiological2010] and the same isometric anchor used by SLI-isometry. A benchmark defined using the OLS scaling relationship, or another scaling exponent, would instead define a different quantity and could consequently favour different approaches (Supplementary Figure X). The purpose of this comparison is therefore not to identify the universally most accurate method, but to characterize how the six approaches behave relative to a clearly defined isometric quantity.

Because SLI-isometry uses the same isometric anchor as $\beta_T$, close agreement between the two is expected by construction. Indeed, the relationship is stronger than simple empirical correspondence: the SLI-isometry coefficient and $\beta_T$ differ only by a known scaling factor, consisting of a variance-correction term derived from the same three simulation parameters used to define $\beta_T$ (Supplementary Information). Dividing the SLI-isometry coefficient by this factor produces an essentially exact correspondence with $\beta_T$, demonstrating that the two quantities are algebraically equivalent up to rescaling. This correspondence provides an algebraic check on the implementation of the isometric benchmark and SLI-isometry calculation, but does not demonstrate that SLI-isometry is superior to the other approaches. Rather, the informative comparisons concern how the remaining approaches, which use different scaling or regression assumptions, systematically diverge from the isometrically defined benchmark and the conditions under which those divergences occur.

## Empirical case studies

I applied all six methods to three independent empirical datasets to assess whether the rank-order patterns observed in simulation were also observed in real data. First, I used morphological data from three migratory nightjar species, European nightjars (Caprimulgus europaeus), eastern whip-poor-wills (Antrostomus vociferus), and common nighthawks (Chordeiles minor), spanning a spatial temperature gradient [@skinnerEnvironmentalGeographicConditions2025]. Second, I applied the six methods to a dataset comprising 43 North American migrant species (1979–2016), using capture year as the gradient of interest [@weeksSharedMorphologicalConsequences2020]. Third, I applied the methods to a dataset of 49 Atlantic Forest bird species using mean annual temperature as the environmental gradient [@rodriguesATLANTICBIRDTRAITS2019]. Together, these datasets span different taxonomic groups, ecological systems, gradients of interest, and putative selective pressures, providing a broad test of whether the same rank-order patterns are observed across empirical contexts.

Age and sex can influence body size independently of the environmental gradient (e.g., through sexual size dimorphism or age-related growth). For each species, I therefore tested whether age or sex covariates predicted mass or wing length while controlling for the gradient variable (i.e., temperature or year). Covariates were retained only when age or sex influenced body size ($p < 0.10$) and when each age or sex group contained at least 100 individuals.

Retained covariates were incorporated in only certain methods. For the ratio ($A/S$) and squared-ratio ($A^2/S$) I assume that age and sex influence wing and mass roughly proportionally, and thus would not importantly modify the ratios. I do not include age and sex in the SLI-isometry approach because $b_{SLI}$ is fixed at 0.33 by definition. For the OLS-residual approach, age and sex were included in the first-stage regression of wing on mass. For multiple-regression, they were included alongside mass and the gradient variable. For SLI-estimated, SMA slopes were estimated separately for each age and sex group and then averaged depending on the demographic status of each individual (e.g., a hatch-year male would receive $(b_{HY} + b_{M}) / 2$).

To evaluate whether the same rank-order patterns were observed in empirical data, I averaged the gradient coefficients within each method family (Ratio: $A / S$ Ratio and $A^2 / S$ Ratio; OLS: multiple regression and OLS-residual; SLI: SLI-isometry and SLI-estimated) for each species and compared their rank order with the simulation results. Because the expected ordering depends on the direction of body size change (Supplementary Figure X), species were first classified according to whether they exhibited Bergmannian, inverse Bergmannian, stable, or mixed size trends (@tbl-classification). Bergmannian species exhibited declines in wing length, body mass, or both along the gradient, whereas inverse Bergmannian species exhibited increases in one or both traits ($p < 0.05$). Stable species showed no significant change in either trait, while mixed species exhibited opposing trends in wing length and body mass. Only Bergmannian and inverse Bergmannian species were used to evaluate agreement with the simulation results because these scenarios correspond directly to the simulated data-generating process. Species with stable or mixed size trends are reported separately in the Supplementary Information. Importantly, SMA regression is unreliable when the correlation between appendage and body mass is weak [@smithUseMisuseReduced2009]. Thus, SLI-estimated was calculated only for species with $r \geq 0.3$.

Finally, past authors have noted that ratio methods tend to be highly correlated with the denominator variable. To test this, I calculated the Pearson correlation between both ratio metrics and body mass for every species in the three empirical datasets.

```{r}
#| label: tbl-classification
#| tbl-cap: "Classification of species according to significant ($p < 0.05$) changes in wing length and body mass along the environmental gradient (temperature or year). Non-significant coefficients did not contribute to classification."

Classification_tbl <- tibble::tribble(
  ~Category,               ~`Classification criterion`,
  "Bergmann's",            "Wing length, body mass, or both decrease",
  "Inverse Bergmann's",    "Wing length, body mass, or both increase",
  "Mixed - Wingyier",      "Wing length increases while body mass decreases",
  "Mixed - Fatter",        "Wing length decreases while body mass increases",
  "Stable",                "No significant change in either trait"
)
knitr::kable(Classification_tbl)
```

# Results

## Simulation performance

```{r}
#| label: plot-approaches-fun
#| include: false

plot_approaches <- function(df, x_txt_size = 9, legend.pos = "top") {
  df <- df %>%
    filter(Temp_eff != "Proportionally larger") %>%
    mutate(Model = str_replace_all(Model, x_labs))

  ggplot(df, aes(x = Model, y = b_temp_inc)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, height = 0, alpha = .5, size = 1.9,
                aes(color = Strength, shape = Scaling)) +
    facet_wrap(vars(Temp_eff)) +
    labs(x = NULL, y = expression(hat(beta)[T] ~ "on relative appendage length"),
         color = "Strength", shape = "Scaling") +
    guides(shape = guide_legend(nrow = 1)) +
    theme(
      axis.text.x = element_text(size = x_txt_size, vjust = .58, angle = 55),
      legend.position = legend.pos, legend.box = "vertical"
    )
}
```

The six methods produced strikingly different estimates of temperature's effect on relative appendage length, and the pattern was consistent across allometric scaling categories and correlation structures (@fig-compare-approaches).

```{r}
#| label: fig-compare-approaches
#| fig-cap: "Estimated effect of temperature on relative appendage length ($\\hat{\\beta}_T$) for each method across Fatter, Proportionally smaller, and Longer species: three different realizations of Bergmann's rule, differing only in how fast appendage length declines with warming relative to mass (faster, equally, or slower, respectively). Positive values indicate that species become relatively longer with increasing temperature. Negative values indicate they become relatively fatter. Each point is a simulated species. The dashed line at zero is the isometric-relative reference value for every Proportionally smaller species, by construction. Point shape indicates allometric scaling category and colour indicates the strength of the difference between temperature's effect on appendage length and mass."
#| fig-width: 9
#| fig-height: 4.5
#| fig-pos: 'H'

plot_approaches(Parms_tbl4, x_txt_size = 11)
```

**Ratio approaches** tended to diverge from the isometric-relative reference quantity in the positive direction, although this was much more prominent when $A/S$ was used as the response variable. When simulated to become relatively fatter, $A^2/S$ disagreed with the reference for only `r Ratio2_fatter_wrong_n` of `r Fatter_n` hypothetical species, whereas $A/S$ disagreed for roughly 1/4 of species (@fig-eval). Among species simulated as Proportionally smaller, where the reference quantity is exactly zero by construction, no species had a 95% confidence interval that included zero for $A/S$, and none for $A^2/S$: both ratio metrics showed statistically significant departures from the isometric-relative reference quantity in every null-effect species, a direct consequence of ratio metrics' mechanical dependence on mass (@tbl-ratio-mass-summary).

**Multiple regression and OLS residuals** approaches consistently diverged from the isometric-relative reference quantity in the negative direction (@fig-eval). Both methods returned negative $\hat{\beta}_T$ in the majority of species simulated to become relatively longer, agreed with the reference for Fatter species, and overcorrected in Proportionally smaller species: only `r Ryding_proportional_null_correct`% of these species had a 95% confidence interval that included zero for either method, showing statistically significant departures from the null isometric-relative reference quantity in the large majority of null-effect species. The two approaches behaved nearly identically, consistent with their mathematical equivalence when predictor variables are uncorrelated.

**SLI-isometry** tracked the isometric-relative reference quantity closely, as expected since both are anchored to the same isometric scaling exponent (@fig-eval). It agreed in sign with the reference for `r Sli.iso_longer_right`% of species simulated to become relatively longer, and `r Sli.iso_fatter_right`% of species simulated to become relatively fatter. Among species simulated as Proportionally smaller, `r Sli.iso_proportional_null_correct`% had a 95% confidence interval that included zero, against `r Sli.est_proportional_null_correct`% for SLI-estimated and `r Ratio_proportional_null_correct`–`r Ryding_proportional_null_correct`% for the four non-SLI methods, the highest agreement rate of any method. This close correspondence is not, by itself, evidence that SLI-isometry outperforms the other five approaches. SLI-isometry's own formula and the isometric-relative reference quantity used to evaluate it are algebraically related by construction, so this result confirms that the simulation and reference calculation behave as expected, rather than establishing SLI-isometry's superior accuracy.

**SLI-estimated** agreed with the reference at a similar rate to SLI-isometry in the Longer and Fatter categories (`r Sli.est_longer_right`% and `r Sli.est_fatter_right`% agreement, respectively; @fig-eval), but diverged sharply among Proportionally smaller species, where only `r Sli.est_proportional_null_correct`% of species had a 95% confidence interval that included zero. This divergence is expected, not a failure of the method. By construction, zeroing $\beta_T$ in this design requires $\rho_{13} \neq \rho_{23}$, and SLI-estimated is structurally sensitive to exactly that differential temperature-association component, since it standardises each species against its own realized scaling relationship rather than against the fixed isometric reference. It therefore retains a statistically detectable signal along that component even when the isometric-relative reference quantity, the quantity SLI-isometry is built to track, is exactly null. SLI-isometry and SLI-estimated are therefore not competing to recover the same quantity in this category. They are answering two different questions that happen to have different answers here.

Results for an additional **Proportionally larger** scenario are presented in the Supplementary Information, but in short these hypothetical species showed the inverse pattern as the Proportionally smaller species.

```{r}
#| label: fig-eval
#| fig-cap: "Each method's estimated coefficient against each species' own analytically-derived isometric-relative reference quantity, one point per simulated species. Colour indicates shape-response category and point shape indicates allometric scaling category. Grey lines mark zero on each axis, the classification boundary. Proportionally smaller species sit exactly on the vertical zero line by construction. r (Pearson correlation with the reference quantity) and the percentage of species in agreement, pooled across all three categories shown in each panel, are annotated in each panel. A species counts as in agreement if its 95% confidence interval falls on the correct side of zero (Longer, Fatter), or if that interval includes zero (Proportionally smaller)."
#| fig-width: 9
#| fig-height: 6.5
#| fig-pos: 'H'

Eval_scatter_df <- Parms_tbl4 %>%
  filter(Temp_eff != "Proportionally larger") %>%
  mutate(Model = str_replace_all(Model, x_labs))

# Includes Longer, Fatter, and Proportionally smaller -- "correct" uses Est_correct (Run_simulation.R,
# Evaluation section) throughout: whole-CI-on-the-correct-side for Longer/Fatter, CI-includes-zero
# (reversed) for Proportionally smaller, since its true effect is exactly 0 by construction.
Eval_scatter_stats <- Eval_scatter_df %>%
  group_by(Model) %>%
  summarize(r = round(cor(beta_iso_true, b_temp_inc), 2),
            pct_correct = round(mean(Est_correct) * 100, 1),
            .groups = "drop")

ggplot(Eval_scatter_df, aes(x = beta_iso_true, y = b_temp_inc, color = Temp_eff, shape = Scaling)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_point(alpha = .55, size = 1.2) +
  geom_text(data = Eval_scatter_stats, aes(label = paste0("r=", r, ", ", pct_correct, "% in agreement")),
            x = -Inf, y = Inf, hjust = -0.05, vjust = 1.3, inherit.aes = FALSE, size = 2.6) +
  facet_wrap(vars(Model), scales = "free_y") +
  labs(x = expression(beta[T] ~ "(isometric-relative reference, per species)"),
       y = expression(hat(beta)[T] ~ "(estimated)"),
       color = "Category", shape = "Scaling") +
  guides(color = guide_legend(nrow = 1), shape = guide_legend(nrow = 1)) +
  theme(legend.position = "top", legend.box = "vertical")
```

## Empirical case studies

```{r}
#| label: tbl-species-accounting
#| tbl-cap: "Number of empirical species classified into each size / shape-shifting category (@tbl-classification), by dataset. Species exhibiting Bergmann's and Inverse Bergmann's rules are shown in the main text (@fig-empirical). Mixed - Wingier, Mixed - Fatter, and Stable species are reported in the Supplementary Information."

dir_order <- c("Bergmann's", "Inverse Bergmann's", "Mixed - Wingier", "Mixed - Fatter", "Stable")

Species_accounting_tbl <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(Study, species, Direction) %>%
  count(Study, Direction) %>%
  pivot_wider(names_from = Direction, values_from = n, values_fill = 0) %>%
  mutate(Study = case_match(Study,
                             "Nightjar" ~ "Skinner et al (2025)",
                             "Weeks (2020)" ~ "Weeks et al (2020)",
                             "Atlantic birds" ~ "Rodrigues et al (2019)"),
         Study = factor(Study, levels = c("Skinner et al (2025)", "Weeks et al (2020)", "Rodrigues et al (2019)"))) %>%
  arrange(Study) %>%
  mutate(Study = as.character(Study)) %>%
  dplyr::select(Study, any_of(dir_order)) %>%
  mutate(Total = rowSums(across(any_of(dir_order))))

n_species_total <- sum(Species_accounting_tbl$Total)

Species_accounting_tbl <- bind_rows(
  Species_accounting_tbl,
  Species_accounting_tbl %>%
    summarise(Study = "Total", across(where(is.numeric), sum))
)

knitr::kable(Species_accounting_tbl,
             col.names = c("Study", "Bergmann's", "Inverse Bergmann's",
                           "Mixed - Wingier", "Mixed - Fatter", "Stable", "Total"),
             align = "c")
```

Across the three empirical datasets, `r n_species_total` species were classified into one of the five size / shape-shifting categories (@tbl-classification; @tbl-species-accounting).

```{r}
#| label: rank-consistency-empirical
#| include: false

Rank_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(species, Study, Direction, rank_consistent) %>%
  filter(Direction %in% c("Bergmann's", "Inverse Bergmann's"))

Rank_summary_dir <- Rank_all %>%
  group_by(Direction) %>%
  summarize(n = n(), n_consistent = sum(rank_consistent),
            pct = round(100 * mean(rank_consistent)), .groups = "drop")

n_rank_total     <- nrow(Rank_all)
pct_rank_overall <- round(100 * mean(Rank_all$rank_consistent))

pull_rank_stat <- function(dir, col) {
  Rank_summary_dir %>% filter(Direction == dir) %>% pull({{ col }})
}

pct_rank_bergmanns     <- pull_rank_stat("Bergmann's", pct)
n_rank_bergmanns       <- pull_rank_stat("Bergmann's", n)
pct_rank_inv_bergmanns <- pull_rank_stat("Inverse Bergmann's", pct)
n_rank_inv_bergmanns   <- pull_rank_stat("Inverse Bergmann's", n)

fig_cap_empirical <- glue::glue("Estimated effect ($\\hat{{\\beta}}$) of the environmental gradient on relative wing length across three empirical avian datasets (Nightjars: red labels; North American migrants: blue labels; Atlantic Forest birds: green labels), grouped by direction of body size change: (a) Bergmann's ($n$ = {n_rank_bergmanns} species) and (b) Inverse Bergmann's ($n$ = {n_rank_inv_bergmanns} species). Only species with a significant ($p < 0.05$) mass or wing response are shown. Points are mean parameter estimates ($\\hat{{\\beta}}$) and error bars are 95% confidence intervals. The y-axis label matches Figure 3's to indicate that both figures show the same quantity, temperature's estimated effect on relative appendage length. A positive slope indicates birds become relatively longer with increasing temperature (or through time). A negative slope indicates they become relatively fatter. Asterisks mark species for which the rank order is consistent with simulation results (Bergmann's: Ratio > SLI > OLS; Inverse Bergmann's: Ratio < SLI < OLS).")
```

The rank-order pattern from simulations was generally reproduced in all three datasets: ratio methods consistently produced the most positive estimates and multiple regression / OLS residuals the most negative, with SLI falling between these extremes. `r pct_rank_overall`% of the `r n_rank_total` species with a directional (Bergmann's or Inverse Bergmann's) classification showed the rank order predicted by the simulations (`r pct_rank_bergmanns`% of `r n_rank_bergmanns` Bergmann's species and `r pct_rank_inv_bergmanns`% of `r n_rank_inv_bergmanns` Inverse Bergmann's species, @fig-empirical).

```{r}
#| label: fig-empirical
#| fig-cap: !expr fig_cap_empirical
#| fig-pos: 'H'
#| fig-height: 6

emp_img  <- png::readPNG("Figures/Empirical_combined.png")
emp_grob <- grid::rasterGrob(emp_img, width = grid::unit(1, "npc"),
                              height = grid::unit(1, "npc"))
cowplot::ggdraw() + cowplot::draw_grob(emp_grob)
```

### Correlation between the wing/mass ratio and body mass

```{r}
#| label: ratio-mass-correlation-empirical
#| include: false

Ratio_mass_cor_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_ratio_mass_cor.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_ratio_mass_cor.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_ratio_mass_cor.csv",  show_col_types = FALSE)
) %>%
  dplyr::select(Study, species, Metric, n, r, p_value)

Ratio_mass_summary_tbl <- Ratio_mass_cor_all %>%
  summarize(N_species = n(), Mean_r = mean(r), Median_r = median(r),
            Min_r = min(r), Max_r = max(r), .by = c(Study, Metric)) %>%
  bind_rows(
    Ratio_mass_cor_all %>%
      summarize(Study = "All studies combined", N_species = n(),
                Mean_r = mean(r), Median_r = median(r),
                Min_r = min(r), Max_r = max(r), .by = Metric)
  ) %>%
  mutate(across(c(Mean_r, Median_r, Min_r, Max_r), \(x) round(x, 2)),
         Metric = factor(Metric, levels = c("Ratio", "Ratio2"))) %>%
  arrange(Study, Metric)

Ratio_mass_species_tbl <- Ratio_mass_cor_all %>%
  arrange(Study, species, Metric) %>%
  mutate(r = round(r, 2),
         p_value = ifelse(p_value < 0.001, "< 0.001", as.character(round(p_value, 3))))

# Simple ratio (A/S) -- the metric the surrounding prose already refers to
Ratio_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio")
n_rmc_total        <- nrow(Ratio_cor_only)
mean_rmc_all       <- round(mean(Ratio_cor_only$r), 2)
range_rmc_all      <- paste0(round(min(Ratio_cor_only$r), 2), " to ",
                             round(max(Ratio_cor_only$r), 2))
pct_rmc_strong_neg <- round(100 * mean(Ratio_cor_only$r < -0.7))

# Squared ratio (A^2/S) -- quantifies how much squaring the appendage term
# weakens (but does not eliminate) the correlation with mass.
Ratio2_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio2")
mean_rmc2_all       <- round(mean(Ratio2_cor_only$r), 2)
range_rmc2_all      <- paste0(round(min(Ratio2_cor_only$r), 2), " to ",
                              round(max(Ratio2_cor_only$r), 2))
pct_rmc2_strong_neg <- round(100 * mean(Ratio2_cor_only$r < -0.7))
```

Across `r n_rmc_total` species, the ratio metric was consistently and strongly negatively correlated with mass (mean $r$ = `r mean_rmc_all`, range `r range_rmc_all`). `r pct_rmc_strong_neg`% of species had a correlation more negative than $-0.7$. This pattern held regardless of taxon, gradient type (spatial or temporal), or geographic region, confirming that the ratio's dependence on body size is a general mechanical property of the metric rather than a quirk of any one dataset (@tbl-ratio-mass-summary). I calculated the same correlation for the squared-ratio metric ($\log(A^2/S)$) - the correlation between $\log(A^2/S)$ and body mass was still strongly negative, but consistently weaker in magnitude than its unsquared counterpart (mean $r$ = `r mean_rmc2_all`, range `r range_rmc2_all`, `r pct_rmc2_strong_neg`% of species below $-0.7$, @tbl-ratio-mass-summary).

```{r}
#| label: tbl-ratio-mass-summary
#| tbl-cap: "Summary of the per-species Pearson correlation between the wing/mass ratio metric ($\\log(A/S)$) and its squared counterpart ($\\log(A^2/S)$) and body mass, by empirical dataset."

Ratio_mass_summary_tbl %>%
  filter(Study != "All studies combined") %>%
  mutate(Metric = recode(Metric, "Ratio" = "Wing / Mass", "Ratio2" = "Wing² / Mass"),
         Range = paste(Min_r, "to", Max_r)) %>%
  dplyr::select(Study, Metric, Median_r, Range) %>%
  knitr::kable(
    col.names = c("Study", "Metric", "Median", "Range")
  )
```

# Discussion

Body-size adjustment is not a neutral statistical preprocessing step [as in @rydingMultipleRegressionNot2025] – it is a modelling decision that defines relative appendage length against a particular reference relationship, and different reference relationships estimate different quantities. Across simulations spanning a realistic range of allometric relationships and temperature responses, the six methods examined here diverged systematically and predictably from the isometric-relative reference quantity and from one another: ratio-based methods tended to diverge in the positive direction, whereas OLS-based approaches diverged in the negative direction. The two SLI parameterisations track two quantities that coincide closely in most scenarios but separate cleanly whenever a species' allometry departs from isometry. I observed the same pattern of systematic divergence among methods across three independent empirical datasets spanning diverse ecosystems and spatiotemporal gradients. Together, these results suggest that many published estimates of climate-associated shape change may depend as much on which quantity the chosen analytical approach targets as on the underlying temperature-associated morphological relationships. More broadly, they highlight that quantifying relative appendage size depends on assumptions about how body size and appendage size should be related and how variation between them should be partitioned, spanning scaling, regression geometry, and the implicit directional interpretation each method carries. Making the standardization's reference relationship explicit, as SLI does, clarifies which specific question a given estimate actually answers, rather than treating body-size adjustment as an interchangeable preprocessing step. The following sections focus on species that conform to Bergmann's rule, although the underlying rationale can be inverted for species exhibiting inverse Bergmann's rule.

The four sections below trace this same conceptual progression through each family of methods in turn. Ratios impose a fixed algebraic scaling relationship, without independent biological justification, on which appendage length is standardized. OLS approaches, meaning multiple regression and the closely related residual-regression method, instead estimate a directional, conditional scaling relationship from the data and attribute the remaining variation to temperature. SLI-isometry defines the reference scaling as geometric/isometric a priori, independent of the data. SLI-estimated instead estimates that reference scaling empirically, via SMA. These four are not merely alternative computational routes to one shared quantity: each defines relative appendage size against a different reference relationship, and the magnitude, and in some categories even the sign, of the inferred temperature effect depends on which reference relationship is chosen.

## Ratios diverge positively from the isometric-relative reference

Ratio-based approaches, particularly $A/S$, diverged from the isometric-relative reference quantity in the positive direction in simulations, often by a large margin. Although ratio variables are commonly interpreted as measures of relative appendage length, they frequently remain strongly correlated with body size itself [@greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996]. Across all three empirical datasets, the wing-to-mass ratio was highly correlated with body mass, indicating that it retained much of the same information as its denominator. Consequently, environmental changes in body size were propagated directly into the response variable, making it difficult to distinguish changes in relative appendage size from changes in overall size. The converse occurred under inverse Bergmann's rule, where increasing body mass caused the ratio to diverge from the reference in the negative direction (Supplementary Figure X). In both cases, the ratio inherited its morphological signal from body size rather than isolating relative appendage size.

The comparatively close tracking of the squared-ratio metric $A^2/S$ is also informative. Unlike the simple ratio, it is motivated by geometric arguments relating surface area to volume and therefore partially accounts for allometric scaling [@dubinerWidespreadRecentChanges2022; @nwaoguTemperatureAridityDetermine2018]. More generally, ratio methods differ in the allometric relationship they assume between appendage length and body size. However, these assumptions are fixed by the algebraic form of the index rather than being specified as a biological hypothesis or estimated from the data. In contrast, SLI explicitly standardizes appendage length to a common body size using either the theoretical expectation under isometry (SLI-isometry) or the observed allometric relationship (SLI-estimated). That both SLI approaches track the isometric-relative reference more closely than the ratio methods do therefore illustrates that making the standardization's allometric assumption explicit, rather than fixing it implicitly through the algebraic form of a ratio, clarifies what quantity is actually being estimated [@nakagawaDivideConquerSize2017].

GPT - More generally, ratio methods define relative appendage length through a fixed algebraic relationship between appendage and body size. For example, the simple appendage-to-mass ratio implies a scaling exponent of 1, whereas the squared ratio implies an exponent of 0.5. These exponents may represent biologically motivated hypotheses about scaling, but the relationship is embedded in the algebraic form of the index rather than estimated from the observed appendage–body-size relationship. Ratios also standardize each individual using its own body size, rather than explicitly fitting a bivariate scaling relationship and projecting individuals to a common reference body size. In contrast, SLI makes the reference relationship explicit and uses SMA to characterize the relationship between appendage and body size, with the exponent either specified a priori under isometry (SLI-isometry) or estimated from the data (SLI-estimated). This produces a size-standardized appendage measure in the original units and allows the reference relationship itself to be defined independently of the algebraic form of an index.

## Multiple regression conditions on a directional statistical relationship

Multiple regression attempts a partitioning goal similar to SLI-estimated's: both try to separate variation in appendage length associated with body size from a residual, temperature-associated component. The model $Appendage \sim \beta_T \cdot Temperature + \beta_M \cdot Mass$ defines the body-size relationship through an OLS conditional statistical relationship, assigning all residual scatter around the fitted line to appendage length, and interprets $\beta_T$ as the remaining, mass-conditional association with temperature. SLI-estimated defines the same relationship symmetrically instead, via SMA regression, and standardizes individuals to a common reference body size. The two approaches therefore attempt much the same partition of morphological variation. What differs is *how* the body-size relationship is defined: directionally for OLS, symmetrically for SMA. It is this difference in definition, not a difference in what each approach is trying to accomplish, that produces their diverging estimates.

Whether the OLS conditional relationship also corresponds to a directional causal relationship between mass and appendage is a separate question from how the two variables are statistically related. Interpreting $\hat{\beta}_T$ causally, as the direct effect of temperature on appendage length, requires a causal structure in which body mass causally influences appendage length ($Mass \rightarrow Appendage$). Alternative causal structures could produce the same observed statistical association but different interpretations of the conditional temperature coefficient. The covariance-based simulation used here deliberately specifies the statistical association between mass and appendage without specifying its causal origin. Consequently, it can evaluate how conditioning on mass behaves statistically, but cannot determine whether the resulting temperature coefficient represents a causal direct effect. The distinction between direct and size-mediated effects under an explicit causal model is developed in the companion paper.

Under the covariance structure used in these simulations, the extent to which conditioning on mass displaces $\hat{\beta}_T$ from the isometric-relative reference quantity depends on the relative strength of the temperature–mass and temperature–appendage associations. The displacement is greatest when the correlation between mass and appendage is low, as the model attributes less variance in appendage length to mass and more to temperature (@fig-neg-bias-mvr). In species decreasing in size proportionally, stronger temperature-morphology correlations produce a more negative $\hat{\beta}_T$ (@fig-neg-bias-mvr, panel a). In species simulated to get relatively longer, the negative displacement is greatest at more negative temperature-appendage correlations, and when the effect of temperature on mass is relatively large compared to its effect on appendage (i.e., low 'Strengths' in @fig-neg-bias-mvr, panel b).

```{r}
#| label: fig-neg-bias-mvr
#| fig-cap: "Magnitude of the negative bias in multiple regression (mass-as-covariate) as a function of correlation strength. (a) In species simulated to decrease in appendage length proportionally with mass (Proportionally smaller), the negative bias grew more extreme as the temperature-morphology correlation became more negative and as the correlation between appendage and mass decreased. Allometric scaling (point shape) did not affect the magnitude of bias. (b) In species simulated to become relatively longer, the negative bias was greatest at more negative temperature-appendage correlations. Point colour shows the difference between temperature's effect on wing and its effect on mass (Strength). Point size shows the appendage-mass correlation, confirming that the bias is more extreme when this correlation is low."
#| fig-width: 10
#| fig-height: 4.5
#| fig-pos: 'H'

p_bias_prop <- Parms_tbl4 %>%
  filter(Temp_eff == "Proportionally smaller" & Model == "Ryding") %>%
  ggplot(aes(x = r_13, y = b_temp_inc, color = factor(r_12))) +
  geom_jitter(aes(shape = Scaling), alpha = .6, size = 3, width = 0.02) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, formula = y ~ poly(x, 2)) +
  labs(x = "Temperature-morphology correlation", y = expression(hat(beta)[T]),
       color = "Mass-\nappendage\ncorrelation") +
  guides(shape = "none")

p_bias_longer <- Parms_tbl4 %>%
  filter(Temp_eff == "Longer" & Model == "Ryding") %>%
  ggplot(aes(x = r_13, y = b_temp_inc)) +
  geom_point(aes(color = Strength, size = r_12), alpha = .6) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, formula = y ~ poly(x, 2)) +
  labs(x = "Temperature-appendage correlation", y = expression(hat(beta)[T]),
       size = "Mass-\nappendage\ncorrelation", color = "Strength\nAllen's")

ggarrange(p_bias_prop, p_bias_longer, labels = "auto")
```

## SLI standardizes against an explicit reference relationship

```{r}
#| label: fig-ols-sma
#| fig-cap: "The allometric relationship between log-mass and log-appendage for a simulated hyperallometric species in which temperature increases produce relatively longer appendages. The SMA regression line (blue) is steeper than the OLS line (red). Individual points are coloured by temperature. Sections labelled I and II highlight regions where OLS and SMA assign opposite signs to residuals: small-bodied individuals (Section I) are classified as having relatively short appendages by OLS but relatively long appendages by SMA. Large-bodied individuals (Section II) are classified in the opposite direction. As temperature increases select for smaller body size (Bergmann's rule), individuals are displaced toward Section I, exactly where OLS assigns more negative residuals than SMA, producing a spurious negative $\\hat{\\beta}_T$."
#| fig-pos: 'H'
#| fig-height: 4.5

Ex_parms_hypo_hyper <- Parms_mat3 %>%
  filter(Scaling != "Isometry" & Temp_eff == "Longer" & r_12 == r_12[1]) %>%
  slice_max(n = 1, order_by = Strength, by = c(Temp_eff, Scaling), with_ties = FALSE)

set.seed(20260710)
Ex_df_hypo_hyper <- gen_ex_data(Ex_parms_hypo_hyper) %>%
  filter(Scaling == "Hyperallometry") %>%
  mutate(Temp_inc = Temp_inc / 3)

xy_lims <- Ex_df_hypo_hyper %>%
  summarize(max.append = max(Append_log), min.append = min(Append_log),
            max.mass   = max(Mass_log),   min.mass   = min(Mass_log))

ggplot(Ex_df_hypo_hyper, aes(x = Mass_log, y = Append_log)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ggpmisc::stat_ma_line(method = sma_or_ma, color = "blue") +
  geom_point(alpha = 0.4, aes(color = Temp_inc)) +
  annotate("text", x = 4.23, y = 5.13, label = "I",  size = 6, color = "black") +
  annotate("text", x = 4.52, y = 5.25, label = "II", size = 6, color = "black") +
  scale_color_viridis_c(option = "plasma") +
  lims(y = c(xy_lims$min.append, xy_lims$max.append),
       x = c(xy_lims$min.mass,   xy_lims$max.mass)) +
  labs(x = "Log(mass)", y = "Log(appendage)", color = "Temperature\nincrease")
```

SMA regression produces a steeper slope than OLS whenever $|r| < 1$, because it minimizes deviations in both dimensions rather than only vertically [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009]. The consequence for OLS-based measures of relative appendage length is illustrated in @fig-ols-sma. Because the OLS slope is shallower, OLS and SMA assign different relative appendage values to individuals at different body sizes. When temperature is associated with declining body size, as under Bergmann's rule, the temperature gradient therefore shifts observations systematically toward region I, where OLS assigns more negative residuals than SMA.

The SLI inherits two additional useful properties from the SMI [@peigNewPerspectivesEstimating2009]. First, the resulting index retains the units of the original measurement (millimetres), which facilitates biological interpretation. Second, the reference body size $M_0$ enables comparison of individuals across populations and studies: by standardising to the same body size, SLI values from different populations are directly comparable, unlike OLS residuals or ratio values whose scale depends on the particular dataset.

Importantly, these results are conditional on the assumed data-generating process. The covariance structure is compatible with a biological system in which mass and appendage size are coordinated expressions of an underlying latent phenotype (e.g., body size), rather than a predictor-response relationship. Under this causal structure, SMA is expected to recover the allometric scaling relationship more faithfully than OLS. Conversely, if a causal model in which mass influences appendage size is biologically appropriate, OLS-based adjustment may be more appropriate for estimating a conditional/direct effect. In empirical morphological datasets, however, the causal relationship between mass and appendage size is often uncertain. Under these circumstances, SLI approaches provide a complementary framework that avoids specifying a directional statistical relationship between body size and appendage size, and may therefore be useful when the causal structure underlying that relationship is uncertain.

## Estimating versus assigning SLI scaling exponents

The two SLI parameterisations do not simply differ in which allometric exponent they happen to use. They differ in which of two mathematically distinct components of a nonzero isometric-relative quantity they are structurally capable of detecting. Rewriting the isometric-relative reference quantity as $\beta_T \propto \rho_{23}(\beta_{SMA} - 0.33) + \beta_{SMA} \cdot \Delta\rho$ shows that a species can depart from $\beta_T = 0$ through two independent components (@fig-two-pathways). The first is an **allometry-vs-isometry component**, in which the species' own SMA slope simply differs from the isometric exponent even with no differential temperature association between mass and appendage. The second is a **differential temperature-association component**, in which temperature is differentially associated with mass and appendage ($\Delta\rho \neq 0$), the classical Bergmann's/Allen's-rule mechanism. For SLI-isometry, whose fixed exponent ($b_{SLI} = 0.33$) generally differs from a species' true SMA slope, both terms are typically nonzero. A significant SLI-isometry coefficient can therefore arise from either component, or a mixture of both, and the coefficient alone cannot distinguish which. For SLI-estimated, whose exponent is by definition the species' own SMA slope, the allometry-vs-isometry term vanishes identically ($\beta_{SMA} - \beta_{SMA} = 0$), so only the differential temperature-association component can produce a nonzero SLI-estimated coefficient. This is a structural property of the two formulas, not a difference in estimation performance, and it is precisely why the two approaches diverge in the Proportionally smaller/larger categories: those species were simulated with zero differential temperature association by construction, so any signal SLI-isometry detects there comes from the allometry-vs-isometry component alone, a different kind of departure from proportional scaling than the differential temperature-association mechanism classically associated with Allen's rule. Reporting both metrics together, rather than treating either alone as "the" Allen's-rule estimate, therefore conveys more information than either can on its own.

```{r}
#| label: fig-two-pathways
#| fig-cap: "Conceptual illustration of two components contributing to an isometry-relative temperature effect, for the same simulated hyperallometric species shown in @fig-ols-sma. Points are a representative subset of individuals, coloured by simulated temperature increase; warmer individuals are smaller in body size and are also displaced relative to the species-specific SMA relationship. The solid blue line is the empirical SMA relationship between appendage length and body mass ($b_{SMA}$, its estimated slope); the dashed grey line is the isometric reference ($b_{SMA} = 0.33$ by definition). The white-filled point marks the centroid of the warmest 15% of individuals. The green double-headed arrow shows the **allometry-isometry component**: the difference between the two reference relationships at a representative body size, which arises because this hyperallometric species' own scaling differs from isometry, even for a point lying exactly on the SMA line. The red arrow continues from where the green arrow meets the isometric line and shows the **differential temperature-association component**: the additional vertical displacement associated with temperature, measured from the isometric line. SLI-isometry, which standardizes against the isometric reference, is sensitive to both components; SLI-estimated, which standardizes against the empirical SMA relationship, isolates the differential temperature-association component alone. Neither component is inherently more meaningful than the other -- both are parts of the isometry-relative quantity, and which one matters depends on the inferential target."
#| fig-pos: 'H'
#| fig-width: 8
#| fig-height: 5.5

# SMA and isometric reference lines, computed directly from the slope/intercept
# formulas already introduced in the SMA-intuition callout (Introduction).
r_pathways    <- cor(Ex_df_hypo_hyper$Mass_log, Ex_df_hypo_hyper$Append_log)
sma_slope     <- sign(r_pathways) * sd(Ex_df_hypo_hyper$Append_log) / sd(Ex_df_hypo_hyper$Mass_log)
sma_intercept <- mean(Ex_df_hypo_hyper$Append_log) - sma_slope * mean(Ex_df_hypo_hyper$Mass_log)
iso_slope     <- 1 / 3
iso_intercept <- mean(Ex_df_hypo_hyper$Append_log) - iso_slope * mean(Ex_df_hypo_hyper$Mass_log)

# Centroid of the warmest 15% of individuals -- the representative body size
# (x_anchor) both component arrows are drawn at.
warm_centroid <- Ex_df_hypo_hyper %>%
  filter(Temp_inc >= quantile(Temp_inc, 0.85)) %>%
  summarize(Mass_log = mean(Mass_log), Append_log = mean(Append_log))
x_anchor <- warm_centroid$Mass_log
y_sma_at_anchor <- sma_intercept + sma_slope * x_anchor
y_iso_at_anchor <- iso_intercept + iso_slope * x_anchor

# Differential-temperature-association displacement: fit each individual's
# vertical distance from the SMA line against its own simulated temperature
# increase, then evaluate that fit at a representative warm temperature.
Ex_df_hypo_hyper <- Ex_df_hypo_hyper %>%
  mutate(sma_resid = Append_log - (sma_intercept + sma_slope * Mass_log))
resid_fit     <- lm(sma_resid ~ Temp_inc, data = Ex_df_hypo_hyper)
temp_warm     <- quantile(Ex_df_hypo_hyper$Temp_inc, 0.95, names = FALSE)
resid_at_warm <- unname(predict(resid_fit, newdata = data.frame(Temp_inc = temp_warm)))
y_actual_at_anchor <- y_sma_at_anchor + resid_at_warm

# The two arrows chain vertically at x_anchor: SMA -> isometric line (green,
# allometry-isometry component), then isometric line -> actual (red,
# differential temperature-association component), meeting exactly at the
# isometric line so the total span reads as one additive decomposition.
x_line_end  <- max(Ex_df_hypo_hyper$Mass_log)
y_range     <- diff(range(Ex_df_hypo_hyper$Append_log))
x_label     <- min(Ex_df_hypo_hyper$Mass_log)
y_green_label <- y_sma_at_anchor - 0.30 * y_range
y_red_label   <- y_actual_at_anchor + 0.30 * y_range

set.seed(20260818)
Ex_df_pathways_plot <- Ex_df_hypo_hyper %>% slice_sample(n = 700)

ggplot(Ex_df_pathways_plot, aes(x = Mass_log, y = Append_log)) +
  geom_point(alpha = 0.28, size = 1.6, aes(color = Temp_inc)) +
  geom_abline(intercept = sma_intercept, slope = sma_slope, color = "blue", linewidth = 1) +
  geom_abline(intercept = iso_intercept, slope = iso_slope, color = "grey35", linetype = "dashed", linewidth = 1) +
  geom_point(data = warm_centroid, aes(x = Mass_log, y = Append_log), inherit.aes = FALSE,
             shape = 21, fill = "white", color = "black", size = 2.6, stroke = 1) +
  annotate("segment", x = x_anchor, xend = x_anchor, y = y_sma_at_anchor, yend = y_iso_at_anchor,
           color = "forestgreen", linewidth = 1, arrow = arrow(length = unit(0.18, "cm"), ends = "both")) +
  annotate("segment", x = x_anchor, xend = x_anchor, y = y_iso_at_anchor, yend = y_actual_at_anchor,
           color = "firebrick", linewidth = 1, arrow = arrow(length = unit(0.18, "cm"), ends = "last")) +
  annotate("segment", x = x_label, xend = x_anchor, y = y_green_label, yend = (y_sma_at_anchor + y_iso_at_anchor) / 2,
           color = "forestgreen", linewidth = 0.4) +
  annotate("segment", x = x_label, xend = x_anchor, y = y_red_label, yend = (y_iso_at_anchor + y_actual_at_anchor) / 2,
           color = "firebrick", linewidth = 0.4) +
  annotate("text", x = x_label, y = y_green_label, label = "Allometry-isometry\ncomponent",
           color = "forestgreen", size = 3.3, hjust = 1, fontface = "bold") +
  annotate("text", x = x_label, y = y_red_label, label = "Differential\ntemperature\nassociation",
           color = "firebrick", size = 3.3, hjust = 1, fontface = "bold") +
  annotate("text", x = x_line_end - 0.06, y = sma_intercept + sma_slope * (x_line_end - 0.06) + 0.025,
           label = "SMA estimated", color = "blue", size = 4.2, hjust = 0, fontface = "bold") +
  annotate("text", x = x_line_end - 0.06, y = sma_intercept + sma_slope * (x_line_end - 0.06) + 0.008,
           label = paste0("b[SMA] == ", round(sma_slope, 2)), parse = TRUE,
           color = "blue", size = 3.4, hjust = 0) +
  annotate("text", x = x_line_end - 0.06, y = iso_intercept + iso_slope * (x_line_end - 0.06) - 0.015,
           label = "Isometric reference", color = "grey35", size = 3.6, hjust = 0, fontface = "bold") +
  annotate("text", x = x_line_end - 0.06, y = iso_intercept + iso_slope * (x_line_end - 0.06) - 0.032,
           label = "b[SMA] == 0.33", parse = TRUE, color = "grey35", size = 3.4, hjust = 0) +
  scale_color_viridis_c(option = "plasma") +
  scale_x_continuous(expand = expansion(mult = c(0.42, 0.40))) +
  labs(x = "Log(mass)", y = "Log(appendage)", color = "Temperature\nincrease")
```

This structural difference is also why the two parameterisations are appropriate for different questions. SLI-isometry uses $b_{SLI} = 0.33$, corresponding to the expected isometric scaling of a linear appendage dimension against a volumetric proxy of body size (i.e., mass), and asks whether appendage size departs from that proportional, geometric expectation [e.g., @youngfleshAbioticConditionsShape2022]. This null model ($\text{Appendage length} \propto \text{Mass}^{0.33}$) is intuitively appealing for tests of Allen's rule, however there is little consensus regarding the implications of biological scaling for thermoregulation [@tabhLimitedEvidenceThat2025; @santoroAllometryEvaluateAllens2022; @rydingResponseAllometryEvaluate2022]. Proportional change in linear dimensions alongside a change in overall body size is nonetheless a reasonable default expectation for "no shape change," and recent treatments of Allen's rule adopt essentially this same isometric/geometric null when distinguishing genuine shape-shifting from size change alone [@symondsItsAllRelative2025; @rydingMultipleRegressionNot2025]. This justification does not imply that empirical appendage-mass relationships are, or must be, isometric. It only means that isometry is one legitimate, externally motivated reference point among the several a study might choose. The distinction between theoretically specified and empirically estimated scaling exponents has precedent in the condition-index literature. @peigParadigmBodyCondition2010 emphasized that departures from isometry can make empirical estimation of the scaling exponent biologically informative, while also recognizing that a theoretically specified exponent can provide a useful alternative.

In contrast, SLI-estimated uses the empirically derived SMA slope and asks instead whether appendage size departs from the species' own observed allometric relationship, which is useful if the objective is to characterise deviations from the observed scaling relationship rather than from a theoretical null. For example, it removes systematic shifts in appendage length associated with hyperallometric or hypoallometric scaling, leaving residual variation in relative morphology. It may also improve comparability across groups such as sex or age classes, where SMA slopes can be estimated separately and incorporated into the standardisation, analogous to the scaled mass index used in body condition studies [@peigParadigmBodyCondition2010]. This approach provides flexibility, but may not be feasible in certain datasets, particularly if appendage–mass relationships are sufficiently weak (e.g., we used $r_{A,S} \geq 0.3$ as a threshold for applying SLI-estimated). Ultimately, the choice between SLI parameterisations depends on whether the inferential target is deviation from a theoretical scaling law (SLI-isometry) or deviation from the observed species-specific scaling structure (SLI-estimated), not on which one is more accurate in the abstract.

## Limitations of the SLI approaches

Despite their advantages, the SLI approaches are not without flaws. Unfortunately, SMA regression is inflexible in its ability to customize models because it can only model the two scaling variables $X$ and $Y$. This inflexibility can make it more challenging to meet SMA regression assumptions relative to OLS assumptions [@wartonBivariateLinefittingMethods2006]. For example, SMA regression assumes homoskedasticity, but missing covariates that influence appendage measurements (e.g., feather wear) cannot be included in SMA regression, which could lead to heteroskedasticity. Similarly, if there is still a non-linear relationship after log-log transforming the linear and volumetric measurement, the SMA assumption of linearity is violated. This non-linearity can be easily accomodated in classical regression [e.g., $\mathit{Body\ size} + \mathit{Body\ size^2}$, @baldwinComplementarityAllensBergmanns2023], but not in SMA regression. In these cases, regression approaches that accommodate additional covariates may be preferable.

SLI-estimated's reference relationship is also sample-dependent, since it is defined by the observed SMA slope rather than a fixed exponent. Consequently, its interpretation can depend on sample composition, including sex and age structure, geographic or population structure, the range of body sizes sampled, and potentially the distribution of temperature across the sampled individuals. SLI-estimated should therefore be interpreted as deviation from the observed allometric relationship, rather than as a universal measure of biological shape change.

An important next step will be to evaluate whether SLI-derived measures correspond to biologically meaningful aspects of morphology or function. @peigNewPerspectivesEstimating2009, for example, found that the Scaled Mass Index predicted variation in fat and protein reserves better than several conventional condition indices across multiple vertebrate taxa. Analogous validation of SLI against independent morphological measures, including geometric morphometric or three-dimensional measurements that more directly quantify appendage surface area and body volume, could test whether particular SLI parameterizations capture biologically meaningful variation in relative appendage size.

## Recommendations for empirical studies

Body-size adjustment is not simply a statistical preprocessing step, but a modelling decision that encodes assumptions about the biological relationship between body size, appendage size, and the environmental gradient of interest. The simulations presented here demonstrate that different adjustment methods target different estimands and can therefore lead to contrasting biological conclusions. Rather than advocating a universally superior method, I recommend that researchers explicitly match their analytical approach to their biological assumptions and inferential objectives.

1.  **Define the inferential objective and, when relevant, state the causal assumptions.** No single method will recover temperature's influence on relative appendage length under every plausible biological system. Researchers should first distinguish whether their goal is to describe a pattern of relative morphological change or to estimate a causal effect. For descriptive analyses, the appropriate standardization method should be chosen according to the morphological quantity of interest. Causal analyses additionally require an explicit causal model and assumptions sufficient to identify the target effect. Causal graphical models provide a useful framework for making these assumptions explicit and identifying situations in which conditioning on body size may induce rather than remove bias [@dudneyCausalInferenceFramework2024].
2.  **Avoid unvalidated ratio metrics.** Ratio methods define a fixed allometric relationship between appendage length and body size, typically without making this assumption explicit or providing a biological justification [@nakagawaDivideConquerSize2017]. Consequently, they may fail to remove allometric variation when the assumed scaling relationship does not reflect the biology of the study system. Simple appendage-to-mass ratios ($A / M$) assume linear scaling and generally retain substantial correlation with body size. Squared ratios ($A^2 / M$) assume a scaling exponent of 0.5, which is meant to approximate the body surface area relative to volume [e.g., @dubinerWidespreadRecentChanges2022; @nwaoguTemperatureAridityDetermine2018]. The validity of this approximation will be highly dependent on organismal shape, and new 3D modeling approaches may replace the need for these rough approximations in the near future [e.g., CITE Adamczack 2019; @rydingMeasuringAvianBill2025].
3.  **Interpret multiple regression coefficients in light of the assumptions required for their causal interpretation.** Including body size as a covariate defines a directional conditional relationship between body size and appendage size. If that relationship is interpreted causally as body size influencing appendage size, and the broader causal assumptions required for identification are satisfied, conditioning on body size can isolate a direct effect of temperature. When those assumptions are violated, conditioning on body size may instead introduce bias or estimate a different causal quantity. Multiple regression should therefore be viewed as a model corresponding to a specific biological hypothesis rather than as a default method of body-size adjustment.
4.  **Use SLI when the directional relationship between body size and appendage size is uncertain.** Unlike multiple regression, the SLI does not require specifying a directional conditional relationship between mass and appendage length. It does, however, still require its own assumptions: SLI-estimated assumes a symmetric allometric (SMA) relationship between mass and appendage, and SLI-isometry additionally imposes a fixed isometric exponent. Making these assumptions explicit, rather than avoiding them altogether, is what allows the reference relationship to be a stated choice rather than an implicit one.
5.  **Choose the SLI exponent according to the biological question**. SLI-isometry ($b=0.33$) targets deviation from the theoretical isometric expectation, whereas SLI-estimated targets deviation from the species' own observed allometric relationship. These are two distinct, legitimate inferential targets rather than competing estimates of one quantity, so the choice should follow from which question is being asked, not from which produces the larger or more "significant" effect. SLI-estimated additionally requires a sufficiently strong appendage-body size correlation (approximately $r \geq 0.3$) and adequate sample size (approximately $n \geq 100$ per species or group) to be feasible at all.
6.  **Evaluate sensitivity to methodological choice.** Because the inferred magnitude and even direction of Allen's rule depended strongly on the adjustment method, analyses should not rely exclusively on a single index. Whenever feasible, authors should compare conclusions across multiple approaches (e.g., SLI-isometry, SLI-estimated, and multiple regression) and discuss whether biological inference is robust to reasonable alternative assumptions.
7.  **Report the underlying allometric relationship.** At minimum, report the estimated allometric scaling exponent and the correlation between appendage size and body size. These quantities facilitate comparison among studies, allow readers to assess the appropriateness of different adjustment methods, and enable future reanalysis under alternative modelling assumptions.

# Conclusions

The method used to quantify relative appendage length has large and predictable effects on biological conclusions in studies of Allen's rule, because different methods encode different assumptions about the scaling relationship between appendage length and body size and consequently estimate different quantities. Ratios fix this relationship implicitly through their algebraic form. Multiple regression and OLS residuals define it through a directional conditional relationship between mass and appendage. That directional relationship can be given a causal interpretation under additional biological and identification assumptions, but the statistical model itself does not establish those assumptions. The two SLI parameterisations make the reference relationship explicit instead, standardizing either against a theoretical isometric expectation (SLI-isometry) or against the species' own observed allometric relationship (SLI-estimated). Across simulations spanning realistic biological conditions, these six approaches diverged from one another and from the isometric-relative reference quantity in systematic, predictable ways, and the same pattern of divergence was reproduced across three independent empirical datasets.

The two-component decomposition developed here shows why this divergence occurs. An apparent isometry-relative temperature effect can combine an allometry-vs-isometry component, arising whenever a species' own scaling departs from the isometric reference, with a differential temperature-association component, arising when temperature is differentially associated with mass and appendage. SLI-isometry retains both components, while SLI-estimated removes the first by construction. This is a mathematical result about what conventional standardization conflates, not a causal claim about which component reflects a direct effect of temperature on appendage length versus an effect mediated by body size. That causal distinction remains open here, since the simulation does not specify a causal structure among temperature, mass, and appendage length.

More broadly, our results suggest that isolating relative appendage length from body size should not be treated as a routine statistical adjustment. Rather, it requires stating which of several legitimate quantities a given analysis is actually targeting: deviation from a theoretical scaling law, deviation from a species' own allometry, or an effect conditional on an assumed causal structure. Different analytical approaches answer subtly different questions because they embody different assumptions about morphology and scaling. Matching the chosen approach to the specific biological question is therefore as important as the biological question itself.

# Data Availability

All analyses can be reproduced using the code and data in the zipped 'for reviewers' file. Nightjar morphological data are available from @skinnerDataEnvironmentalGeographic2025. North American migrants data are available from @weeksSharedMorphologicalConsequences2020. Atlantic Forest bird data are available from @rodriguesATLANTICBIRDTRAITS2019.

# Acknowledgements

I thank Andy Green for helpful comments on the development of the Standardized Length Index and its relationship to the Standardized Mass Index, and David Warton for clarifying the assumptions and properties of SMA regression.

{{< pagebreak >}}

# References
