---
title: "The Shape of Things to Come: Different Estimators Lead to Different Conclusions in Tests of Allen’s Rule"
number-sections: true
execute:
  echo: false
  message: false
  warning: false
format: #docx
  elsevier-pdf:
    toc: false
    header-includes:
      - |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhead[L]{}
        \fancyhead[R]{SLI to examine Allen's rule}
        \usepackage{makecell}
        \usepackage{lineno}
        \linenumbers
        \usepackage[margin=1in]{geometry}
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
        \usepackage{tabularray}
    journal:
      name: Methods in Ecology and Evolution
      formatting: doubleblind #doubleblind suppresses authors!, return to preprint
      cite-style: authoryear
bibliography: ../../Suppfiles/Morphology.bib
link-citations: true
csl: ../../_extensions/quarto-journals/elsevier/bib/elsevier-harvard.csl
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
editor_options:
  chunk_output_type: console
metadata-files: 
  - ../../Suppfiles/abstract_keywords.yml
  - ../../Suppfiles/authors.yml
---

{{< pagebreak >}}

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(png)
library(grid)
library(ggpubr)
library(cowplot)
library(scales)
library(smatr)
library(MASS)
library(ggpmisc)
ggplot2::theme_set(theme_cowplot())
```

```{r}
#| label: load-lit-review-results
#| include: false

# Literature review is processed separately by Scripts/02_Lit_review.R (re-run it any time the underlying spreadsheet or classification logic changes) and saved to Derived/Rds/lit_review_results.rds. This brings Lit_review, Unclear_exclude, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/lit_review_results.rds"), envir = .GlobalEnv)
```

```{r}
#| label: Percents-in-line

Per_class <- Lr %>%
  summarize(Percent = n() / nrow(Lr) * 100, .by = class) %>%
  mutate(Percent = paste0(round(Percent, 0), "%"))

per_aves <- Per_class %>% 
  filter(class == "Aves") %>% 
  pull(Percent)

# Pooled (both classes combined) approach-usage counts/percentages, cited
# inline below and matching the combined @fig-lit-review panel.
pull_pooled <- function(approach_name, stat) {
  Lr_per_pooled %>% filter(approach == approach_name) %>% pull({{ stat }})
}

tot_approach_uses <- unique(Lr_per_pooled$Tot_studies)
n_multi_approach  <- sum(duplicated(Lr_pivot$study))
pct_mv_reg <- pull_pooled("multiple_regression", Per)
pct_raw    <- pull_pooled("raw_values", Per)
pct_other  <- pull_pooled("other", Per)
pct_ratio  <- pull_pooled("ratio", Per)
pct_resid  <- pull_pooled("residuals", Per)

# How many studies had multiple approaches? 
Num_mult_approach <- Lr_pivot %>% 
  count(study, sort = T) %>% 
  filter(n > 1) %>% 
  nrow()

# Studies examining relative (not absolute) appendage length, by whether mass
# was the body-size-control variable used.
n_relative_studies <- sum(Mass_control_tbl$n)
n_mass_control      <- Mass_control_tbl %>% filter(Mass_control == TRUE) %>% pull(n)
pct_mass_control    <- round(100 * n_mass_control / n_relative_studies, 0)

# Studies using an OLS-based approach (multiple regression or residual regression),
# as a percentage of studies examining relative (not just absolute) appendage length --
# same n_relative_studies denominator as the mass-as-body-size-control stat above.
n_ols_studies <- Lr_pivot %>%
  filter(approach %in% c("multiple_regression", "residuals")) %>%
  distinct(study) %>%
  nrow()
pct_ols_relative <- round(100 * n_ols_studies / n_relative_studies, 0)
```

```{r}
#| label: source-key-fns
#| include: false
source("Scripts/00_Key_allometry_fns.R")
```

```{r}
#| label: load-simulation-results
#| include: false

# Simulation is run separately by Scripts/01_Run_simulation.R (re-run it any time the parameter grid, sample size, or evaluation logic changes) and saved to Derived/Rds/simulation_results.rds
# Both this manuscript and supporting_info.qmd just read the result. This brings N_spp, Parms_mat3, Parms_tbl4, Eval_tbl, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/simulation_results.rds"), envir = .GlobalEnv)
```

# Introduction

Body size is strongly correlated with many biological traits, meaning variation in a focal trait or process will often be strongly influenced by overall body size [@debieBodySizeDispersal2012; @vasseur2025organism; @jetzScalingAnimalSpace2004]. The challenge is that ecology and evolution are often interested in questions regarding variation in a trait or process that is independent of overall body size. For example, appendage length influences diverse aspects of organismal performance, including biomechanics, the energetics of locomotion, and resource acquisition, but is strongly correlated with overall body size [@martin-serraThreeDimensionalAnalysisMorphological2014; @demesBiomechanicsAllometricScaling1989; @alexanderAllometryLimbBones1979]. Thus, many biological questions concern whether appendages are disproportionately long or short relative to overall body size.

One prominent example of this is Allen's rule, which predicts that appendages should be relatively shorter in colder environments and relatively longer in warmer environments [@allen1877influence]. Accordingly, testing Allen's rule requires methods that can distinguish variation in appendage size from variation in overall body size [@symondsItsAllRelative2025]. Yet the statistical methods used to quantify relative appendage length have received far less attention than the biological hypothesis itself [reviewed in @rydingShapeshiftingChangingAnimal2021].

There are two common estimators used in studies of Allen's rule: multiple regression and ratios. The multiple regression approach includes body size and temperature (or a proxy, e.g., latitude, year) as covariates, interpreting the slope of temperature (hereafeter, $\hat{\beta}_T$) as temperature's conditional effect on appendage length (after accounting for body size). The ratio approach uses appendage length divided by body size as the response variable. Despite their widespread use, both the multiple regression and ratio approaches have important statistical limitations [@glazierComplicationsBodysizeCorrection2022]. Ratio-based metrics imply a specific scaling ratio that is often not justified by the scaling variables and are frequently correlated with body size itself, limiting their utility as measures of relative shape [@jakobEstimatingFitnessComparison1996; @greenMassLengthResiduals2001]. Multiple regression, on the other hand, makes a key assumption – it assigns all unexplained variation around the fitted allometric line to the $Y$ variable (appendage length in this case). Consequently, every deviation from the fitted line is interpreted as evidence of an unexpectedly long or short appendage, rather than allowing deviations to be shared between both size variables (@nte-sma-intuition). This assumption is often difficult to justify in studies of allometric scaling [@greenMassLengthResiduals2001; @smithUseMisuseReduced2009].

Some have suggested that a framework grounded in allometric scaling theory is necessary for studies of Allen's rule [@santoroAllometryEvaluateAllens2022], although there has been little consensus [@rydingResponseAllometryEvaluate2022]. Allometric scaling theory approaches biological scaling through the equation $Y=aX^b$, where body size $X$ and a second trait $Y$ scale at a rate determined by the exponent $b$ [@huxley1932problems; @shingletonAllometryStudyBiological2010]. Under this framework, relative appendage length can be understood as the deviation of appendage length from an expected allometric relationship with body size. One approach to estimating that allometric relationship is using standardized major axis (SMA) regression, which treats variation in body size and appendage length symmetrically rather than assigning all unexplained variation to the response variable [@wartonBivariateLinefittingMethods2006] (@nte-sma-intuition). Consequently, SMA estimates steeper allometric slopes whenever the correlation between $X$ and $Y$ is $< 1$. SMA has been recommended for studies of allometric scaling [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009; @greenMassLengthResiduals2001] and has been used in studies examining Allen's rule [e.g., @adamczakBiggerBodiesRequire2020]. However, its implications for quantifying relative appendage length have not been systematically compared with the multiple regression and ratio approaches commonly used in studies of Allen's rule. Understanding the statistical behavior of these approaches, and the conditions under which they diverge, is therefore critical for determining when and to what extent methodological choices influence conclusions about climate-associated morphological change.

Here I introduce the Scaled Length Index (SLI), an analogue of the Scaled Mass Index [@peigNewPerspectivesEstimating2009] but adapted to measure appendage length relative to body size. The SLI uses the SMA slope to standardize each individual's appendage length to a common reference body size, producing a size-corrected shape index in the original units of the appendage. I present two parameterisations, corresponding to two distinct questions about relative appendage change. First, SLI-isometry fixes the scaling exponent at 0.33 (the isometric expectation for a linear versus volumetric measurement like mass) and asks whether appendage size departs from that proportional, geometric expectation. Second, SLI-estimated estimates the exponent empirically from the data and asks whether appendage size departs from the species' own observed allometric relationship.

Using simulated data from `r N_spp` hypothetical species spanning a realistic range of allometric relationships and temperature sensitivities, I characterize how these three approach families (SLI metrics, ratios, and OLS-based approaches like multiple regression) diverge from one another under different combinations of allometric scaling and temperature-associated changes in body size and appendage length. I also examine whether the rank-order patterns that emerge during simulation are reproduced across three independent empirical datasets.

::: {#nte-sma-intuition .callout-note icon="false"}
## Building intuition for SMA regression

SMA and OLS regression differ in two closely related ways. First, SMA regression estimates a steeper allometric slope whenever the correlation between variables is \< 1. Second, because it treats variation in both variables symmetrically, relative appendage length and relative body size become reciprocal descriptions of the same deviation from the allometric relationship. The equations below show why these properties arise.

`\vspace{1.5em}`{=latex}

*SMA estimates a steeper allometric slope*

There is a direct mathematical link between the slopes of OLS and SMA regression.\
The slope of a predictor variable $X$ (e.g., body size) is related to the response variable $Y$ (e.g., appendage length) in OLS regression via the following equation:

- $\beta_{\text{OLS}} = r * \frac{Sy}{Sx}$ where $r$ is the correlation between $X$ and $Y$, and $S$ is the standard deviation

The slope of SMA regression is estimated by:

- $\beta_{\text{SMA}} = sign(r) * \frac{Sy}{Sx}$ , i.e., if $r$ is negative then $\beta_{\text{SMA}}$ is also negative.

A simple substitution of $\frac{Sy}{Sx}$ in the first equation shows that

$$\beta_{\text{OLS}} = r * \beta_{\text{SMA}}$$

This makes it evident that as $r$ increases, $\beta_{\text{OLS}} \rightarrow \beta_{\text{SMA}}$, and that the difference in estimated slopes will be highest when $X$ and $Y$ are only weakly correlated (SI Figure 2).

`\vspace{1.5em}`{=latex}

*SMA gives a symmetric geometric interpretation*

Unlike OLS, the SMA line of best fit is unchanged when the $X$ and $Y$ axes are flipped (@fig-sma_flip_axes), leading to a useful geometric property: swapping the axes simply reflects the points across the fitted line. Consequently, the individual highlighted in green can be described either as having relatively short appendages for its body size (panel a) or as having a relatively large body for its appendage length (panel b). These are reciprocal descriptions of the same deviation from the allometric relationship (i.e., residuals retain the same magnitude but reverse sign). In contrast, because the OLS line changes when the axes are exchanged, this reciprocity does not hold.

![In SMA regression the X and Y axes can be flipped and the line of best fit will not change. Here I flip the axes and plot the OLS (red) and SMA (blue) regression lines of best fit, as well as the same individual (highlighted in green) in both panels. SMA residuals are equally correlated with both $X$ and $Y$ ($|r_{sma}| = 0.53$), whereas OLS residuals remain uncorrelated with the predictor but strongly correlated with the response ($r_{ols} = 0.89$).](../../Figures/sma_flip_axes.png){#fig-sma_flip_axes width="100%"}
:::

# Literature review

I reviewed `r nrow(Lit_review)` empirical studies published between 2021 and 2025 that quantified appendage length in the context of Allen's rule in endotherms [i.e., those published in @skinnerBergmannAllenUnified2026]. Of these, `r nrow(Unclear_exclude)` were excluded due to unclear methodologies, leaving `r nrow(Lr)` studies. `r number_to_word(n_multi_approach)` of these studies used multiple approaches to examine Allen's rule, resulting in `r tot_approach_uses` analytical approaches that were classified into five categories: ratio-based methods, two-step residual regression, multiple regression with body size as a covariate, use of absolute appendage length, or other. The 'other' category included shape axes from principal component analyses, path analyses, and geometric morphometric approaches, among other creative solutions [e.g., @youngfleshAbioticConditionsShape2022]. Multiple regression was the most common approach (`r pct_mv_reg`%), followed by absolute appendage length (`r pct_raw`%), ratio-based methods (`r pct_ratio`%), other approaches (`r pct_other`%), and residual regression (`r pct_resid`%; SI Figure 1). Note that absolute appendage length has been compared with methods that explicitly account for body size elsewhere [@symondsItsAllRelative2025], so I do not consider it further. The majority of studies (`r per_aves`) examined Allen's rule in birds. The remainder used mammal focal taxa.

# Materials and Methods

## Data simulation

I simulated appendage length, mass, and a temperature increase for `r N_spp` hypothetical species, each species with `r N_ind` individuals. I simulated mass measurements as a proxy for body size as it was the single most frequent variable (`r pct_mass_control`%) used to control for body size in studies of Allen’s rule. Data were drawn from a multivariate normal distribution:

$$\begin{bmatrix}
\log(\text{Appendage}) \\
\log(\text{Mass}) \\
\text{Temperature}
\end{bmatrix}
\sim \mathrm{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

with $\boldsymbol{\mu} = [\log(180),\ \log(80),\ 1.0]^\top$ representing a species with a mean appendage length of 180 mm and mean mass of 80 g, experiencing an average 1°C temperature increase. The 1°C temperature increase could be imagined spatially or temporally (in the context of climate change) - the statistical implications are the same. The covariance matrix $\boldsymbol{\Sigma}$ was constructed from three Pearson correlations ($\rho_{12}$: appendage–mass; $\rho_{13}$: appendage–temperature; $\rho_{23}$: mass–temperature) and the allometric scaling slope. Hypothetical species were constructed by crossing the following parameter values: appendage–mass correlation $\rho_{12} \in {`r paste(r_12, collapse = ", ")`}$, temperature–morphology correlations $\rho_{13}, \rho_{23} \in {`r paste(r_13, collapse = ", ")`}$, crossed factorially to produce species with different relative temperature associations of mass and appendage, and allometric scaling slope $b \in {`r paste(b_avg_12, collapse = ", ")`}$, representing hypoallometry, isometry, and hyperallometry respectively. Standard deviations of log-mass and log-appendage were set so that the average of the OLS and SMA slopes equaled the specified allometric slope, allowing me to vary allometric scaling independently of the appendage–mass correlation. Correlations were chosen to represent a broad but realistic range observed in the literature: appendage–mass [@greenMassLengthResiduals2001; @goslerFieldDeterminationBody1998] and temperature-morphology correlations [@ashtonPatternsWithinSpeciesBody2002; @ashtonBergmannsRuleValid2000].

## Temperature effect classification

Importantly, when simulating data from a joint multivariate normal distribution rather than from a directional model (e.g., $Appendage \sim Mass + Temperature + \epsilon$), there is no fitted regression coefficient representing a universal "true" effect. To help characterize the statistical behavior of each approach I characterize an isometrically-defined reference quantity, $\beta_T$, as a point of comparison for the various methods. $\beta_T$ captures departures from proportional (isometric) scaling between appendage length and body size:

$$\beta_T = \rho_{13} \cdot \beta_{SMA} - 0.33 \cdot \rho_{23}, \qquad \beta_{SMA} = \frac{2b}{\rho_{12}+1}$$ {#eq-b-true}

where $b$ is the simulated allometric scaling slope, and $\beta_{SMA}$ is the species' population SMA slope implied by its simulated parameters. Importantly, this is only one theoretically motivated definition of relative appendage change, and alternative definitions are possible (e.g., SI Figure 6 shows the same reference quantity defined by OLS scaling). Another consequence of the current definition of $\beta_T$, using the SMA scaling relationship anchored to an exponent of 0.33, is that the SLI-isometry estimator is benefited by design. Indeed, it was built to reflect the theoretical expectation that $A \propto M^{1/3}$ [@shingletonAllometryStudyBiological2010]. The SLI-isometry coefficient $\hat\beta_T$ and the isometric reference quantity $\beta_T$ differ only by a known scaling factor (a variance-correction term; SI Figure 5). Thus, the purpose of the isometric reference quantity is not to identify the universally most accurate method, but to characterize how the six approaches behave relative to a clearly defined reference quantity.

As simulated temperature increased, species with $\beta_T > 0$ were classified as getting relatively longer and $\beta_T < 0$ species were classified as getting relatively stouter.

I additionally simulated a set of species with $\beta_T = 0$ exactly, decreasing in size but with the isometric-relative effect nulled by construction ("Null-effect smaller"). I solved for $\rho_{13}$ algebraically based on a species $\rho_{23}$ and $\beta_{SMA}$, following directly from @eq-b-true:

$$\rho_{13} = \frac{0.33\ \cdot \rho_{23}}{\beta_{SMA}}$$

guaranteeing $\beta_T = 0$ exactly, regardless of allometric scaling category. Stouter, Longer, and Null-effect smaller are all realizations of Bergmann's rule, as warming decreased overall body size in these categories.

Finally, I simulated species that increased in size with warmer temperatures, in line with several recent studies that have documented inverse Bergmannian clines [@stapelfeldtLongtermFieldStudy2023; @castillo-figueroaDoesBergmannsRule2022; @jumanRecentRapidEcogeographical2022]. The 'Null-effect larger' scenario was the mirror image of the 'Null-effect smaller' category, solving $\rho_{13}$ the same way from positive $\rho_{23} \in \{0.15, 0.3, 0.5, 0.7\}$, $b \in \{0.22, 0.33, 0.44\}$, and $\rho_{12} \in \{0.3, 0.45, 0.6\}$, representing species that grow larger with warming while maintaining the isometric-relative reference quantity at zero.

## Statistical approaches

I applied six methods from three families to each simulated species (@tbl-methods-estimands).

**Ratio approaches: (**1) The appendage-to-mass ratio $log(A / S)$ was regressed on temperature. (2) The squared appendage-to-mass ratio $log(A^2 / S)$ was regressed on temperature, reflecting a surface-area-to-volume formulation used in some empirical studies [e.g., @dubinerWidespreadRecentChanges2022; @nwaoguTemperatureAridityDetermine2018]. Importantly, ratios are also sometimes formulated as $log(A) / log(S)$ [e.g., @rydingMultipleRegressionNot2025]; thus, I examined the robustness of my findings to this alternative formulation as well. Given that results were qualitatively similar for both ratio parameterizations (SI Figure 3), and $log(A) / log(S)$ was not unit-invariant (SI Figure 4), I only report the results of $log(A / S)$ in the main text.

**Ordinary least squares (OLS) regression approaches: (**1) A multiple regression approach where appendage was modelled as a linear function of mass and temperature simultaneously. The temperature coefficient $\hat{\beta}_T$ was interpreted as temperature's conditional effect on appendage length, holding mass constant. (2) A two-step residual regression approach was conducted where appendage was regressed on mass using OLS, and the resulting residuals were regressed on temperature.

**Scaled Length Index — estimated (SLI-estimated):** Following @peigNewPerspectivesEstimating2009, I introduce a related parameterization:

$$SLI_i = A_i \cdot \left(\frac{M_0}{M_i}\right)^{b_{SLI}}$$

where $A_i$ is the appendage length of individual $i$, $M_i$ is the mass of individual $i$, $M_0$ is the mean mass of the sampled population, and $b_{SLI}$ comes from the empirical SMA slope of $\log(A)$ on $\log(S)$ estimated from the data [@wartonSmatr3Package2012]. The SLI reprojects each individual's appendage length to the value it would take at the population mean body size (@fig-sli-concept). $SLI_i$ is then regressed on temperature.

**Scaled Length Index — isometry (SLI-isometry):** Identical to SLI-estimated but replacing the estimated $b_{SLI}$ by $b_{SLI} = 0.33$, the isometric scaling exponent for a linear versus volumetric measurement.

I log-transformed all variables and then standardised to unit variance before fitting regression models so that coefficients were comparable across species and methods.

```{r}
#| label: tbl-methods-estimands
#| tbl-cap: "The six statistical approaches compared in this study, the scaling relationship each assumes or estimates between body mass and appendage length, whether that relationship is fixed a priori or estimated from the data, and the resulting measure of relative appendage size. 'Directional' denotes an assumed direction of the mass–appendage relationship (OLS), and 'symmetric' denotes no assumed direction (SMA); neither implies a causal claim."

Methods_estimands_tbl <- tibble::tribble(
  ~Method,                    ~`Scaling relationship`,                          ~`Specification`,            ~`Relative appendage length`,
  "Ratio ($A/S$)",            "$b = 1$",                                        "Fixed",                     "Appendage size relative to linear scaling",
  "Squared ratio ($A^2/S$)",  "$b = 0.5$",                                      "Fixed",                     "Appendage size relative to surface-area scaling",
  "OLS multiple regression",  "$\\hat{\\beta}_{\\mathrm{Mass}}^{\\,\\mathrm{OLS}}$, conditional on temperature", "Estimated (directional)",   "Defined in the context of the model (not standalone)",
  "OLS residuals",            "$\\hat{\\beta}_{\\mathrm{Mass}}^{\\,\\mathrm{OLS}}$", "Estimated (directional)",   "Residual appendage size after OLS adjustment",
  "SLI-isometry",             "$b = 0.33$",                                     "Fixed",                     "Appendage size relative to isometric scaling",
  "SLI-estimated",            "$\\hat{\\beta}_{\\mathrm{Mass}}^{\\,\\mathrm{SMA}}$", "Estimated (symmetric)",     "Appendage size relative to species-specific allometry"
)

knitr::kable(Methods_estimands_tbl, format = "latex", align = "c", booktabs = TRUE, escape = FALSE, linesep = "") %>%
  kableExtra::column_spec(1, width = "2.6cm", latex_valign = "m") %>%
  kableExtra::column_spec(2, width = "3.1cm", latex_valign = "m") %>%
  kableExtra::column_spec(3, width = "3.4cm", latex_valign = "m") %>%
  kableExtra::column_spec(4, width = "4.2cm", latex_valign = "m") %>%
  kableExtra::kable_styling(font_size = 10)
```

### SLI parameterizations: Two distinct estimators

Returning to @eq-b-true, the definition of $\beta_T$ makes it clear that there are two ways a species could depart from $\beta_T = 0$ (the isometric reference): (1) temperature having a differential association with mass vs. appendage ($\rho_{13} \neq \rho_{23}$), or (2) the species exhibiting non-isometric scaling ($\beta_{SMA} \neq 0.33$; @fig-sli-concept). Rewriting $\rho_{13} = \rho_{23} + \Delta\rho$, where $\Delta\rho$ is the difference between $\rho_{13}$ and $\rho_{23}$, decomposes @eq-b-true into these same two distinct components, expressed additively as:

$$\beta_T \;\propto\; \underbrace{\rho_{23}\left(\beta_{SMA} - 0.33\right)}_{\text{allometry-vs-isometry component}} \;+\; \underbrace{\beta_{SMA} \cdot \Delta\rho}_{\substack{\text{differential temperature-} \\ \text{association component}}}$$The two SLI estimators reflect this decomposition. Because SLI-isometry standardizes against the fixed isometric exponent it is capable of detecting departures due to the allometry-vs-isometry component and the temperature-association component. SLI-estimated, on the other hand, standardizes against each species' own SMA slope, effectively removing any impact due to non-isometric scaling. Thus, it is only sensitive to the differential temperature-association component (@fig-sli-concept, panel b). This decomposition is algebraic rather than causal.

```{=html}
<!-- Both panels of @fig-sli-concept are built in Scripts/04a_Concept_figures.R (re-run it any
time either panel's code changes) and saved to Figures/SLI_concept.png; that script also
exports the panel-(b) example species to Derived/Rds/Ex_df_hypo_hyper.rds for fig-ols-sma
(Discussion) to reuse. This script-generated PNG is the reproducible baseline Sylvia's
polished PDF (below) is hand-edited from -- re-run 04a_Concept_figures.R and hand over the
fresh PNG whenever the underlying data/code changes, so her edit stays in sync with the
source. -->
```

![The Scaled Length Index (SLI), adapted from the Scaled Mass Index [@peigNewPerspectivesEstimating2009]. (a) The SLI reprojects each individual's appendage length to the value it would take at the population's mean mass, $M_0$. Two example individuals (filled black circles) are shown sliding along their own standardization curves (dashed black line) to their standardized values (open points). If linearized, the standardization curves would be parallel to the population-level slope (solid grey curve). (b) SLI-estimated and SLI-isometry estimate two different quantities. Points are a representative subset of individuals (log-log scale) for a simulated hyperallometric species whose appendage length increases with warming, coloured by simulated temperature increase. The white-filled point marks the expected phenotype for an individual experiencing a 0.5°C temperature increase. The solid blue line ('Estimated allometry') is the empirical SMA relationship ($\hat\beta_{SMA} = 0.68$), which provides the reference relationship for SLI-estimated; the dashed grey line is the 'isometric reference' ($b = 0.33$), which provides the reference relationship for SLI-isometry. The green arrow shows the allometry-isometry component, i.e., the gap between the two reference lines at this individual's predicted body size. The red arrow shows the temperature-association component, i.e., the additional displacement associated with temperature, measured from the isometric line to this individual's predicted appendage length.](../../Figures/Sylvia/Figure2_SLI_concept_Sept16_edits.pdf){#fig-sli-concept width="100%" fig-pos="H"}

## Statistical approach classification

I evaluated the six statistical approaches using two criteria relative to the isometrically defined benchmark quantity $\beta_T$. First, I assessed whether each approach agreed with the sign of $\beta_T$: a positive $\hat{\beta}_T$ when $\beta_T > 0$, a negative $\hat{\beta}_T$ when $\beta_T < 0$, and $\beta_T \approx 0$ when $\beta_T = 0$. I considered $\hat{\beta}_T$ to agree with the benchmark when its 95% confidence interval excluded zero in cases where $\beta_T \neq 0$, or included zero when $\beta_T = 0$. Second, I assessed how closely $\hat{\beta}_T$ tracked $\beta_T$ in magnitude using a bivariate regression of $\hat{\beta}_T$ on $\beta_T$ across species. An approach that reproduced the benchmark exactly would yield no sign misclassifications, $r = 1$, and estimates falling on the 1:1 line.

## Empirical case studies

```{r}
#| label: n-species-per-dataset
#| include: false

n_nightjar_spp <- read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE) %>% distinct(species) %>% nrow()
n_weeks_spp    <- read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE) %>% distinct(species) %>% nrow()
n_atlantic_spp <- read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE) %>% distinct(species) %>% nrow()
```

I applied all six methods to three independent empirical datasets to assess whether the rank-order patterns observed in simulation were also observed in real data. First, I used morphological data from `r n_nightjar_spp` migratory nightjar species, European nightjars (*Caprimulgus europaeus*), eastern whip-poor-wills (*Antrostomus vociferus*), and common nighthawks (*Chordeiles minor*), spanning a spatial temperature gradient [@skinnerEnvironmentalGeographicConditions2025]. Second, I applied the six methods to a dataset comprising `r n_weeks_spp` North American migrant species (1979–2016), using capture year as a proxy for temperature change [@weeksSharedMorphologicalConsequences2020]. Third, I applied the methods to a dataset of `r n_atlantic_spp` Atlantic Forest bird species spanning a spatial temperature gradient [@rodriguesATLANTICBIRDTRAITS2019]. Together, these datasets span different taxonomic groups, ecological systems, gradients of interest, and putative selective pressures, providing a broad test of whether the same rank-order patterns are observed across empirical contexts.

Age and sex can influence body size independently of temperature (i.e., through sexual size dimorphism or age-related growth). For each species, I therefore tested whether age or sex predicted mass or wing length while controlling for the gradient variable (i.e., temperature or year), retaining a covariate for further consideration only when it significantly predicted body size ($p < 0.10$). Incorporation of demographic covariates varied by the statistical approach (SI section 'Accounting for age and sex in empirical datasets').

To evaluate whether the same rank-order patterns were observed in empirical data, I averaged the gradient coefficients ($\hat\beta_T$) within each method family (Ratio: $A / S$ and $A^2 / S$; OLS: multiple regression and OLS-residual; SLI: SLI-isometry and SLI-estimated) for each species and compared their rank order with the simulation results. While this provides only a coarse assessment of rank-order correspondence, my objective was to identify broad patterns across method families to assess qualitative alignment between simulations and empirical data. A stricter classification criteria examining the rank-order congruence of individual methods returned similar results (see the section 'Rank order without method-family averaging' in the SI). Because the expected ordering depends on the direction of body size change (SI Figure 7), species were first classified according to whether they exhibited Bergmannian, inverse Bergmannian, stable, or mixed size trends (@tbl-classification). Bergmannian species exhibited declines in wing length, body mass, or both along the gradient, whereas inverse Bergmannian species exhibited increases in one or both traits ($p < 0.05$). Stable species showed no significant change in either trait, while mixed species exhibited opposing trends in wing length and body mass. Only Bergmannian and inverse Bergmannian species were used to evaluate agreement with the simulation results because these represent the situations where relative appendage length is particularly sensitive to the method used to account for body size. Species with stable or mixed size trends are reported in SI Figures 8 and 9.

Past authors have noted that ratio methods can remain strongly correlated with body size when the exponent imposed by the ratio does not match the underlying scaling relationship [@greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996]. To test this, and to assess whether SLI's use of an explicit allometric exponent reduces this dependence, I calculated the Pearson correlation between body mass and each of the ratio metrics, SLI-isometry, and SLI-estimated for every species in the three empirical datasets.

```{r}
#| label: tbl-classification
#| tbl-cap: "Classification of species according to significant ($p < 0.05$) changes in wing length and body mass across temperature or year gradients."

Classification_tbl <- tibble::tribble(
  ~Category,               ~`Classification criterion`,
  "Bergmann's",            "Wing length, body mass, or both decrease",
  "Inverse Bergmann's",    "Wing length, body mass, or both increase",
  "Mixed - Wingyier",      "Wing length increases while body mass decreases",
  "Mixed - Stouter",        "Wing length decreases while body mass increases",
  "Stable",                "No significant change in either trait"
)
knitr::kable(Classification_tbl)
```

## Use of AI tools

Claude Code (Anthropic) was used throughout development of the analysis pipeline and manuscript: drafting and debugging R code, proofreading manuscript text, and checking internal consistency (e.g., cross-references, reported statistics) between the code and the manuscript. All code and prose produced with this assistance were reviewed, verified against the underlying data and simulation output, and are the responsibility of the author.

# Results

## Simulation performance

```{r}
#| label: plot-approaches-fun
#| include: false

plot_approaches <- function(df, x_txt_size = 9, legend.pos = "top") {
  df <- df %>%
    filter(Temp_eff != "Null-effect larger") %>%
    mutate(Model = str_replace_all(Model, x_labs))

  ggplot(df, aes(x = Model, y = b_temp_inc)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, height = 0, alpha = .5, size = 1.9,
                aes(color = Strength, shape = Scaling)) +
    facet_wrap(vars(Temp_eff)) +
    labs(x = NULL, y = expression(hat(beta)[T] ~ "on relative appendage length"),
         color = "Strength", shape = "Scaling") +
    guides(shape = guide_legend(nrow = 1)) +
    theme(
      axis.text.x = element_text(size = x_txt_size, vjust = .58, angle = 55),
      legend.position = legend.pos, legend.box = "vertical"
    )
}
```

The six methods produced strikingly different estimates of temperature's effect on relative appendage length, and the pattern was largely consistent across allometric scaling categories and correlation structures (@fig-compare-approaches). The **Null-effect larger** species showed the inverse pattern as the Null-effect smaller species (SI Figure 7).

```{r}
#| label: fig-compare-approaches-source
#| include: false

# Reproducible baseline for @fig-compare-approaches -- Sylvia's polished PDF (embedded below)
# is hand-edited from this exported PDF. Re-run this chunk and hand over the fresh export
# whenever plot_approaches() or Parms_tbl4 changes, so her edit stays in sync with the source.
Fig3_compare_approaches <- plot_approaches(Parms_tbl4, x_txt_size = 11)

dir.create("Figs_share", showWarnings = FALSE)
ggsave("Figs_share/Figure3_compare_approaches.pdf", Fig3_compare_approaches,
       width = 9, height = 4.5, units = "in")
```

![Estimated effect of temperature on relative appendage length ($\hat{\beta}_T$) for each method across three different realizations of Bergmann's rule (Stouter, Null-effect smaller, and Longer species). Positive values of $\hat{\beta}_T$ indicate that species were estimated to become relatively longer with increasing temperature. Negative values indicate they become relatively stouter. Each point is a simulated species. Point shape indicates allometric scaling category and colour indicates the strength of the difference between temperature's effect on appendage length and mass.](../../Figures/Sylvia/Figure3_compare_approaches_Sept16_edits.pdf){#fig-compare-approaches width="100%" fig-pos="H"}
**Ratio approaches** tended to diverge from the isometric-relative reference quantity in the positive direction, although this was much more prominent when $A/S$ was used as the response variable. When simulated to become relatively stouter, $A^2/S$ disagreed with the reference for only `r Ratio2_fatter_wrong_n` of `r Fatter_n` hypothetical species, whereas $A/S$ disagreed for roughly 1/4 of species (@fig-eval). Among species simulated as Null-effect smaller, where $\beta_T = 0$, no species had 95% confidence intervals containing $0$ for either $A/S$ or $A^2/S$.

**Multiple regression and OLS residuals** approaches consistently diverged from the isometric-relative reference quantity in the negative direction (@fig-eval). Both methods returned negative $\hat{\beta}_T$ in the majority of species simulated to become relatively longer, agreed with the reference for Stouter species, and overcorrected in Null-effect smaller species: only `r Ryding_proportional_null_correct`% of these species had a 95% confidence interval that included $0$ for either method, showing statistically significant departures from the null isometric-relative reference quantity in the large majority of null-effect species. The two approaches behaved nearly identically, consistent with their mathematical equivalence when predictor variables are uncorrelated [@freckletonMisuseResidualsEcology2002].

**SLI-isometry** tracked the isometric-relative reference quantity closely, as expected since both are anchored to the same isometric scaling exponent (@fig-eval). Its 95% confidence interval excluded zero on the correct side (matching the reference's sign) for `r Sli.iso_longer_right`% of species simulated to become relatively longer, and `r Sli.iso_fatter_right`% of species simulated to become relatively stouter. Among species simulated as Null-effect smaller, `r Sli.iso_proportional_null_correct`% had a 95% confidence interval that included zero. **SLI-estimated** agreed with the isometric reference quantity in the Longer and Stouter categories (`r Sli.est_longer_right`% and `r Sli.est_fatter_right`% agreement, respectively; @fig-eval), but diverged sharply among Null-effect smaller species, where only `r Sli.est_proportional_null_correct`% of species had a 95% confidence interval that included zero.

```{r}
#| label: fig-eval
#| fig-cap: "Each method's estimated coefficient against its own analytically-derived isometric-relative reference quantity, one point per simulated species. Grey lines mark zero on each axis and serve as the classification boundaries - 'Stouter' species should sit in the bottom left quadrant, 'Longer' species in the top right quadrant, and 'Null-effect smaller' species at 0,0 (intersection of the grey lines). Species in agreement with the reference are faded; species not in agreement are drawn at full opacity. The Pearson correlation between $\\hat\\beta_T$ and $\\beta_T$ ($r$) and the percentage of species in agreement are annotated in each panel. A species counts as 'in agreement' if its 95% confidence interval falls on the correct side of zero (Longer, Stouter), or if that interval includes zero (Null-effect smaller)."
#| fig-width: 9
#| fig-height: 6.5
#| fig-pos: 'H'

Eval_scatter_df <- Parms_tbl4 %>%
  filter(Temp_eff != "Null-effect larger") %>%
  mutate(Model = str_replace_all(Model, x_labs))

# Includes Longer, Stouter, and Null-effect smaller -- "correct" uses Est_correct (01_Run_simulation.R,
# Evaluation section) throughout: whole-CI-on-the-correct-side for Longer/Stouter, CI-includes-zero
# (reversed) for Null-effect smaller, since its true effect is exactly 0 by construction.
Eval_scatter_stats <- Eval_scatter_df %>%
  group_by(Model) %>%
  summarize(r = round(cor(beta_iso_true, b_temp_inc), 2),
            pct_correct = round(mean(Est_correct) * 100, 1),
            .groups = "drop")

ggplot(Eval_scatter_df, aes(x = beta_iso_true, y = b_temp_inc, color = Temp_eff, shape = Scaling)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  # Species in agreement with the reference recede (low alpha); species not in
  # agreement stay at full opacity so the structure of the disagreements -- which
  # categories, which quadrant -- is what the eye lands on. Small jitter spreads
  # the Proportionally-smaller pile that otherwise stacks exactly on x = 0.
  geom_point(data = \(d) dplyr::filter(d, Est_correct),  alpha = 0.28, size = 1.7,
             position = position_jitter(width = 0.006, height = 0.006, seed = 1)) +
  geom_point(data = \(d) dplyr::filter(d, !Est_correct), alpha = 0.9,  size = 2.1,
             position = position_jitter(width = 0.006, height = 0.006, seed = 1)) +
  geom_text(data = Eval_scatter_stats, aes(label = paste0("r=", r, ", ", pct_correct, "% in agreement")),
            x = -Inf, y = Inf, hjust = -0.05, vjust = 1.3, inherit.aes = FALSE, size = 2.6) +
  facet_wrap(vars(Model), scales = "free_y") +
  labs(x = expression(beta[T] ~ "(isometric-relative reference, per species)"),
       y = expression(hat(beta)[T] ~ "(estimated)"),
       color = "Category", shape = "Scaling") +
  guides(color = guide_legend(nrow = 1), shape = guide_legend(nrow = 1)) +
  theme(legend.position = "top", legend.box = "vertical")
```

## Empirical case studies

```{r}
#| label: tbl-species-accounting
#| tbl-cap: "Number of empirical species classified into each size / shape-shifting category (@tbl-classification), by dataset. Species exhibiting Bergmann's and Inverse Bergmann's rules are shown in the main text (@fig-empirical)."

dir_order <- c("Bergmann's", "Inverse Bergmann's", "Mixed - Wingier", "Mixed - Stouter", "Stable")

Species_accounting_tbl <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(Study, species, Direction) %>%
  count(Study, Direction) %>%
  pivot_wider(names_from = Direction, values_from = n, values_fill = 0) %>%
  mutate(Study = case_match(Study,
                             "Nightjar" ~ "Skinner et al (2025)",
                             "Weeks (2020)" ~ "Weeks et al (2020)",
                             "Atlantic birds" ~ "Rodrigues et al (2019)"),
         Study = factor(Study, levels = c("Skinner et al (2025)", "Weeks et al (2020)", "Rodrigues et al (2019)"))) %>%
  arrange(Study) %>%
  mutate(Study = as.character(Study)) %>%
  dplyr::select(Study, any_of(dir_order)) %>%
  mutate(Total = rowSums(across(any_of(dir_order))))

n_species_total <- sum(Species_accounting_tbl$Total)

Species_accounting_tbl <- bind_rows(
  Species_accounting_tbl,
  Species_accounting_tbl %>%
    summarise(Study = "Total", across(where(is.numeric), sum))
)

Dir_totals_row        <- Species_accounting_tbl %>% filter(Study == "Total")
pct_bergmanns_all     <- round(100 * Dir_totals_row$`Bergmann's` / n_species_total)
pct_inv_bergmanns_all <- round(100 * Dir_totals_row$`Inverse Bergmann's` / n_species_total)
pct_mixed_wingier_all <- round(100 * Dir_totals_row$`Mixed - Wingier` / n_species_total)
n_mixed_fatter_all    <- Dir_totals_row$`Mixed - Stouter`
pct_stable_all        <- round(100 * Dir_totals_row$Stable / n_species_total)

knitr::kable(Species_accounting_tbl,
             col.names = c("Study", "Bergmann's", "Inverse Bergmann's",
                           "Mixed - Wingier", "Mixed - Stouter", "Stable", "Total"),
             align = "c")
```

Across the `r n_species_total` species examined here, the majority of species (`r pct_bergmanns_all`%) adhered to Bergmann's rule in at least one body size metric, whereas `r pct_inv_bergmanns_all`% exhibited a pattern consistent with Inverse Bergmann's rule. `r pct_mixed_wingier_all`% of species increased in wing length while simultaneously getting lighter, while only `r n_mixed_fatter_all` species got heavier and simultaneously decreased wing length. `r pct_stable_all`% of species showed no significant response in either mass or wing (@tbl-species-accounting).

```{r}
#| label: rank-consistency-empirical
#| include: false

Rank_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(species, Study, Direction, rank_consistent) %>%
  filter(Direction %in% c("Bergmann's", "Inverse Bergmann's"))

Rank_summary_dir <- Rank_all %>%
  group_by(Direction) %>%
  summarize(n = n(), n_consistent = sum(rank_consistent),
            pct = round(100 * mean(rank_consistent)), .groups = "drop")

n_rank_total     <- nrow(Rank_all)
pct_rank_overall <- round(100 * mean(Rank_all$rank_consistent))

pull_rank_stat <- function(dir, col) {
  Rank_summary_dir %>% filter(Direction == dir) %>% pull({{ col }})
}

pct_rank_bergmanns     <- pull_rank_stat("Bergmann's", pct)
n_rank_bergmanns       <- pull_rank_stat("Bergmann's", n)
pct_rank_inv_bergmanns <- pull_rank_stat("Inverse Bergmann's", pct)
n_rank_inv_bergmanns   <- pull_rank_stat("Inverse Bergmann's", n)
```

The rank-order pattern from simulations was generally reproduced in all three datasets: ratio methods consistently produced the most positive estimates and multiple regression / OLS residuals the most negative, with SLI falling between these extremes (@fig-empirical). `r pct_rank_overall`% of the `r n_rank_total` species with a directional (Bergmann's or Inverse Bergmann's) classification showed the rank order predicted by the simulations (`r pct_rank_bergmanns`% of `r n_rank_bergmanns` Bergmann's species and `r pct_rank_inv_bergmanns`% of `r n_rank_inv_bergmanns` Inverse Bergmann's species; SI Figure 10).

```{=html}
<!-- Figures/Empirical_combined.png (Scripts/04b_Empirical_combined.R) is the reproducible
baseline Sylvia's polished PDF (below) is hand-edited from -- re-run that script and hand
over the fresh PNG whenever the underlying data/code changes, so her edit stays in sync
with the source. Caption counts (n_rank_bergmanns / n_rank_inv_bergmanns) are inlined live,
computed above. -->
```

![Estimated effect ($\hat{\beta}$) of temperature or year on relative wing length for each of the six methods, across three empirical avian datasets (Nightjars: red points; North American migrants: blue points; Atlantic Forest birds: green points), grouped by direction of body size change: (a) Bergmann's ($n$ = `r n_rank_bergmanns` species) and (b) Inverse Bergmann's ($n$ = `r n_rank_inv_bergmanns` species). Each point is one species' estimate ($\hat{\beta}$) for a given method; boxplots summarize the distribution of these estimates across species within each method. The percentage in the top-right corner of each boxplot is the proportion of species whose rank order across methods matched the simulation-predicted order (Bergmann's: Ratio \> SLI \> OLS; Inverse Bergmann's: Ratio \< SLI \< OLS). The stacked barplots on the right of each panel show binary adherence to Allen's rule based on species' 95% confidence intervals. Species were classified as significantly longer if their 95% confidence interval was above zero, significantly stouter if entirely below zero, or not significant if it overlapped zero.](../../Figures/Sylvia/Figure5_Empirical_combined_Sept16_edits.pdf){#fig-empirical width="100%" fig-pos="H"}

```{r}
#| label: significance-classification-empirical
#| include: false

Sig_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  filter(
    Direction %in% c("Bergmann's", "Inverse Bergmann's"),
    Approach  %in% c("Ratio", "Ratio2", "Ryding", "Resid_ols", "Sli_est", "Sli_iso"),
    std.error  < 0.5
  ) %>%
  mutate(Call = case_when(
    LCI95 > 0 ~ "Sig. longer",
    UCI95 < 0 ~ "Sig. stouter",
    TRUE      ~ "Not significant"
  ))

Sig_summary <- Sig_all %>%
  count(Direction, Approach, Call) %>%
  group_by(Direction, Approach) %>%
  mutate(pct = round(100 * n / sum(n))) %>%
  ungroup()

pull_sig_pct <- function(dir, approach, call) {
  Sig_summary %>%
    filter(Direction == dir, Approach == approach, Call == call) %>%
    pull(pct)
}

pct_bergmanns_ratio_longer     <- pull_sig_pct("Bergmann's", "Ratio", "Sig. longer")
pct_bergmanns_ryding_fatter    <- pull_sig_pct("Bergmann's", "Ryding", "Sig. stouter")
pct_bergmanns_olsresid_fatter  <- pull_sig_pct("Bergmann's", "Resid_ols", "Sig. stouter")
pct_bergmanns_sliiso_longer    <- pull_sig_pct("Bergmann's", "Sli_iso", "Sig. longer")
pct_bergmanns_sliiso_fatter    <- pull_sig_pct("Bergmann's", "Sli_iso", "Sig. stouter")

pct_invbergmanns_ratio_longer    <- pull_sig_pct("Inverse Bergmann's", "Ratio", "Sig. longer")
pct_invbergmanns_ryding_longer   <- pull_sig_pct("Inverse Bergmann's", "Ryding", "Sig. longer")
pct_invbergmanns_olsresid_longer <- pull_sig_pct("Inverse Bergmann's", "Resid_ols", "Sig. longer")

# Pooled across the two OLS-based methods (multiple regression + OLS residuals) rather than a mean of the two
# already-rounded per-method percentages above, so this is the % of all OLS-based-method x species rows in the
# given Call, not an average of two averages.
pct_bergmanns_ols_fatter <- Sig_all %>%
  filter(Direction == "Bergmann's", Approach %in% c("Ryding", "Resid_ols")) %>%
  summarize(pct = round(100 * mean(Call == "Sig. stouter"))) %>%
  pull(pct)

pct_bergmanns_ols_longer <- Sig_all %>%
  filter(Direction == "Bergmann's", Approach %in% c("Ryding", "Resid_ols")) %>%
  summarize(pct = round(100 * mean(Call == "Sig. longer"))) %>%
  pull(pct)

pct_invbergmanns_ols_longer <- Sig_all %>%
  filter(Direction == "Inverse Bergmann's", Approach %in% c("Ryding", "Resid_ols")) %>%
  summarize(pct = round(100 * mean(Call == "Sig. longer"))) %>%
  pull(pct)

# Per-species concordance in the significance call across methods, by direction.
# The six-method figure is restricted to species where SLI-estimated is defined
# (the per-group allometric-correlation filter); the five-method figure drops
# SLI-estimated and so covers every species in each directional set.
Sig_concord <- Sig_all %>%
  group_by(Direction, Study, species) %>%
  summarize(
    has_all6 = all(c("Ratio", "Ratio2", "Ryding", "Resid_ols", "Sli_est", "Sli_iso") %in% Approach),
    agree6   = n_distinct(Call) == 1,
    agree5   = n_distinct(Call[Approach != "Sli_est"]) == 1,
    .groups  = "drop"
  )

concord_pct <- function(dir, col, six = FALSE) {
  d <- Sig_concord %>% filter(Direction == dir)
  if (six) d <- d %>% filter(has_all6)
  round(100 * mean(d[[col]]))
}

pct_bergmanns_agree6    <- concord_pct("Bergmann's", "agree6", six = TRUE)
pct_bergmanns_agree5    <- concord_pct("Bergmann's", "agree5")
pct_invbergmanns_agree6 <- concord_pct("Inverse Bergmann's", "agree6", six = TRUE)
pct_invbergmanns_agree5 <- concord_pct("Inverse Bergmann's", "agree5")
```

Beyond differences in rank order, the six methods often disagreed on whether a species adhered to Allen's rule at all, and in which direction (@fig-empirical, bar plots). Among Bergmann's species, `r pct_bergmanns_agree5`% of species showed concordance across all methods examined (SLI-estimated excluded). This difference was most pronounced when comparing the wing/mass ratio and the OLS-based approaches, which classified `r pct_bergmanns_ratio_longer`% and `r pct_bergmanns_ols_fatter`% as significantly longer and significantly stouter, respectively. SLI-isometry fell between these extremes, with confidence intervals split almost evenly between significantly longer (`r pct_bergmanns_sliiso_longer`%) and significantly stouter (`r pct_bergmanns_sliiso_fatter`%). Among Inverse Bergmann's species, `r pct_invbergmanns_agree6`% of species showed concordance across all six methods (`r pct_invbergmanns_agree5`% when comparing the five methods in the larger dataset). Here the methods generally agreed that species were getting relatively longer, but detection rates varied substantially -- `r pct_invbergmanns_ratio_longer`% of species were classified as significantly longer using the wing/mass ratio, compared to roughly `r pct_invbergmanns_ols_longer`%, on average, for multiple regression and OLS residuals. These results indicate that the choice of body-size-standardization method can determine not only the magnitude and rank order of an estimated effect, but whether a researcher would conclude that Allen's rule holds for a given species at all.

### Correlation between the wing/mass ratio and body mass

```{r}
#| label: ratio-mass-correlation-empirical
#| include: false

Ratio_mass_cor_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_mass_cor.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_mass_cor.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_mass_cor.csv",  show_col_types = FALSE)
) %>%
  dplyr::select(Study, species, Metric, n, r, p_value)

# One row per Study x Metric (Ratio, Ratio2, Sli_iso, Sli_est), median r only --
# pivoted to one column per metric for display (tbl-ratio-mass-summary).
Ratio_mass_summary_tbl <- Ratio_mass_cor_all %>%
  summarize(N_species = n(), Median_r = median(r), .by = c(Study, Metric)) %>%
  mutate(Median_r = round(Median_r, 2),
         Metric = factor(Metric, levels = c("Ratio", "Ratio2", "Sli_iso", "Sli_est"))) %>%
  arrange(Study, Metric)

Ratio_mass_species_tbl <- Ratio_mass_cor_all %>%
  arrange(Study, species, Metric) %>%
  mutate(r = round(r, 2),
         p_value = ifelse(p_value < 0.001, "< 0.001", as.character(round(p_value, 3))))

# Simple ratio (A/S) -- the metric the surrounding prose already refers to
Ratio_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio")
n_rmc_total        <- nrow(Ratio_cor_only)
mean_rmc_all       <- round(mean(Ratio_cor_only$r), 2)
range_rmc_all      <- paste0(round(min(Ratio_cor_only$r), 2), " to ",
                             round(max(Ratio_cor_only$r), 2))
pct_rmc_strong_neg <- round(100 * mean(Ratio_cor_only$r < -0.7))

# Squared ratio (A^2/S) -- quantifies how much squaring the appendage term
# weakens (but does not eliminate) the correlation with mass.
Ratio2_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio2")
mean_rmc2_all       <- round(mean(Ratio2_cor_only$r), 2)
range_rmc2_all      <- paste0(round(min(Ratio2_cor_only$r), 2), " to ",
                              round(max(Ratio2_cor_only$r), 2))
pct_rmc2_strong_neg <- round(100 * mean(Ratio2_cor_only$r < -0.7))

# SLI-isometry / SLI-estimated species counts, for the table caption's sample-size
# caveat -- SLI-estimated is dropped (not NA-padded) for species that
# failed the per-group allometric-correlation filter, so it always has n <= SLI-isometry's.
n_sli_iso_total <- Ratio_mass_cor_all %>% filter(Metric == "Sli_iso") %>% nrow()
n_sli_est_total <- Ratio_mass_cor_all %>% filter(Metric == "Sli_est") %>% nrow()

# SLI-isometry / SLI-estimated mass correlations -- the SLI counterpart to the
# Ratio/Ratio2 stats above, for the Results paragraph's brief SLI comparison.
Sli_iso_cor_only <- Ratio_mass_cor_all %>% filter(Metric == "Sli_iso")
mean_rmc_sli_iso  <- round(mean(Sli_iso_cor_only$r), 2)
range_rmc_sli_iso <- paste0(round(min(Sli_iso_cor_only$r), 2), " to ",
                            round(max(Sli_iso_cor_only$r), 2))

Sli_est_cor_only <- Ratio_mass_cor_all %>% filter(Metric == "Sli_est")
mean_rmc_sli_est  <- round(mean(Sli_est_cor_only$r), 2)
range_rmc_sli_est <- paste0(round(min(Sli_est_cor_only$r), 2), " to ",
                            round(max(Sli_est_cor_only$r), 2))

tbl_cap_ratio_mass_summary <- glue::glue("Median per-species Pearson correlation with body mass, by empirical dataset, for the wing/mass ratio $\\log(A/S)$, its squared counterpart $\\log(A^2/S)$, SLI-isometry, and SLI-estimated. SLI-estimated is reported for fewer species ({n_sli_est_total} of {n_sli_iso_total} across all three datasets) than the other three metrics, because it is undefined for species that failed the per-group allometric-correlation filter.")
```

Across `r n_rmc_total` species, the ratio metric was consistently and strongly negatively correlated with mass (mean $r$ = `r mean_rmc_all`, range `r range_rmc_all`). `r pct_rmc_strong_neg`% of species had a correlation more negative than $-0.7$. This pattern held regardless of taxon, gradient type (spatial or temporal), or geographic region, confirming that the ratio's dependence on body size is a general mechanical property of the metric rather than a quirk of any one dataset (@tbl-ratio-mass-summary). The correlation between $\log(A^2/S)$ and body mass was still strongly negative, but consistently weaker in magnitude than its unsquared counterpart (mean $r$ = `r mean_rmc2_all`, range `r range_rmc2_all`, `r pct_rmc2_strong_neg`% of species below $-0.7$, @tbl-ratio-mass-summary). Both SLI parameterizations were less correlated with mass than either ratio metric (SLI-isometry: mean $r$ = `r mean_rmc_sli_iso`, range `r range_rmc_sli_iso`; SLI-estimated: mean $r$ = `r mean_rmc_sli_est`, range `r range_rmc_sli_est`) (@tbl-ratio-mass-summary; SI Figure 11).

```{r}
#| label: tbl-ratio-mass-summary
#| tbl-cap: !expr tbl_cap_ratio_mass_summary

Ratio_mass_summary_tbl %>%
  dplyr::select(Study, Metric, Median_r) %>%
  pivot_wider(names_from = Metric, values_from = Median_r) %>%
  knitr::kable(
    col.names = c("Study", "Wing / Mass", "Wing² / Mass", "SLI-isometry", "SLI-estimated")
  )
```

# Discussion

These results demonstrate that body-size adjustment is not a neutral preprocessing step in studies of Allen's rule. The six approaches examined here produced systematically different estimates of relative appendage morphology in simulations, and this rank ordering was reproduced across three empirical datasets spanning `r n_rank_total` species and substantially different taxa, ecosystems, and geographic contexts. Importantly, biological conclusions regarding adherence to Allen's rule varied tremendously depending on which statistical method was employed. The statistical approaches quantify morphological change in different ways - the key for researchers is to match the estimator to the biological question of interest.

## Ratios and implicit scaling relationships

Ratios are attractive because they are simple, familiar, and directly interpretable. However, their algebraic form implicitly specifies a scaling relationship between appendage length and body size. The scaling assumptions underlying these ratios become clearer when both ratios and SLI metrics are expressed in the general form $A/S^\beta$, where $\beta$ represents the scaling slope. Under log-log transformation, the simple ratio $A/S$ is equivalent to $A/S^1$ (i.e., $\beta = 1$), while the squared-ratio metric $A^2/S$ is linearized as $A/S^{0.5}$. The dimensional appeal of dividing an area-like quantity ($A^2$) by mass is understandable, but allometric scaling theory and empirical data [@youngfleshAbioticConditionsShape2022] suggest that $1/3$ is a more appropriate scaling reference for a linear metric than $1/2$. When using a volumetric proxy for body size (such as body mass, $M$), SLI-isometry uses the geometrically expected isometric exponent of $1/3$ and is therefore algebraically equivalent to the ratio $A/M^{1/3}$ (up to a constant multiplier determined by the reference mass $M_0$). This dimensional bridge exposes a fundamental scaling mismatch in some past studies of climate-driven morphological change. Indeed, studies utilizing a simple ratio of appendage to body mass ($A/M$) to represent relative appendage size [e.g. @jirinecMorphologicalConsequencesClimate2021; @ballingerContributionGeneticEnvironmental2022] implicitly force a scaling exponent of $\beta = 1$ despite that the geometrically expected scaling relationship is $1/3$.

This distinction is particularly important when body size itself varies along an environmental gradient. When overall body mass declines in response to warming, consistent with Bergmannian patterns, the scaling mismatch imposed by $A/M$ can create a strong negative correlation between the ratio and body mass [e.g., @frohlichAllometryRevealsTradeoffs2023]. Consequently, $A/M$ can increase even when appendage length scales strictly isometrically with body mass [@rydingMultipleRegressionNot2025]. A temperature-associated increase in $A/M$ could therefore arise from declining body size alone, rather than from appendages becoming disproportionately longer [a mathematical artifact also documented in condition-index studies; @greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996]. Such a pattern could be misclassified as adaptive shape-shifting [@glazierComplicationsBodysizeCorrection2022].

This also highlights that SLI-isometry is not fundamentally distinct from ratio approaches; rather, it represents a mathematically explicit, geometrically-justified ratio. I argue that the primary advantages of the SLI framework come from transparency, interpretability, and flexibility. Rather than embedding scaling assumptions implicitly in a ratio, the SLI framework makes the scaling assumption, the reference body size ($M_0$), and the resulting units explicit. The same framework can then be used with an empirically estimated scaling exponent, as in SLI-estimated, rather than requiring a separate ratio formulation for each scaling assumption. These favorable properties make the SLI framework preferable to conventional ratio approaches.

```{r}
#| label: fig-ols-sma
#| fig-cap: "The allometric relationship between log-mass and log-appendage for a simulated hyperallometric species in which temperature increases produce relatively longer appendages. The SMA regression line (blue) is steeper than the OLS line (red). Individual points are coloured by temperature. Sections labelled I and II highlight regions where OLS and SMA assign opposite signs to residuals: small-bodied individuals (Section I) are classified as having relatively short appendages by OLS but relatively long appendages by SMA. Large-bodied individuals (Section II) are classified in the opposite direction. Where temperature increases select for smaller body size (Bergmann's rule), individuals are displaced toward Section I, exactly where OLS assigns more negative residuals than SMA, producing a more negative $\\hat{\\beta}_T$."
#| fig-pos: 'H'
#| fig-height: 4.5

# Reuses Ex_df_hypo_hyper, the same example species already generated for @fig-sli-concept
# panel (b) by Scripts/04a_Concept_figures.R, which exports it here for reuse.
Ex_df_hypo_hyper <- readRDS("Derived/Rds/Ex_df_hypo_hyper.rds")
xy_lims <- Ex_df_hypo_hyper %>%
  summarize(max.append = max(Append_log), min.append = min(Append_log),
            max.mass   = max(Mass_log),   min.mass   = min(Mass_log))

ggplot(Ex_df_hypo_hyper, aes(x = Mass_log, y = Append_log)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ggpmisc::stat_ma_line(method = sma_or_ma, color = "blue") +
  geom_point(alpha = 0.4, aes(color = Temp_inc)) +
  annotate("text", x = 4.23, y = 5.13, label = "I",  size = 6, color = "black") +
  annotate("text", x = 4.52, y = 5.25, label = "II", size = 6, color = "black") +
  scale_color_viridis_c(option = "plasma") +
  lims(y = c(xy_lims$min.append, xy_lims$max.append),
       x = c(xy_lims$min.mass,   xy_lims$max.mass)) +
  labs(x = "Log(mass)", y = "Log(appendage)", color = "Temperature\nincrease")
```

## Multiple regression for conditional associations

Multiple regression takes a different approach to body-size adjustment, estimating the relationship between appendage size and temperature conditional on body size. The properties and assumptions of multiple regression are well understood and are well-suited to questions regarding the conditional association between two variables [@zuurProtocolConductingPresenting2016; @freckletonMisuseResidualsEcology2002]. For these reasons, several authors have argued that multiple regression is the most appropriate way to adjust for body size in questions of Allen's rule [@nakagawaDivideConquerSize2017; @rydingResponseAllometryEvaluate2022; @rydingMultipleRegressionNot2025], including over the OLS residuals approach [@freckletonMisuseResidualsEcology2002]. Ultimately, multiple regression is appropriate if the goal of Allen's rule is to describe a pattern of covariation between temperature and appendage length [@symondsItsAllRelative2025; also see @wattBergmannsRuleConcept2010], conditional on body size. However, it is important to note that multiple regression does not isolate departures from isometry or any other geometric null hypothesis.

Relative to the other approaches examined, multiple regression consistently returned more negative estimates of $\hat\beta_T$, which can largely be understood by closer examination of the scaling slope. Relative to SMA regression, OLS produces a shallower slope whenever $|r| < 1$, because it minimizes deviations vertically rather than in both dimensions [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009]. Because the OLS slope is shallower, OLS and SMA assign different relative appendage values to individuals at different body sizes. When temperature is associated with declining body size, as under Bergmann's rule, the temperature gradient therefore shifts observations systematically toward region I, where OLS assigns more negative residuals than SMA (@fig-ols-sma).

This systematic trend observed in OLS methods has important implications for interpreting morphological responses to climate change. OLS-based approaches were used in `r pct_ols_relative`% of studies examining relative appendage length, which raises the possibility that OLS-based estimates of appendage elongation may be systematically conservative in settings where temperature is associated with declining body size. Indeed, in the three empirical datasets, `r pct_bergmanns_ratio_longer`% of Bergmann's-rule species were classified as significantly longer (i.e., adhering to Allen's rule) using the wing/mass ratio, compared to only `r pct_bergmanns_ols_longer`% using the OLS-based approaches. Importantly, @rydingMultipleRegressionNot2025 found similar methodological divergences between ratio-based and multiple-regression approaches using a different simulation approach and analytical framework, lending strength to our findings. These broad swings in inference highlight the importance of distinguishing these statistical quantities when evaluating claims that climate-driven morphological change represents adaptive shape-shifting [e.g., @rydingShapeshiftingChangingAnimal2021].

## What constitutes an Allen's-rule response?

The estimators described here clearly have distinct statistical behavior, which raises a more fundamental question for studies of Allen's rule - what is the 'correct' estimator to detect a change in relative appendage morphology? In Allen's original paper he notes, “certain parts of the organism vary more than does general size...”, perhaps alluding to isometry as a reasonable theoretical expectation when distinguishing genuine shape-shifting from size change alone [@allen1877influence]. Recent simulation studies have also implied that isometric changes in appendage length are a reasonable null hypothesis [@symondsItsAllRelative2025; @rydingMultipleRegressionNot2025], although importantly, the consequences for thermoregulation are not known precisely [@santoroAllometryEvaluateAllens2022; but see @tabhLimitedEvidenceThat2025].

Other scaling references may be more appropriate when non-thermoregulatory forces are known to produce systematic departures from isometry. For example, bird wingspan scales with positive allometry due to the biomechanics of flight [@raynerFormFunctionAvian1988], and certain ornaments or weaponry used in mate acquisition can experience strong positive allometry due to sexual selection [@shingletonAllometryStudyBiological2010; @kodric-brownAllometryOrnamentsWeapons2006; @cuervoAllometricPatternSexually2009]. The goal then may be to remove systematic differences in appendage length due to allometric scaling in order to isolate any additional elongation due to temperature (or some other non-scaling factor). In these cases, SLI-estimated provides an excellent tool as, by construction, it removes the signal in relative appendage length that is due to positive allometry.

## Limitations and next steps for SLI approaches

```{r}
#| label: sli-limitations-stats
#| include: false

Allometric_cor_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_allometric_cor.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_allometric_cor.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_allometric_cor.csv",  show_col_types = FALSE)
)

pct_sli_est_total <- round(100 * n_sli_est_total / n_sli_iso_total)

# Fraction of species per study whose mass-wing correlation falls below the
# cor_min = 0.3 threshold used to gate SLI-estimated (Methods).
pct_atlantic_weak_cor <- Allometric_cor_all %>%
  filter(Study == "Atlantic birds") %>%
  summarise(pct = round(100 * mean(cor_mw < 0.3))) %>%
  pull(pct)

pct_weeks_weak_cor <- Allometric_cor_all %>%
  filter(Study == "Weeks (2020)") %>%
  summarise(pct = round(100 * mean(cor_mw < 0.3))) %>%
  pull(pct)
```

Despite their advantages, the SLI approaches are not without limitations. Unfortunately, SMA regression is inflexible in its ability to customize models because it can only model the two scaling variables $X$ and $Y$. This inflexibility can make it more challenging to meet SMA regression assumptions relative to OLS assumptions [@wartonBivariateLinefittingMethods2006]. For example, SMA regression assumes homoskedasticity, but missing covariates that influence appendage measurements (e.g., feather wear) cannot be included in SMA regression, which could lead to heteroskedasticity. Similarly, if a non-linear relationship remains after log-log transforming the linear and volumetric measurement, the SMA assumption of linearity is violated. This non-linearity can be easily accommodated in classical regression [e.g., $\mathit{Body\ size} + \mathit{Body\ size^2}$, @baldwinComplementarityAllensBergmanns2023], but not in SMA regression. In these cases, regression approaches that accommodate additional covariates may be preferable [but also see @glazierComplicationsBodysizeCorrection2022].

One limitation particular to SLI-estimated is that the SMA slope is unreliable at small sample sizes or when there is little allometric signal [i.e., mass–appendage correlation is weak, @smithUseMisuseReduced2009]. In the empirical datasets examined here, a per-species SMA slope could be estimated for only `r pct_sli_est_total`% of species. Interestingly, there was a general morphological decoupling of mass and wing in Atlantic Forest birds (`r pct_atlantic_weak_cor`% below r \< 0.3) that was less prominent in North American migrants [`r pct_weeks_weak_cor`%, @skinnerDataEnvironmentalGeographic2025; @weeksSharedMorphologicalConsequences2020]. While this dependence on a minimum correlation to use SLI-estimated can be seen as a limitation, researchers should consider whether body size needs to be controlled for in the first place if the correlation between size and appendage is minimal.

One important next step is to examine the relationship between SLI-derived measures and biologically meaningful aspects of morphology or function. For example, @peigNewPerspectivesEstimating2009 found that the Scaled Mass Index predicted variation in fat and protein reserves better than several conventional condition indices across multiple vertebrate taxa. Analogous validation against independent morphological measures, such as geometric morphometric or three-dimensional measurements of appendage surface area and body volume [e.g., @rydingMeasuringAvianBill2025; @adamczakUsing3DModels2019], could similarly test whether particular SLI parameterizations capture biologically meaningful variation in relative appendage size. Such validation should span taxa near the center and the periphery of morphospace [@pigotMacroevolutionaryConvergenceConnects2020]. For example, SLI indices may capture relative appendage length well in taxa with ‘standard’ body plans (e.g., sparrows), but struggle to capture more unusual body plans (i.e., hummingbirds, penguins).

## Future work: From association to causal inference

For those who seek to move beyond documenting patterns, causal inference provides a framework for asking how environmental changes would alter biological responses under specified interventions [@franksEcologyNeedsCausal2025]. For example, predicting the effects of climate change on species or populations, separately from other environmental changes such as land-use change, requires asking how morphology would change under an intervention on temperature, precipitation, or any other environmental gradient, relative to an appropriate counterfactual in which that intervention did not occur [@dudneyCausalInferenceFramework2024]. Thus, conservation in the face of global change will increasingly require approaches that distinguish causal effects from statistical associations to distinguish the consequences of distinct, but often correlated, environmental drivers [@franksEcologyNeedsCausal2025].

Another benefit of a causal inference framework is the ability to decompose a total effect into direct and indirect effects. In the context of Allen's rule, the effect of temperature on appendage length could be decomposed into an indirect effect mediated through body size (temperature $\rightarrow$ body size $\rightarrow$ appendage length) and a direct effect representing the temperature $\rightarrow$ appendage pathway. Path analysis or structural equation modeling within an explicit causal framework are particularly well suited to estimating such decompositions [e.g., @xuEcologicalPredictorsInterspecific2023], which could help clarify the combined effects of Bergmann's and Allen's rules [@skinnerBergmannAllenUnified2026]. Importantly, the SLI decomposition should not itself be interpreted as a causal decomposition. The components describing allometric and temperature-associated variation are statistical quantities defined by the chosen standardization framework, but are not causal pathways that cleanly separate into direct and indirect effects of temperature on appendage length. Where causal interpretation is the objective, researchers should articulate the relevant causal assumptions and consider whether an explicit causal framework is warranted [@siegelFoundationsFutureDirections2025].

# Conclusions

Relative appendage length is not a uniquely defined quantity. Different approaches encode different assumptions about how body size and appendage morphology are related, and these assumptions lead to systematically different estimates of morphological change. The SLI metrics presented here provide a flexible and interpretable framework for scalar relative appendage size, with two parameterizations that correspond to distinct biological questions. I also hope that the SLI metrics will encourage biologists to think critically about allometric scaling and, more importantly, about whether the statistical estimator matches the biological question at hand. I recommend that adjustment for body size should be treated as part of the inferential design of a study rather than as a routine preprocessing step.

{{< pagebreak >}}

# References
