---
title: "Supplementary Information: The Shape of Things to Come — Simulations Show Promise in a New Metric, the Standardized Length Index"
number-sections: true
execute:
  echo: false
  message: false
  warning: false
format:
  pdf:
    toc: false
    include-in-header:
      - text: |
          \usepackage{authblk}
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
editor_options:
  chunk_output_type: console
metadata-files:
  - ../../Suppfiles/authors.yml
---

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(ggpubr)
library(cowplot)
library(ggpmisc)
ggplot2::theme_set(theme_cowplot())
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: load-simulation-results
#| include: false

# Simulation is run by Scripts/Run_simulation.R and saved to
# Derived/Rds/simulation_results.rds; the main manuscript reads the same file.
# This brings Parms_mat3, Parms_tbl4, Parms_temp_bs, sma_or_ma, x_labs, and
# every Proportionally-larger-scenario percentage into scope by name.
list2env(readRDS("Derived/Rds/simulation_results.rds"), envir = environment())

Parms_tbl4_bigger <- Parms_tbl4 %>% filter(Temp_eff == "Proportionally larger")
```

# Validation of simulated species direction

The main text simulates species by crossing correlations between temperature, mass, and appendage length; whether a given hypothetical species actually moved in its intended direction (e.g., "Fatter" species becoming relatively shorter-winged) was validated by examining the correlation between binned temperature and the SMA intercept, and species with the wrong-signed correlation were excluded from further analysis (main text, Data simulation). This validation check reflects each species' own realized (SMA-estimated) scaling relationship, not the fixed isometric anchor used to define the main text's ground truth (main text, Effect classification). For Proportionally smaller and Proportionally larger species, whose isometric-relative effect is exactly null by design, this means the SMA-intercept correlation is *not* expected to be centered on zero: guaranteeing an exactly null isometric-relative effect for these species necessarily requires $\rho_{13} \neq \rho_{23}$ (main text, Effect classification), and it is exactly that differential correlation the SMA-intercept check picks up — the same pathway responsible for SLI-estimated's divergence from SLI-isometry among these species (main text, Simulation performance; Proportionally larger scenario below).

```{r}
#| label: fig-cor-allometry-values
#| fig-cap: "Correlation between binned temperature and the SMA intercept (a proxy for each species' own realized scaling relationship, not the isometric ground truth used elsewhere in this paper) for every simulated species, used to validate that Fatter and Longer species moved in their intended direction. Fatter species should show a negative correlation and Longer species a positive correlation. Proportionally smaller and Proportionally larger species are not expected to straddle zero here, despite having an exactly null isometric-relative effect by design (see main text); species falling on the wrong side of zero for their category (e.g., points above the dashed line for Fatter species) were excluded from further analysis, an exclusion rule that was not applied to the Proportionally smaller/larger categories. Point shape distinguishes models fit with a fixed slope ('No interaction') from models allowing the slope to vary with temperature ('Interaction'); point size indicates allometric scaling category and colour indicates the strength of the difference between temperature's effect on appendage length and mass."
#| fig-pos: 'H'

Parms_temp_bs %>%
  ggplot(aes(x = Temp_eff, y = Correlation)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = .3), alpha = .3,
             aes(color = Strength, shape = SMA_mod, size = Scaling)) +
  labs(x = NULL, y = "Correlation between temperature\nand SMA intercept",
       shape = "Model", size = "Scaling", color = "Strength\nAllen's") +
  theme(axis.text.x = element_text(size = 10, vjust = .58, angle = 30))
```

# Allometric scaling and correlation strength

```{r}
#| label: fig-wingy-varied-scaling
#| fig-cap: "Allometric relationships between log-mass and log-appendage for nine hypothetical species simulated to become relatively longer with increasing temperature, illustrating how allometric scaling (rows) and the appendage-mass correlation (columns) jointly shape the geometry of the point cloud. Within each row the SMA slope (blue) is held constant while the correlation increases across columns, producing a tighter spread of points around the line of best fit; across rows the SMA slope increases from hypoallometric to hyperallometric scaling, rotating the cloud from a more horizontal to a more vertical orientation. Points are coloured by the amount of temperature increase; the OLS line of best fit (red) is shown for comparison."
#| fig-width: 9
#| fig-height: 9
#| fig-pos: 'H'

vary_scaling_parms <- Parms_mat3 %>%
  filter(Temp_eff == "Longer") %>%
  slice_max(n = 1, order_by = tibble(Strength),
            by = c(Scaling, r_12), with_ties = FALSE)

## gen_ex_data() draws new individuals on every render; seed it so the figure is reproducible.
set.seed(20260711)
vary_scaling_tbl <- gen_ex_data(vary_scaling_parms) %>%
  mutate(Temp_inc = Temp_inc / 3,
         Scaling  = factor(Scaling, levels = c("Hypoallometry", "Isometry", "Hyperallometry")),
         r_12_lab = paste("r =", r_12))

vary_scaling_tbl %>%
  ggplot(aes(x = Mass_log, y = Append_log)) +
  geom_point(alpha = 0.4, aes(color = Temp_inc)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ggpmisc::stat_ma_line(method = sma_or_ma, se = FALSE, color = "blue") +
  scale_color_viridis_c(option = "plasma") +
  facet_grid(Scaling ~ r_12_lab) +
  labs(x = "Log(mass)", y = "Log(appendage)", color = "Temperature\nincrease")
```

# Correlation between relative appendage length metrics and body mass

The main text attributes the ratio methods' upward bias under Bergmann's rule to a mechanical correlation between the ratio metric and body mass, a pattern confirmed empirically across three datasets (main text, ratio/mass correlation table). Here I extend that check to the simulated data and to all five methods that produce an individual-level shape metric; mass-as-covariate is excluded because it returns only a regression coefficient, not an index that can be correlated with mass.

```{r}
#| label: mass-cor-sim-stats
#| include: false

n_sim_species_cor <- nrow(Parms_tbl3)

Mass_cor_summary <- Mass_cor_tbl %>%
  summarize(Mean_r = mean(r_mass), Min_r = min(r_mass), Max_r = max(r_mass), .by = Model) %>%
  mutate(across(c(Mean_r, Min_r, Max_r), \(x) round(x, 2)))

pull_mass_cor <- function(model, stat) {
  Mass_cor_summary %>% filter(Model == model) %>% pull({{ stat }})
}

mean_r_ratio     <- pull_mass_cor("Ratio",     Mean_r)
range_r_ratio    <- paste(pull_mass_cor("Ratio", Min_r), "to", pull_mass_cor("Ratio", Max_r))
mean_r_ratio2    <- pull_mass_cor("Ratio2",    Mean_r)
range_r_ratio2   <- paste(pull_mass_cor("Ratio2", Min_r), "to", pull_mass_cor("Ratio2", Max_r))
mean_r_sli_iso   <- pull_mass_cor("Sli_iso",   Mean_r)
range_r_sli_iso  <- paste(pull_mass_cor("Sli_iso", Min_r), "to", pull_mass_cor("Sli_iso", Max_r))
mean_r_sli_est   <- pull_mass_cor("Sli_est",   Mean_r)
range_r_sli_est  <- paste(pull_mass_cor("Sli_est", Min_r), "to", pull_mass_cor("Sli_est", Max_r))
max_abs_r_ols    <- Mass_cor_tbl %>% filter(Model == "Ols_resid") %>% pull(r_mass) %>% abs() %>% max() %>% signif(2)

fig_cap_mass_cor <- glue::glue("Per-species Pearson correlation between body mass and each method's individual-level relative appendage length metric, across all {n_sim_species_cor} simulated species. Mass-as-covariate is excluded because it produces only a model coefficient, not an individual-level index. Point colour shows the underlying appendage-mass correlation ($\\rho_{{12}}$) used to simulate each species.")
```

```{r}
#| label: fig-mass-cor-sim
#| fig-cap: !expr fig_cap_mass_cor
#| fig-width: 8
#| fig-height: 5
#| fig-pos: 'H'

Mass_cor_tbl %>%
  mutate(r_12  = as_factor(r_12),
         Model = str_replace_all(Model, x_labs),
         Model = factor(Model, levels = c("Appendage / mass", "Appendage² / mass",
                                           "OLS residuals", "SLI isometry", "SLI estimated"))) %>%
  ggplot(aes(x = Model, y = r_mass)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, aes(color = r_12, shape = Scaling)) +
  labs(x = NULL, y = "Correlation with body mass", color = expression(rho[12])) +
  theme(axis.text.x = element_text(size = 10, vjust = .58, angle = 30))
```

The simulated data reproduce the empirical pattern and clarify its mechanism. **OLS residuals** showed correlations with mass of essentially zero in every simulated species (maximum $|r|$ = `r max_abs_r_ols`) — not because residual regression is unbiased, but because least-squares residuals are mathematically orthogonal to the predictor by construction; a correlation of zero here reflects a property of the algebra, not evidence that the method isolates a true shape signal. **Ratio** ($A/S$) showed the strongest correlation with mass (mean $r$ = `r mean_r_ratio`, range `r range_r_ratio`), confirming the mechanical dependence responsible for its bias under Bergmann's rule (main text). **Ratio2** ($A^2/S$) was weaker but still consistently negative (mean $r$ = `r mean_r_ratio2`, range `r range_r_ratio2`), consistent with the partial correction from squaring the appendage term. **SLI-isometry** (mean $r$ = `r mean_r_sli_iso`, range `r range_r_sli_iso`) and **SLI-estimated** (mean $r$ = `r mean_r_sli_est`, range `r range_r_sli_est`) fell between the ratio and OLS-residual extremes: weaker than either ratio metric, but — unlike OLS residuals — not exactly zero. This is expected, since SLI standardizes appendage length using a fixed or estimated allometric exponent rather than an OLS-fitted line, so it is not mathematically guaranteed to be orthogonal to mass unless that exponent happens to equal the OLS slope. SLI's good directional-accuracy performance (main text, Simulation performance) therefore does not depend on being perfectly uncorrelated with mass, but on using an exponent that better approximates the true allometric relationship than either the ratio's implicit exponent or OLS's attenuated slope.

# Proportionally larger scenario: simulated analogue of Inverse Bergmann's responses

The main text simulation compares six methods for estimating relative appendage length across three realizations of Bergmann's rule (Fatter, Proportionally smaller, Longer), in which body mass never increases with warming. Here I additionally simulated a **Proportionally larger** scenario — the mirror image of Proportionally smaller, in which temperature increases wing length and body mass with equal, not differential, pull ($b \in \{0.22, 0.33, 0.44\}$, $\rho_{12} \in \{0.3, 0.45, 0.6\}$, $\rho_{13}$ solved from $\rho_{23} \in \{0.15, 0.3, 0.5, 0.7\}$ so the true isometric-relative effect is exactly null; main text, Effect classification; 36 simulated species). Because the true effect is exactly zero for every species in this design, an unbiased method should classify exactly half of Proportionally larger species as becoming relatively longer, matching the true value of `r Bigger_longer_pct`%. This scenario is the simulated analogue of empirically observed **Inverse Bergmann's** responses (main text classification table), in which mass, wing length, or both increase with warming.

```{r}
#| label: fig-bigger-compare
#| fig-cap: "Estimated effect of temperature on relative appendage length ($\\hat{\\beta}_T$) for each method among species simulated as Proportionally larger (body mass and wing length both increasing with warming; 36 simulated species). Positive values indicate that species were classified as becoming relatively longer; the true effect is exactly zero in every species, so the expected value is zero for an unbiased method. Point shape indicates allometric scaling category."
#| fig-width: 8
#| fig-height: 5
#| fig-pos: 'H'

Parms_tbl4_bigger %>%
  mutate(Model = str_replace_all(Model, x_labs)) %>%
  ggplot(aes(x = Model, y = b_temp_inc)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(shape = Scaling), width = 0.15, alpha = 0.6, size = 2.5) +
  labs(x = NULL, y = expression(hat(beta)[T] ~ "on relative appendage length")) +
  theme(axis.text.x = element_text(size = 10, vjust = .58, angle = 55))
```

The pattern mirrors, and inverts, the classical biases described in the main text. Because mass now increases rather than decreases with temperature, the ratio approach's denominator grows instead of shrinking: only `r Ratio_bigger_pos`% of Proportionally larger species (versus the true proportion of `r Bigger_longer_pct`%) were classified as becoming relatively longer — the opposite-direction analogue of the ratio's upward bias in shrinking (Bergmann's) species. Multiple regression and OLS residuals show the mirror-image error: `r Ryding_bigger_pos`% of Proportionally larger species were classified as becoming relatively longer despite the complete absence of a true shape signal, reflecting the same attenuation mechanism described in the main text, reinforcing rather than opposing the (null) signal because mass is increasing rather than decreasing. SLI-isometry showed a smaller, but still real, bias toward classifying Proportionally larger species as relatively fatter (`r Sli.iso_bigger_pos`%, versus the true `r Bigger_longer_pct`%), while SLI-estimated (`r Sli.est_bigger_pos`%) deviated further from the true proportion — the same pattern, and for the same structural reason, as in the Proportionally smaller scenario (main text, Simulation performance): forcing the isometric-relative true effect to exactly zero necessarily opens a real differential-correlation pathway that SLI-estimated, but not SLI-isometry, is structurally sensitive to. Accounting for uncertainty makes the gap between the two SLI variants decisive: a 95% confidence interval correctly included zero for `r Sli.iso_bigger_null_correct`% of Proportionally larger species under SLI-isometry, against `r Sli.est_bigger_null_correct`% under SLI-estimated and `r Ratio_bigger_null_correct`–`r Ryding_bigger_null_correct`% for the four non-SLI methods — matching the main text's Proportionally smaller result almost exactly. This is based on a comparatively modest sample of 36 simulated species.

# Robustness check: anchoring the ground truth to OLS instead of SMA

The main text anchors "isometric" (0.33) to the SMA scale (main text, Approach classification), matching SLI's own error-in-both-variables logic. As a check on how much that choice matters, I recomputed an alternative ground truth anchored to the OLS scale instead: $\beta_{OLS} = \rho_{12} \cdot \beta_{SMA}$ and $\beta_{T,OLS} = \rho_{13} \cdot \beta_{OLS} - 0.33 \cdot \rho_{23}$, then re-scored all six methods' sign-agreement against it among Longer and Fatter species (Proportionally smaller/larger are excluded here, as in the main text's calibration scatter (main text, Simulation performance), since their isometric-relative effect is exactly null by design and so has no signed value to agree or disagree with under either anchor).

```{r}
#| label: tbl-ols-anchor
#| tbl-cap: "Sign-agreement of each method's estimated coefficient against the adopted (SMA-anchored) ground truth versus an alternative OLS-anchored ground truth, among Longer and Fatter species."

knitr::kable(Ols_anchor_summary_tbl)
```

SLI-isometry's agreement drops sharply (`r Sliiso_pct_sma`% to `r Sliiso_pct_ols`%) under OLS-anchoring, as expected since the ground truth no longer matches its own formula. Ratio, by contrast, improves substantially (`r Ratio_pct_sma`% to `r Ratio_pct_ols`%) — closing much of the gap with the SLI variants, though not matching them or Ratio² under either anchor — while Mass-as-covariate and OLS-residuals both fall further behind (`r Ryding_pct_sma`% to `r Ryding_pct_ols`%, and `r Olsresid_pct_sma`% to `r Olsresid_pct_ols`%, respectively) — the opposite of what "they're OLS-based methods, so they should track an OLS-anchored truth" would predict, because both test deviation from each species' own estimated relationship rather than from a fixed external anchor, and so are structurally insensitive to the allometry-vs-anchor pathway regardless of which scale that anchor is stated on. This confirms the SMA-vs-OLS choice for 0.33 is consequential, not a relabeling, while leaving the case for anchoring to SMA (main text, Approach classification) unchanged.

```{r}
#| label: fig-calibration-ols
#| fig-cap: "The main text's per-species calibration scatter (main text, Simulation performance) recreated with the ground truth anchored to OLS instead of SMA (x-axis: the OLS-anchored true effect), among Longer and Fatter species. Same species and estimated coefficients as the main text figure; only the definition of the true value has moved."
#| fig-width: 9
#| fig-height: 6.5
#| fig-pos: 'H'

ggplot(Ols_anchor_df, aes(x = beta_iso_true_ols, y = b_temp_inc, color = Temp_eff, shape = Scaling)) +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_point(alpha = .55, size = 1.2) +
  geom_text(data = Ols_anchor_stats, aes(label = paste0("r=", r, ", ", pct_correct, "% correct")),
            x = -Inf, y = Inf, hjust = -0.05, vjust = 1.3, inherit.aes = FALSE, size = 2.6) +
  facet_wrap(vars(Model), scales = "free_y") +
  labs(x = expression(beta[T*","*OLS] ~ "(analytic, OLS-anchored)"),
       y = expression(hat(beta)[T] ~ "(estimated)"),
       color = "Category", shape = "Scaling") +
  guides(color = guide_legend(nrow = 1), shape = guide_legend(nrow = 1)) +
  theme(legend.position = "top", legend.box = "vertical")
```

# Why SLI-isometry's correlation with the ground truth isn't exactly 1

The main text notes that anchoring $\beta_T$ to the same isometric scale SLI-isometry itself uses makes it circular to "test" SLI-isometry against the other approaches (Approach classification). If that's right, SLI-isometry's fitted coefficient should be an almost-exact function of $\beta_T$, not just a strongly correlated one — the raw correlation in the main text's calibration scatter (main text, Simulation performance) is r $\approx$ 0.97, not 1, which at first glance looks like room for genuine daylight between the two. It isn't: the gap is (almost entirely) a missing scaling term, not noise.

$\beta_T$ is defined from $\mathrm{Cov}(\log SLI, \mathrm{Temperature})$ alone (main text, Effect classification), but a *fitted regression coefficient* is that covariance divided by $sd(\log SLI)$ too — a correlation, not a covariance. $sd(\log SLI)^2$ is proportional to

$$V = \beta_{SMA}^2 + 0.33^2 - 2 \cdot 0.33 \cdot \rho_{12} \cdot \beta_{SMA}$$

which varies species to species (it depends on each species' own $\beta_{SMA}$ and $\rho_{12}$), and $\beta_T$ never accounted for it — only the *sign* claim survives that omission, since $\sqrt{V}>0$ always. Dividing $\beta_T$ by $\sqrt{V}$ restores the missing term.

```{r}
#| label: tbl-sli-iso-rescale
#| tbl-cap: "Correlation between SLI-isometry's fitted coefficient and the ground truth, before and after the variance correction."

tibble::tibble(Version = c("Raw", "Variance-corrected"), r = c(r_sli_iso_raw, r_sli_iso_rescaled)) %>%
  knitr::kable()
```

```{r}
#| label: fig-sli-iso-rescale
#| fig-cap: "SLI-isometry's fitted coefficient against the ground truth, raw (left) and variance-corrected (right; main text, Effect classification and this section). Longer, Fatter, and Proportionally smaller species, matching the main text's calibration scatter; the dashed line is 1:1, not a fitted trend."
#| fig-width: 8
#| fig-height: 4.5
#| fig-pos: 'H'

Sli_iso_rescale_long <- Sli_iso_rescale_df %>%
  pivot_longer(cols = c(beta_iso_true, beta_T_rescaled), names_to = "Version", values_to = "x") %>%
  mutate(Version = if_else(Version == "beta_iso_true", "Raw", "Variance-corrected"),
         Version = factor(Version, levels = c("Raw", "Variance-corrected")))

Sli_iso_rescale_stats_long <- tibble::tibble(
  Version = factor(c("Raw", "Variance-corrected"), levels = c("Raw", "Variance-corrected")),
  r = c(r_sli_iso_raw, r_sli_iso_rescaled)
)

ggplot(Sli_iso_rescale_long, aes(x = x, y = b_temp_inc, color = Temp_eff, shape = Scaling)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = .55, size = 1.2) +
  geom_text(data = Sli_iso_rescale_stats_long, aes(label = paste0("r=", r)),
            x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5, inherit.aes = FALSE, size = 3) +
  facet_wrap(vars(Version), scales = "free") +
  labs(x = "Ground truth", y = expression(hat(beta)[T] ~ "(SLI-isometry, estimated)"),
       color = "Category", shape = "Scaling") +
  guides(color = guide_legend(nrow = 1), shape = guide_legend(nrow = 1)) +
  theme(legend.position = "top", legend.box = "vertical")
```

The variance correction takes the correlation from `r round(r_sli_iso_raw,2)` to `r round(r_sli_iso_rescaled,4)`, and the points sit almost exactly on the 1:1 line once it's applied. This isn't unique to SLI-isometry — every one of the six methods has a closed-form standardized coefficient that reaches a similarly exact correlation (0.9999–1.0000) against its own correctly-scaled natural anchor (e.g. Mass-as-covariate against the textbook standardized partial-regression coefficient $(\rho_{13}-\rho_{12}\rho_{23})/(1-\rho_{23}^2)$), confirming that the ground truth used in the main text is not merely correlated with SLI-isometry's formula but algebraically equivalent to it up to this one scaling factor — which is precisely why SLI-isometry's own strong performance should be read as confirmation that the simulation and ground-truth calculation behave as expected, not as a comparative win over the other five approaches (main text, Approach classification).

# Mixed - Wingier and Mixed - Fatter empirical species

In addition to the Bergmann's and Inverse Bergmann's species reported in the main text (Empirical case studies section), some species in the empirical datasets showed **Mixed - Wingier** responses: wing length increasing while body mass decreased, so that Allen's and Bergmann's rules reinforce one another. This is the empirical analogue of the simulated species whose mass and wing responses combine most strongly to produce an unambiguously longer relative appendage. A single Atlantic Forest species showed the opposite **Mixed - Fatter** response (wing length decreasing while body mass increased; main text species-accounting table).

```{r}
#| label: load-mixed-data
#| include: false

study_order  <- c("Nightjar", "Weeks (2020)", "Atlantic birds")
study_colors <- c("Nightjar" = "#E41A1C", "Weeks (2020)" = "#377EB8", "Atlantic birds" = "#4DAF4A")

approach_labs <- c(
  "Ratio"     = "Wing / Mass",
  "Ratio2"    = "Wing² / Mass",
  "Ryding"    = "Mass as covariate",
  "Resid_ols" = "OLS residuals",
  "Sli_est"   = "SLI estimated",
  "Sli_iso"   = "SLI isometry"
)

shape_scale <- c("Both" = 15, "Mass" = 16, "Neither" = 17, "Wing" = 18)

load_direction <- function(direction) {
  bind_rows(
    read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
    read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
    read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
  ) %>%
    filter(
      Direction == direction,
      Approach  %in% c("Ratio", "Ratio2", "Ryding", "Resid_ols", "Sli_est", "Sli_iso"),
      std.error < 0.5
    ) %>%
    mutate(
      Sig_trait = str_to_sentence(Sig_trait),
      Approach  = factor(Approach, levels = c("Ratio", "Ratio2", "Ryding", "Resid_ols", "Sli_est", "Sli_iso"))
    )
}

# Builds one panel: orders species by mean estimate within study, colors axis text by study (matching main-text @fig-empirical), and optionally shows a legend.
build_mixed_plot <- function(df_panel, show_legend = FALSE) {
  species_meta_p <- df_panel %>%
    group_by(species, Study) %>%
    summarise(mean_est = mean(estimate, na.rm = TRUE), .groups = "drop") %>%
    mutate(Study = factor(Study, levels = study_order)) %>%
    arrange(Study, mean_est)

  species_levels_p <- species_meta_p$species
  axis_colors_p    <- study_colors[as.character(species_meta_p$Study)]

  p <- df_panel %>%
    mutate(species = factor(species, levels = species_levels_p)) %>%
    ggplot(aes(x = species, y = estimate, color = Approach,
               group = interaction(species, Approach))) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_errorbar(aes(ymin = LCI95, ymax = UCI95),
                  alpha = 0.8, width = 0,
                  position = position_dodge(width = 0.75)) +
    geom_point(aes(shape = Sig_trait), size = 1.5,
               position = position_dodge(width = 0.75)) +
    scale_color_hue(labels = approach_labs) +
    scale_shape_manual(values = shape_scale, drop = FALSE) +
    labs(x = NULL, y = expression(beta ~ "on wing shape"), color = NULL, shape = NULL) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8, color = axis_colors_p))

  if (show_legend) {
    p <- p + theme(legend.position = "top", legend.box = "horizontal")
  } else {
    p <- p + theme(legend.position = "none")
  }
  p
}

parms_wingier <- load_direction("Mixed - Wingier")
parms_fatter  <- load_direction("Mixed - Fatter")

n_mixed_wingier <- n_distinct(parms_wingier$species)
n_mixed_fatter  <- n_distinct(parms_fatter$species)

fig_cap_mixed <- glue::glue("Estimated effect of the environmental gradient on relative wing length across three empirical datasets (Nightjars — red labels; North American migratory birds — blue labels; Atlantic Forest birds — green labels): (a) Mixed - Wingier species (wing length increasing while body mass decreased; $n$ = {n_mixed_wingier}) and (b) Mixed - Fatter species (wing length decreasing while body mass increased; $n$ = {n_mixed_fatter}). Only species with a significant ($p < 0.05$) mass or wing response are shown. Points are mean parameter estimates; error bars are 95% confidence intervals.")
```

```{r}
#| label: fig-mixed-wingier
#| fig-cap: !expr fig_cap_mixed
#| fig-width: 8
#| fig-height: 6
#| fig-pos: 'H'

p_wingier <- build_mixed_plot(parms_wingier)
p_fatter  <- build_mixed_plot(parms_fatter)
shared_legend_mixed <- ggpubr::get_legend(build_mixed_plot(parms_wingier, show_legend = TRUE))

mixed_panels <- plot_grid(p_wingier, p_fatter, labels = c("a", "b"), label_size = 10,
                           ncol = 2, rel_widths = c(14, 1))
plot_grid(shared_legend_mixed, mixed_panels, ncol = 1, rel_heights = c(0.08, 1))
```

Across `r n_mixed_wingier` Mixed - Wingier species (panel a), the rank-order pattern among methods is visibly more variable than for Bergmann's or Inverse Bergmann's species (main text Empirical case studies section), consistent with the absence of a clear rank-order signal in the analogous simulated scenarios (main text, Simulation performance). The single Mixed - Fatter species (panel b) is shown for completeness; with only one species, no rank-order pattern can be assessed.

# Stable empirical species

Species showing no significant ($p < 0.05$) change in either mass or wing length along the environmental gradient were classified as **Stable** (main text classification table) and excluded from the main-text rank-order comparison, since the predicted rank order (Ratio \> SLI \> OLS, or its reverse) is only defined relative to a directional shape response.

```{r}
#| label: load-stable-data
#| include: false

parms_stable <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  filter(
    Direction == "Stable",
    Approach  %in% c("Ratio", "Ratio2", "Ryding", "Resid_ols", "Sli_est", "Sli_iso"),
    std.error < 0.5
  ) %>%
  mutate(
    Sig_trait = str_to_sentence(Sig_trait),
    Approach  = factor(Approach, levels = c("Ratio", "Ratio2", "Ryding", "Resid_ols", "Sli_est", "Sli_iso"))
  )

species_meta_stable <- parms_stable %>%
  group_by(species, Study) %>%
  summarise(mean_est = mean(estimate, na.rm = TRUE), .groups = "drop") %>%
  mutate(Study = factor(Study, levels = study_order)) %>%
  arrange(Study, mean_est)

species_levels_stable <- species_meta_stable$species
axis_colors_stable     <- study_colors[as.character(species_meta_stable$Study)]
n_stable                <- n_distinct(parms_stable$species)

fig_cap_stable <- glue::glue("Estimated effect of the environmental gradient on relative wing length for Stable species (no significant change in mass or wing length) across three empirical datasets (Nightjars — red labels; North American migratory birds — blue labels; Atlantic Forest birds — green labels); $n$ = {n_stable} species. Points are mean parameter estimates; error bars are 95% confidence intervals. Because neither trait changed significantly, these species have no true directional shape signal to recover, and the six methods are not expected to agree on rank order.")
```

```{r}
#| label: fig-stable
#| fig-cap: !expr fig_cap_stable
#| fig-width: 8
#| fig-height: 6
#| fig-pos: 'H'

parms_stable %>%
  mutate(species = factor(species, levels = species_levels_stable)) %>%
  ggplot(aes(x = species, y = estimate, color = Approach,
             group = interaction(species, Approach))) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = LCI95, ymax = UCI95),
                alpha = 0.8, width = 0,
                position = position_dodge(width = 0.75)) +
  geom_point(aes(shape = Sig_trait), size = 1.5,
             position = position_dodge(width = 0.75)) +
  scale_color_hue(labels = approach_labs) +
  scale_shape_manual(values = shape_scale, drop = FALSE) +
  labs(x = NULL, y = expression(beta ~ "on wing shape"), color = NULL, shape = NULL) +
  theme(
    axis.text.x  = element_text(angle = 60, hjust = 1, size = 8, color = axis_colors_stable),
    legend.position = "top"
  )
```

Across `r n_stable` Stable species, estimates cluster closer to zero than in the directional categories, as expected, though the six methods still disagree in both magnitude and, occasionally, sign — underscoring that even a "null" result can look different depending on the method used.

# EXTRAS - delete

These classical approaches share a critical assumption: that body size is measured without error. This is the operating assumption of ordinary least squares regression and is retained in most implementations of multiple regression, maximum likelihood, and Bayesian models unless the model is explicitly formulated otherwise. In practice, this assumption is nearly always violated. Mass — the most commonly used body size proxy — reflects structural body size but also transient biological variation such as gut fill, hydration state, and reproductive condition. Instrument and observer error introduce additional noise. When predictor variables are measured with error, OLS slope estimates are attenuated toward zero — a bias that grows as the correlation between the predictor and response weakens [@wartonBivariateLinefittingMethods2006; @greenMassLengthResiduals2001]. By ignoring this attenuation, classical regression approaches underestimate the allometric relationship between mass and appendage length, which propagates as bias into downstream estimates of temperature's effect on shape.

For instance, @rydingMultipleRegressionNot2025 found that multiple regression was unbiased in their simulation study, which targeted a different causal structure. These findings are therefore not contradictory, but instead reflect sensitivity to differing assumptions about the underlying causal system.
