---
title: "The Shape of Things to Come: Simulations Show Promise in a New Metric, the Standardized Length Index"
number-sections: true
execute:
  echo: false
  message: false
  warning: false
format: #docx
  elsevier-pdf: 
    toc: false
    header-includes:
      - |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhead[R]{SLI to examine Allen's rule}
        \usepackage{makecell}
        \usepackage{lineno}
        \linenumbers
        \usepackage[margin=1in]{geometry}
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
        \usepackage{tabularray}
    journal:
      name: Methods in Ecology and Evolution
      formatting: preprint #doubleblind suppresses authors!, return to preprint
      cite-style: authoryear
bibliography: ../Suppfiles/Morphology.bib
link-citations: true
csl: ../Suppfiles/methods-in-ecology-and-evolution.csl
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
editor_options:
  chunk_output_type: console
metadata-files: 
  - ../Suppfiles/abstract_keywords.yml
  - ../Suppfiles/authors.yml
---

{{< pagebreak >}}

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(readxl)
library(janitor)
library(png)
library(grid)
library(ggpubr)
library(cowplot)
library(scales)
library(smatr)
library(MASS)
library(ggpmisc)
library(dagitty)
library(ggdag)
ggplot2::theme_set(theme_cowplot())
```

```{r}
#| label: format-Lr-tbl
#| include: false

Lit_review <- read_excel("../Lit_review/Methods_ecogeographic_rules_5.7.26.xlsx",
                         sheet = "Allens") %>%
  clean_names()

Unclear_exclude <- Lit_review %>% filter(include_exclude == "Exclude")
Lr <- Lit_review %>%
  anti_join(Unclear_exclude) %>%
  dplyr::select(authors, year, regression_used, class, starts_with("approach")) %>%
  mutate(study = factor(row_number()))

Lr_pivot <- Lr %>%
  pivot_longer(cols = c(starts_with("approach"), -approach_unclear),
               names_to = "approach",
               values_to = "primary_v_secondary") %>%
  mutate(approach = str_remove(approach, "approach_")) %>%
  filter(!is.na(primary_v_secondary) & primary_v_secondary != "M") %>%
  mutate(approach_unclear = ifelse(is.na(approach_unclear), "No", "Yes")) %>%
  filter(approach_unclear == "No")

Lr_p <- Lr_pivot %>%
  summarize(num_studies = n(), .by = c(approach, class)) %>%
  mutate(Tot_studies = sum(num_studies),
         Prop = num_studies / Tot_studies,
         .by = class) %>%
  mutate(Per = Prop * 100,
         approach_label = str_to_sentence(approach),
         approach_label = str_replace_all(approach_label, "_", "\n"),
         approach_label = as.factor(approach_label)) %>%
  mutate(Per = round(Per, 0), Prop = round(Prop, 2)) %>%
  arrange(desc(Per))

Lr_stack <- Lr_pivot %>%
  arrange(class, study) %>%
  mutate(ypos = row_number(), .by = c(approach, class)) %>%
  left_join(Lr_p[, c("approach", "approach_label", "class", "num_studies")])

Lr_p2 <- Lr_p %>%
  group_by(class) %>%
  mutate(approach_facet = fct_reorder(paste(class, approach_label, sep = "___"),
                                      num_studies, .desc = TRUE),
         Per = paste0(Per, "%")) %>%
  ungroup()

Lr_stack2 <- Lr_stack %>%
  left_join(Lr_p2 %>% distinct(class, approach_label, approach_facet),
            by = c("class", "approach_label"))
```

```{r}
#| label: percentages-methods
#| include: false

Per_class <- Lr %>%
  summarize(Percent = n() / nrow(Lr) * 100, .by = class) %>%
  mutate(Percent = paste0(round(Percent, 0), "%"))

per_aves <- Per_class %>% filter(class == "Aves") %>% pull(Percent)

pull_per_aves <- function(approach_name) {
  result <- Lr_p %>%
    filter(class == "Aves" & str_detect(approach, approach_name)) %>%
    pull(Per)
  if (length(result) == 0) 0 else result
}

per_mv_reg   <- pull_per_aves("multivariate")
per_residual <- pull_per_aves("two_step|residual")
per_ratios   <- pull_per_aves("^ratio")
per_raw      <- pull_per_aves("raw")
```

```{r}
#| label: source-key-fns
#| include: false
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: load-simulation-results
#| include: false

# Simulation is run separately by Scripts/Run_simulation.R (re-run it any time the parameter grid, sample size, or evaluation logic changes) and saved to Derived/Rds/simulation_results.rds
# Both this manuscript and supplementary_info.qmd just read the result. This brings N_spp, Parms_mat3, Parms_tbl4, Eval_tbl, and every inline-cited percentage into scope by name.

list2env(readRDS("Derived/Rds/simulation_results.rds"), envir = .GlobalEnv)
```

# Introduction

Wildlife around the world are responding to contemporary climate change with shifts in morphology. Studies have documented declines in body size consistent with Bergmann's rule across diverse taxa, and growing evidence suggests that body shape is also changing: appendage-to-body-size ratios are shifting in birds, mammals, and other endotherms in a pattern broadly consistent with Allen's rule [@rydingShapeshiftingChangingAnimal2021; @rydingLongShorttermResponses2024]. Because Allen's rule is fundamentally about shape, that is, the proportional length of an appendage relative to overall body size, testing it requires methods that can distinguish variation in appendage length associated with the environmental gradient from variation attributable to body size [@symondsItsAllRelative2025]. Yet the statistical methods used to quantify relative appendage length have received far less attention than the biological hypothesis itself.

A literature review of studies testing Allen's rule since @rydingShapeshiftingChangingAnimal2021's key review reveals that three broad approaches account for nearly all empirical work (@fig-lit-review). The simplest uses absolute appendage length as the response variable, regressed against temperature or spatial or temporal proxies (e.g., latitude or year), hereafter referred to as the 'gradient of interest'. Throughout this paper I focus on temperature, although the methods apply equally to any of these gradients, and refer to the corresponding regression coefficient simply as $\beta_T$. This approach has been compared with methods that explicitly account for body size elsewhere [e.g., @symondsItsAllRelative2025], so I do not consider it further. A second approach, currently the most common, includes body size as a covariate in a multiple regression, interpreting $\beta_T$ as the effect of temperature on relative appendage length. A third approach takes the ratio of appendage length divided by a body size metric as the response variable. Despite their widespread use, both approaches have important statistical limitations.

Ratio-based metrics are frequently correlated with body size itself, limiting their utility as measures of relative shape [@jakobEstimatingFitnessComparison1996; @greenMassLengthResiduals2001], yet they remain widely used [e.g., @jirinecMorphologicalConsequencesClimate2021; @dubinerWidespreadRecentChanges2022]. The limitations of ordinary least squares (OLS) regression-based methods are more subtle. OLS regression attributes all unexplained variation around the fitted allometric relationship to appendage length rather than body size. Consequently, every deviation from the fitted line is interpreted as evidence of an unexpectedly long or short appendage, rather than allowing deviations to be shared between both variables (@nte-sma-intuition). This assumption is often difficult to justify in observational studies of morphology. Importantly, this same assumption underlies the vast majority of regression models, including maximum likelihood and Bayesian implementations, unless explicitly formulated otherwise.

Standardized major axis (SMA) regression provides an alternative line-fitting approach by treating variation in body size and appendage length symmetrically rather than assigning all unexplained variation to the response variable (@wartonBivariateLinefittingMethods2006; @nte-sma-intuition). Consequently, SMA estimates steeper allometric slopes whenever the correlation between $X$ and $Y$ is imperfect. SMA has long been recommended for studies of allometric scaling [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009; @greenMassLengthResiduals2001] and has been proposed as the preferred regression approach in studies of Allen's rule [@santoroAllometryEvaluateAllens2022], yet its performance has never been systematically compared with the classical methods used to quantify relative appendage length.

Here I introduce the **Standardized Length Index (SLI)**, an analogue of the Standardized Mass Index [@peigNewPerspectivesEstimating2009] adapted to measure appendage length relative to body size. The SLI uses the SMA slope to standardize each individual's appendage length to a common reference body size, producing a size-corrected shape index in the original units of the appendage. I present two parameterisations: SLI-isometry, which fixes the scaling exponent at 0.33 (the isometric expectation for a linear versus volumetric measurement like mass), and SLI-estimated, which estimates the exponent empirically from the data. Using simulated data from `r N_spp` hypothetical species spanning a realistic range of allometric relationships and temperature sensitivities, I show that both SLI approaches outperform the classical methods. I validate the rank-order pattern of method performance across three independent empirical datasets, and provide recommendations for empirical studies of Allen's rule.

::: {#nte-sma-intuition .callout-note icon="false"}
## Building intuition for SMA regression

SMA and OLS regression differ in two closely related ways. First, SMA regression estimates a steeper allometric slope whenever the correlation between variables is \< 1. Second, because it treats variation in both variables symmetrically, relative appendage length and relative body size become reciprocal descriptions of the same deviation from the allometric relationship. The equations below show why these properties arise.

*SMA estimates a steeper allometric slope*

There is a direct mathematical link between the slopes of OLS and SMA regression.\
The slope of a predictor variable $X$ (e.g., body size) is related to the response variable $Y$ (e.g., appendage length) in OLS regression via the following equation:

- $\beta_{ols} = r * \frac{Sy}{Sx}$ where $r$ is the correlation between $X$ and $Y$, and $S$ is the standard deviation

The slope of SMA regression is estimated by:

- $\beta_{sma} = sign(r) * \frac{Sy}{Sx}$ , i.e., if $r$ is negative then $\beta_{sma}$ is also negative.

A simple substitution of $\frac{Sy}{Sx}$ in the first equation shows that

$$\beta_{ols} = r * \beta_{sma}$$

This makes it evident that as $r$ increases, $\beta_{\text{OLS}} \rightarrow \beta_{\text{SMA}}$, and that the difference in estimated slopes will be highest when $X$ and $Y$ are only weakly correlated.

*SMA gives a symmetric geometric interpretation*

Unlike OLS, the SMA line of best fit is unchanged when the $X$ and $Y$ axes are exchanged (@fig-sma_flip_axes), leading to a useful geometric property: swapping the axes simply reflects the points across the fitted line. Consequently, the individual highlighted in green can be described either as having relatively short appendages for its body size (panel A) or as having a relatively large body for its appendage length (panel B). These are reciprocal descriptions of the same deviation from the allometric relationship (i.e., residuals retain the same magnitude but reverse sign). In contrast, because the OLS line changes when the axes are exchanged, this reciprocity does not hold.

![In SMA regression the X and Y axes can be flipped and the line of best fit will not change. Here I flip the axes and plot the OLS (red) & SMA (blue) regression lines of best fit, as well as the same individual from my hypothetical species (highlighted in green). SMA residuals are equally correlated with both $X$ and $Y$ ($|r_{sma}| = 0.59$, whereas OLS residuals remain uncorrelated with the predictor but strongly correlated with the response ($r_{ols} = 0.96$).](../Figures/sma_flip_axes.png){#fig-sma_flip_axes width="100%"}
:::

# Materials and Methods

## Literature review

I reviewed studies published between 2021 and 2025 that quantified relative appendage length in the context of Allen's rule in endotherms. Papers were identified by searching for the keywords Allen\* AND relative appendage OR limb OR body shape AND climate change OR warming OR temperature OR latitude, supplemented by reference-list searches. I recorded the primary analytical approach used by each study and classified it into one of five categories: ratio-based methods, two-step residual regression, multiple regression with body size as a covariate, use of absolute appendage length only, or other. In total, I scored `r nrow(Lit_review)` papers, of which `r nrow(Unclear_exclude)` were excluded for not meeting inclusion criteria. Studies with unclear methodology were excluded from approach tallies. The majority of studies (`r per_aves`) examined Allen's rule in birds; the remainder used mammal focal taxa.

```{r}
#| label: fig-lit-review
#| fig-cap: "Frequency of the primary methodological approaches used in studies testing Allen's rule since 2021. Each point represents one study; colored lines connect studies that used more than one approach, indicating non-independence. Percentages are computed within each taxonomic class as a proportion of all approach-uses rather than all studies."
#| fig-height: 6.5
#| fig-pos: 'H'

ggplot() +
  geom_col(data = Lr_p2,
           aes(x = approach_facet, y = num_studies),
           fill = "grey85", color = "black") +
  geom_point(data = Lr_stack2,
             aes(x = approach_facet, y = ypos, color = study),
             alpha = 0.7, size = 3) +
  geom_line(data = Lr_stack2,
            aes(x = approach_facet, y = ypos, group = study, color = study),
            alpha = 0.75) +
  geom_text(data = Lr_p2,
            aes(x = approach_facet, y = num_studies, label = Per),
            vjust = -0.7, size = 4.5) +
  facet_wrap(~class, scales = "free_x",
             labeller = labeller(.default = label_value)) +
  scale_x_discrete(labels = function(x) sub(".*___", "", x)) +
  theme(axis.text.x = element_text(size = 9, vjust = 0.58, angle = 55),
        legend.position = "none") +
  labs(x = NULL, y = "Number of studies")
```

## Data simulation

I simulated appendage length, body size (mass), and a temperature index for `r N_spp` hypothetical species, each represented by approximately `r N_ind` individuals. Data were drawn from a multivariate log-normal distribution:

$$\begin{bmatrix}
\log(\text{Appendage}) \\
\log(\text{Mass}) \\
\text{Temperature}
\end{bmatrix}
\sim \mathrm{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

with $\boldsymbol{\mu} = [\log(180),\ \log(80),\ 1.0]^\top$ representing a species with a mean wing length of 180 mm and mean mass of 80 g over an average 1°C temperature increase. Simulating from a joint distribution rather than from a directional model (e.g., $Appendage \sim Mass + Temperature + \epsilon$) ensures that no method is advantaged by design. The covariance matrix $\boldsymbol{\Sigma}$ was parameterised by three Pearson correlations ($\rho_{12}$: appendage–mass; $\rho_{13}$: appendage–temperature; $\rho_{23}$: mass–temperature) and the allometric scaling slope. Hypothetical species were constructed by crossing the following parameter values: appendage–mass correlation $\rho_{12} \in \{`r paste(r_12, collapse = ", ")`\}$; temperature–morphology correlations $\rho_{13} = \rho_{23} \in \{`r paste(r_13, collapse = ", ")`\}$; and allometric scaling slope $b \in \{`r paste(b_avg_12, collapse = ", ")`\}$, representing hypoallometry, isometry, and hyperallometry respectively. Standard deviations of log-mass and log-appendage were set so that the average of the OLS and SMA slopes equalled the specified allometric slope, further ensuring that neither SMA nor OLS regression was advantaged by the simulation structure.

Species were classified by the direction of temperature's effect on shape: **Longer** ($\rho_{13} > \rho_{23}$, temperature reduces mass more than appendage length), **Fatter** ($\rho_{13} < \rho_{23}$, temperature reduces appendage more than mass), **Proportionally smaller** ($\rho_{13} = \rho_{23} < 0$, temperature reduces both equally), and **Proportionally larger** ($\rho_{13} = \rho_{23} > 0$, temperature increases both equally). Fatter, Proportionally smaller, and Longer are all different realizations of Bergmann's rule as warming always decreases overall body size. Combinations where temperature had no effect on either trait ($\rho_{13} = \rho_{23} = 0$) were excluded. I additionally simulated species that increase in size with warmer temperatures, in line with several recent studies that have documented inverse Bergmannian clines in mammals [@stapelfeldtLongtermFieldStudy2023; @castillo-figueroaDoesBergmannsRule2022; @jumanRecentRapidEcogeographical2022]. The 'Bigger' scenario was the mirror image of the Proportional category in which temperature increases wing length and body mass by equal amount ($b \in \{0.22, 0.33, 0.44\}$, $\rho_{12} \in \{0.3, 0.45, 0.6\}$, $\rho_{13} = \rho_{23} \in \{0.15, 0.3, 0.5, 0.7\}$) — species that grow uniformly larger with warming without any true change in shape. To validate that the simulated direction of temperature's effect on shape matched the intended direction I examined the SMA intercepts, where larger SMA intercepts represent proportionally longer appendages for a given body size [@shingletonAllometryStudyBiological2010], across binned temperature values (@fig-temp-body-shape-plot). Simulated datasets where the correlation had the wrong sign due to simulation noise were removed (`r nrow(Sim_fail)` of `r N_spp` species; Supplementary Figure X), yielding a final sample of `r nrow(Parms_tbl3)` hypothetical species.

```{r}
#| label: ex-parms-select
#| include: false

Ex_parms <- Parms_mat3 %>%
  filter(Scaling == "Hyperallometry" & r_13 != 0 & r_23 != 0) %>%
  slice_max(n = 1, order_by = tibble(Strength, r_12),
            by = Temp_eff, with_ties = FALSE) %>%
  filter(Temp_eff %in% c("Fatter", "Longer")) %>%
  dplyr::select(-Strength)
```

```{r}
#| label: ex-parms-temp-body-shape
#| include: false

Ex_df_ex <- gen_ex_data(Ex_parms)

Temp_eff_vec <- setNames(c("Fatter", "Longer"), c("Fatter", "Longer"))
ex_parms_temp_bs <- map(Temp_eff_vec, \(shp) {
  df <- Ex_df_ex %>% filter(Temp_eff == shp)
  bind_rows(
    format_sma_parms(run_sma_mod(df, interaction = FALSE)) %>% mutate(Slopes_vary = "No"),
    format_sma_parms(run_sma_mod(df, interaction = TRUE))  %>% mutate(Slopes_vary = "Yes")
  )
}) %>% list_rbind(names_to = "Temp_eff")
```

```{r}
#| label: fig-temp-body-shape-plot
#| fig-cap: The allometric scaling slope determines the rate of increase of appendage length with an increase in body size. The intercept reflects the proportions of the appendage relative to body size; when slopes are constant, the group with the higher intercept has proportionally larger appendages across all body sizes. As temperature increases, species were simulated to get fatter (left panel) or longer (right panel). The figure represents a model where SMA slopes were held constant while the intercept was allowed to vary.
#| fig-pos: 'H'

ex_parms_temp_bs %>%
  filter(Slopes_vary == "No") %>%
  mutate(x = 0) %>%
  ggplot() +
  geom_abline(aes(intercept = elevation, slope = slope, color = Temp_inc)) +
  ggrepel::geom_text_repel(aes(x = x, y = elevation, label = Temp_label)) +
  labs(x = "Log mass", y = "Log appendage") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) +
  guides(color = "none") +
  facet_wrap(~Temp_eff, scales = "free_y")
```

## Statistical approaches

I applied six methods to each simulated dataset, log-transforming all variables and then standardising to unit variance before fitting regression models so that coefficients were comparable across species and methods.

**Ratio approaches.** (1) The appendage-to-mass ratio $log(A / S)$ was regressed on temperature. (2) The squared appendage-to-mass ratio $log(A^2 / S)$ was regressed on temperature, reflecting a surface-area-to-volume formulation used in some empirical studies [e.g., @nwaoguTemperatureAridityDetermine2018; @dubinerWidespreadRecentChanges2022].

**Two-step residual regression (OLS residuals).** Appendage was regressed on mass using OLS, and the resulting residuals were regressed on temperature.

**Multiple regression.** Appendage was modelled as a linear function of mass and temperature simultaneously; the temperature coefficient $\beta_T$ was interpreted as the effect of temperature on relative appendage length.

**Standardized Length Index — isometry (SLI-isometry).** Following @peigNewPerspectivesEstimating2009, I computed:

$$SLI_i = A_i \cdot \left(\frac{M_0}{M_i}\right)^{b_{SLI}}$$

where $A_i$ is the appendage length of individual $i$, $M_i$ is the mass of individual $i$, $M_0$ is the mean mass of the species, and $b_{SLI} = 0.33$ (the isometric scaling exponent for a linear versus volumetric measurement). The SLI reprojects each individual's appendage length to the value it would take at the population mean body size, then the SLI was regressed on temperature.

**Standardized Length Index — estimated (SLI-estimated).** Identical to SLI-isometry but with $b_{SLI}$ replaced by the empirical SMA slope of $\log(A)$ on $\log(S)$ estimated from the data.

## Evaluation

Because data were simulated from a joint distribution without a specified effect size, there is no single true $\beta_T$ to recover. I therefore evaluated methods on whether they correctly recovered the **direction** of temperature's effect on relative appendage length: a positive $\hat{\beta}_T$ for species becoming relatively longer, and a negative $\hat{\beta}_T$ for species becoming relatively fatter. For species simulated as Proportional or Bigger — categories with no true shape signal, differing only in whether overall body size shrinks or grows with temperature — I compared the proportion classified as longer or fatter against the true proportion (approximately 50%). An ideal method scores 1.0 on directional accuracy for Longer and Fatter species, and classifies approximately 50% of Proportional and Bigger species in each direction.

## Empirical case studies

I applied all six methods to three independent empirical datasets to assess whether the rank-order pattern observed in the simulations generalised to real data. First, I used morphological data from three migratory nightjar species, European nightjars (Caprimulgus europaeus), eastern whip-poor-wills (Antrostomus vociferus), and common nighthawks (Chordeiles minor), spanning a spatial temperature gradient [@skinnerEnvironmentalGeographicConditions2025]. Second, I applied the methods to the dataset of @weeksSharedMorphologicalConsequences2020, comprising 43 North American migratory bird species (1979–2016), using capture year as the temporal gradient. Third, I applied the methods to a dataset of 49 Atlantic Forest bird species from @rodriguesATLANTICBIRDTRAITS2019, using mean annual temperature as the environmental gradient. Together, these datasets span different taxonomic groups, ecological systems, gradients of interest, and putative selective pressures, providing a broad test of whether simulation results generalise across empirical contexts.

Age and sex can influence body size independently of the environmental gradient (e.g., through sexual size dimorphism or age-related growth). For each species, I therefore tested whether age or sex covariates predicted mass or wing length while controlling for the gradient variable (i.e., temperature or year). Covariates were retained only when age or sex influenced body size ($p < 0.10$) and when each age or sex group contained at least 100 individuals.

Retained covariates were incorporated in only certain methods. For the ratio ($A/S$) and squared-ratio ($A^2/S$) I assume that age and sex influence wing and mass roughly proportionally, and thus would not importantly modify the ratios. I do not include age and sex in the SLI-isometry approach because $b_{SLI}$ is fixed at 0.33 by definition. For the OLS-residual approach, age and sex were included in the first-stage regression of wing on mass. For multiple-regression, they were included alongside mass and the gradient variable. For SLI-estimated, SMA slopes were estimated separately for each age and sex group and then averaged depending on the demographic status of each individual (e.g., a hatch-year male would receive use a different SMA slope than an adult male).

To evaluate whether the simulation results generalized to empirical data, I averaged the gradient coefficients within each method family (Ratio: $A / S$ Ratio and $A^2 / S$ Ratio; OLS: multiple regression and OLS-residual; SLI: SLI-isometry and SLI-estimated) for each species and compared their rank order with the simulation predictions. Because the expected ordering depends on the direction of body size change, species were first classified according to whether they exhibited Bergmannian, inverse Bergmannian, stable, or mixed size trends (@tbl-classification). Bergmannian species exhibited declines in wing length, body mass, or both along the gradient, whereas inverse Bergmannian species exhibited increases in one or both traits ($p < 0.05$). Stable species showed no significant change in either trait, while mixed species exhibited opposing trends in wing length and body mass. Only Bergmannian and inverse Bergmannian species were used to evaluate agreement with the simulation results because these scenarios correspond directly to the simulated data-generating process. Species with stable or mixed size trends are reported separately in the Supplementary Information. Finally, because SMA regression is unreliable when the correlation between wing length and body mass is weak [@smithUseMisuseReduced2009], SLI-estimated was calculated only for species with $r \geq 0.3$.

```{r}
#| label: tbl-classification
#| tbl-cap: "Classification of species according to significant ($p < 0.05$) changes in wing length and body mass along the environmental gradient (temperature or year). Non-significant coefficients did not contribute to classification."

Classification_tbl <- tibble::tribble(
  ~Category,               ~`Classification criterion`,
  "Bergmann's",            "Wing length, body mass, or both decrease",
  "Inverse Bergmann's",    "Wing length, body mass, or both increase",
  "Mixed - Wingyier",      "Wing length increases while body mass decreases",
  "Mixed - Fatter",        "Wing length decreases while body mass increases",
  "Stable",                "No significant change in either trait"
)
knitr::kable(Classification_tbl)
```

# Results

## Simulation performance

```{r}
#| label: plot-approaches-fun
#| include: false

plot_approaches <- function(df, much_longer = FALSE, ylim = TRUE,
                             x_txt_size = 9, legend.pos = "right") {
  df <- df %>%
    mutate(r_12 = as_factor(r_12),
           Model = str_replace_all(Model, x_labs))
  if (!much_longer) { df <- df %>% filter(Temp_eff != "Inverse") }
  if (much_longer)  { df <- df %>% filter(Temp_eff == "Inverse") }

  p <- df %>%
    ggplot(aes(x = Model, y = b_temp_inc)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_boxplot(position = position_dodge(width = .75), outlier.shape = NA) +
    labs(x = NULL, y = expression(hat(beta)[T] ~ "on relative appendage length")) +
    facet_wrap(vars(Temp_eff)) +
    theme(
      axis.text.x = element_text(size = x_txt_size, vjust = .58, angle = 55),
      legend.position = legend.pos
    )

  if (ylim) p <- p + ylim(c(-.25, .25))

  if (!much_longer) {
    p <- p + geom_point(
      position = position_jitterdodge(jitter.width = 2),
      alpha = .5, size = 2.5,
      aes(color = Strength, shape = Scaling, group = Model)
    )
  } else {
    p <- p + geom_point(
      position = position_jitterdodge(jitter.width = 2),
      alpha = .5,
      aes(color = Strength, shape = factor(r_23), size = abs(r_13), group = Model)
    ) +
    guides(size = guide_legend("r_13"), shape = guide_legend("r_23"))
  }
  p
}
```

The six methods produced strikingly different estimates of temperature's effect on relative appendage length, and the pattern was consistent across allometric scaling categories and correlation structures (@fig-compare-approaches).

```{r}
#| label: fig-compare-approaches
#| fig-cap: "Estimated effect of temperature on relative appendage length ($\\hat{\\beta}_T$) for each method across Fatter, Proportional, and Longer species — three different realizations of Bergmann's rule, differing only in how fast appendage length declines with warming relative to mass (faster, equally, or slower, respectively). Positive values indicate that species become relatively longer with increasing temperature; negative values indicate they become relatively fatter. Each point is a simulated species. The dashed line at zero indicates no effect. Point shape indicates allometric scaling category and colour indicates the strength of the difference between temperature's effect on appendage length and mass. The 'Proportional' panel contains species with no true shape signal, so the expected value is near zero for unbiased methods."
#| fig-width: 9
#| fig-height: 4.5
#| fig-pos: 'H'

plot_approaches(Parms_tbl4 %>% filter(Temp_eff != "Bigger"),
                x_txt_size = 11, much_longer = FALSE,
                legend.pos = "top", ylim = FALSE)
```

**Ratio approaches** produced consistently overestimated effects. Both $A/S$ and $A^2/S$ assigned large positive $\hat{\beta}_T$ values to species simulated to become relatively longer, but also classified most proportional species as becoming relatively longer — and even many species simulated to become relatively fatter. This positive bias arises mechanically: when temperature reduces body size (Bergmann's rule), the denominator of the ratio decreases, inflating the ratio regardless of whether the appendage changed.

**Multiple regression and OLS residuals** produced consistently underestimated — and often reversed — effects. Both methods returned negative $\hat{\beta}_T$ in the majority of species simulated to become relatively longer, correctly classified Fatter species, and overcorrected in Proportional species (@fig-eval). The two approaches behaved nearly identically, consistent with their mathematical equivalence when predictor variables are uncorrelated.

**SLI-isometry** was among the most accurate methods across the standard shape-response categories. It correctly classified the direction of temperature's effect in `r Sli.iso_longer_right`% of species simulated to become relatively longer, and `r Sli.iso_fatter_right`% of species simulated to become relatively fatter (@fig-eval). Among species simulated as proportional, it classified approximately `r Sli.iso_proportional_pos`% as becoming relatively longer, close to the true proportion of `r round(Prop_longer * 100, 0)`%.

**SLI-estimated** performed similarly to SLI-isometry across the Longer, Fatter, and Proportional categories (@fig-eval).

Results for an additional **Bigger** scenario — species that grow uniformly larger with warming, the mirror image of Proportional and the simulated analogue of empirically observed Inverse Bergmann's responses — are presented in the Supplementary Information.

```{r}
#| label: fig-eval
#| fig-cap: "Evaluation of method accuracy across shape-response categories. (a) Proportion of species in each category for which the method correctly recovered the direction of temperature's effect. An optimal method scores 1.0 for Longer and Fatter categories. (b) For species simulated as Proportional (temperature reduces appendage and mass equally, so there is no true shape signal), the proportion classified as becoming relatively longer or fatter. The dashed line marks the true proportion of species that happened to become relatively fatter due to random variation in the simulation. An ideal method gives approximately equal classifications."
#| fig-width: 10
#| fig-height: 6.5
#| fig-pos: 'H'

Prop_correct_p <- Eval_tbl %>%
  mutate(Model = str_replace_all(Model, x_labs)) %>%
  ggplot(aes(x = Model, y = Prop_correct, fill = Temp_eff)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = NULL, y = "Proportion correctly classified", fill = "True effect") +
  theme(axis.text.x = element_text(size = 9, vjust = .58, angle = 55),
        legend.position = "top")

Pos_neg_p <- Proportional_methods_eval %>%
  ggplot(aes(x = Model, y = Proportion, fill = fct_rev(Direction))) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = Prop_fatter, linetype = "dashed") +
  labs(x = NULL, y = "Proportion") +
  theme(axis.text.x = element_text(size = 9, vjust = .58, angle = 55),
        legend.position = "top") +
  scale_fill_discrete(name = NULL, labels = c("Longer", "Fatter"))

ggarrange(Prop_correct_p, Pos_neg_p, labels = "auto")
```

## Empirical case studies

```{r}
#| label: rank-consistency-empirical
#| include: false

Rank_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_parms.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_parms.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_parms.csv",  show_col_types = FALSE)
) %>%
  distinct(species, Study, Direction, rank_consistent) %>%
  filter(Direction %in% c("Bergmann's", "Inverse Bergmann's"))

Rank_summary_dir <- Rank_all %>%
  group_by(Direction) %>%
  summarize(n = n(), n_consistent = sum(rank_consistent),
            pct = round(100 * mean(rank_consistent)), .groups = "drop")

n_rank_total     <- nrow(Rank_all)
pct_rank_overall <- round(100 * mean(Rank_all$rank_consistent))

pull_rank_stat <- function(dir, col) {
  Rank_summary_dir %>% filter(Direction == dir) %>% pull({{ col }})
}

pct_rank_bergmanns     <- pull_rank_stat("Bergmann's", pct)
n_rank_bergmanns       <- pull_rank_stat("Bergmann's", n)
pct_rank_inv_bergmanns <- pull_rank_stat("Inverse Bergmann's", pct)
n_rank_inv_bergmanns   <- pull_rank_stat("Inverse Bergmann's", n)
```

The rank-order pattern from simulations was reproduced in nightjars: ratio methods consistently produced the most positive estimates and multiple regression / OLS residuals the most negative, with SLI falling between these extremes. This pattern extended across all three datasets: `r pct_rank_overall`% of the `r n_rank_total` species with a directional (Bergmann's or Inverse Bergmann's) classification showed the rank order predicted by the simulations (`r pct_rank_bergmanns`% of `r n_rank_bergmanns` Bergmann's species and `r pct_rank_inv_bergmanns`% of `r n_rank_inv_bergmanns` Inverse Bergmann's species; @fig-empirical). The strong negative correlation between the wing-to-mass ratio and body mass ($r < -0.9$ in all three nightjar species) explains the ratio's upward bias: as mass declines with increasing temperature (Bergmann's rule), the ratio's denominator shrinks, producing artifactually large increases irrespective of true shape change.

```{r}
#| label: fig-empirical
#| fig-cap: "Estimated effect of the environmental gradient on relative wing length across three empirical avian datasets (Nightjars — red labels; North American migrants  — blue labels; Atlantic Forest birds — green labels), grouped by direction of body size change. Only species with a significant ($p < 0.05$) mass or wing response are shown. Points are mean parameter estimates; error bars are 95% confidence intervals. A positive slope indicates birds become relatively longer with increasing temperature (or over time); a negative slope indicates they become relatively fatter. Stars mark species for which the rank order Ratio > SLI > OLS (Bergmann's) or Ratio < SLI < OLS (Inverse Bergmann's) is consistent with simulation predictions."
#| fig-pos: 'H'
#| fig-height: 6

emp_img  <- png::readPNG("Figures/Empirical_combined.png")
emp_grob <- grid::rasterGrob(emp_img, width = grid::unit(1, "npc"),
                              height = grid::unit(1, "npc"))
cowplot::ggdraw() + cowplot::draw_grob(emp_grob)
```

### Correlation between the wing/mass ratio and body mass

```{r}
#| label: ratio-mass-correlation-empirical
#| include: false

Ratio_mass_cor_all <- bind_rows(
  read_csv("Derived/Csv/Nightjar_ratio_mass_cor.csv",  show_col_types = FALSE),
  read_csv("Derived/Csv/Weeks_ratio_mass_cor.csv",     show_col_types = FALSE),
  read_csv("Derived/Csv/Atlantic_ratio_mass_cor.csv",  show_col_types = FALSE)
) %>%
  dplyr::select(Study, species, Metric, n, r, p_value)

Ratio_mass_summary_tbl <- Ratio_mass_cor_all %>%
  summarize(N_species = n(), Mean_r = mean(r), Median_r = median(r),
            Min_r = min(r), Max_r = max(r), .by = c(Study, Metric)) %>%
  bind_rows(
    Ratio_mass_cor_all %>%
      summarize(Study = "All studies combined", N_species = n(),
                Mean_r = mean(r), Median_r = median(r),
                Min_r = min(r), Max_r = max(r), .by = Metric)
  ) %>%
  mutate(across(c(Mean_r, Median_r, Min_r, Max_r), \(x) round(x, 2)),
         Metric = factor(Metric, levels = c("Ratio", "Ratio2"))) %>%
  arrange(Study, Metric)

Ratio_mass_species_tbl <- Ratio_mass_cor_all %>%
  arrange(Study, species, Metric) %>%
  mutate(r = round(r, 2),
         p_value = ifelse(p_value < 0.001, "< 0.001", as.character(round(p_value, 3))))

# Simple ratio (A/S) -- the metric the surrounding prose already refers to
Ratio_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio")
n_rmc_total        <- nrow(Ratio_cor_only)
mean_rmc_all       <- round(mean(Ratio_cor_only$r), 2)
range_rmc_all      <- paste0(round(min(Ratio_cor_only$r), 2), " to ",
                             round(max(Ratio_cor_only$r), 2))
pct_rmc_strong_neg <- round(100 * mean(Ratio_cor_only$r < -0.7))

# Squared ratio (A^2/S) -- quantifies how much squaring the appendage term
# weakens (but does not eliminate) the correlation with mass.
Ratio2_cor_only     <- Ratio_mass_cor_all %>% filter(Metric == "Ratio2")
mean_rmc2_all       <- round(mean(Ratio2_cor_only$r), 2)
range_rmc2_all      <- paste0(round(min(Ratio2_cor_only$r), 2), " to ",
                              round(max(Ratio2_cor_only$r), 2))
pct_rmc2_strong_neg <- round(100 * mean(Ratio2_cor_only$r < -0.7))
```

The discussion below attributes the ratio methods' upward bias to a mechanical correlation between the ratio metric and body mass. To confirm this holds beyond the nightjar case study, I calculated the Pearson correlation between the wing/mass ratio metric ($\log(A/S)$, following the log-ratio convention used throughout) and body mass separately for every species in all three empirical datasets (@tbl-ratio-mass-summary). Across `r n_rmc_total` species, the ratio metric was consistently and strongly negatively correlated with mass (mean $r$ = `r mean_rmc_all`, range `r range_rmc_all`); `r pct_rmc_strong_neg`% of species had a correlation more negative than $-0.7$. This pattern held regardless of taxon, gradient type (spatial or temporal), or geographic region, confirming that the ratio's dependence on body size is a general mechanical property of the metric rather than a quirk of any one dataset. I calculated the same correlation for the squared-ratio metric ($\log(A^2/S)$); consistent with squaring the appendage term relative to a linear mass denominator, this correlation was consistently weaker in magnitude but still strongly negative (mean $r$ = `r mean_rmc2_all`, range `r range_rmc2_all`; `r pct_rmc2_strong_neg`% of species below $-0.7$; @tbl-ratio-mass-summary).

```{r}
#| label: tbl-ratio-mass-summary
#| tbl-cap: "Summary of the per-species Pearson correlation between the wing/mass ratio metric ($\\log(A/S)$) and its squared counterpart ($\\log(A^2/S)$) and body mass, by empirical dataset."

Ratio_mass_summary_tbl %>%
  filter(Study != "All studies combined") %>%
  mutate(Metric = recode(Metric, "Ratio" = "Wing / Mass", "Ratio2" = "Wing² / Mass"),
         Range = paste(Min_r, "to", Max_r)) %>%
  dplyr::select(Study, Metric, N_species, Median_r, Range) %>%
  knitr::kable(
    col.names = c("Study", "Metric", "N species", "Median", "Range")
  )
```

# Discussion

The method used to quantify relative appendage length had a large and predictable influence on inferences about Allen's rule. Across simulations spanning a realistic range of allometric relationships and temperature responses, ratio-based methods consistently overestimated changes in relative appendage length, whereas OLS-based approaches consistently underestimated them. The Standardized Length Index (SLI) provided the most accurate estimates overall, and the same rank-order pattern was reproduced across three independent empirical datasets. Together, these results suggest that many published estimates of climate-driven shape change may depend as much on analytical choice as on biological signal. More broadly, they highlight that quantifying body shape is fundamentally an allometric problem, and that methods developed to account for allometric scaling offer a more reliable framework for testing ecogeographical rules

## Why ratio approaches overestimate temperature effects

Ratio-based approaches, whether $A/S$ or $A^2 / S$, consistently overestimated $\hat{\beta}_T$ throughout my simulations, often by a large margin. Although ratio variables are commonly interpreted as measures of relative appendage length, they frequently remain strongly correlated with body size itself [@greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996]. Across all three empirical datasets, the wing-to-mass ratio was highly correlated with body mass, indicating that it retained much of the same information as its denominator. Consequently, environmental changes in body size were propagated directly into the response variable, making it difficult to distinguish changes in body shape from changes in overall size.

This behaviour explains the systematic bias observed in the simulations. Under Bergmann's rule, body mass declined with increasing temperature, causing the wing-to-mass ratio to increase even when wing length changed only in proportion to body size. The ratio therefore exaggerated evidence for Allen's rule by conflating shrinking body size with increasing relative appendage length. The converse occurred under inverse Bergmann's rule, where increasing body mass caused the ratio to underestimate changes in relative appendage length ( Supplementary Figure X). In both cases, the ratio inherited part of the environmental signal acting on body size, rather than isolating shape alone.

The comparatively good performance of the squared-ratio metric $A^2/S$ is also informative. Unlike the simple ratio, it was motivated by geometric arguments relating surface area to volume and therefore partially accounts for allometric scaling [@dubinerWidespreadRecentChanges2022; @nwaoguTemperatureAridityDetermine2018]. More generally, ratio methods differ in the allometric relationship they assume between appendage length and body size. However, these assumptions are fixed by the algebraic form of the index rather than being specified as a biological hypothesis or estimated from the data. In contrast, SLI explicitly standardizes appendage length to a common body size using either the theoretical expectation under isometry (SLI-isometry) or the observed allometric relationship (SLI-estimated). The superior performance of both SLI approaches therefore suggests that accurately accounting for allometric scaling, rather than simply transforming the response into a ratio, is more effective for isolating changes in relative appendage length.

## Why classical regression underestimates temperature effects

Causal modeling helps make explicit which paths are being estimated by the regression model, and which are blocked when we control for body size (i.e., mass in our simulation). The model $Appendage \sim \beta_T * \mathit{Temperature\ increase} + \beta_M * Mass$ implies a causal relationship of both predictors (temperature increase and body size) on appendage length. If we assume Bergmann's rule then there is also a causal relationship between temperature and mass, which corresponds to panel a) in @fig-dag.

This directed acyclic graph (DAG) implies that the total effect of temperature on appendage length consists of a direct path $Temperature \rightarrow Appendage$ and an indirect path that operates via changes in body size, $Temperature \rightarrow Mass \rightarrow Appendage$. The DAG also exposes a key assumption of the multiple regression approach: that $Mass \rightarrow Appendage$. When this assumption is true, a multiple regression model conditioning on mass removes the indirect path and returns an unbiased estimate of $\beta_T$. However, alternative models where $Appendage \rightarrow Mass$ (@fig-dag, panel b) or some unknown variable causes correlation in mass and appendage (@fig-dag, panel c) highlight other plausible causal structures (also see Supplementary information section X). The data simulation protocol of this study most resembles @fig-dag, panel c, where mass and appendage are correlated but not causally related. Indeed, under this model mass is actually an endogenous collider, and conditioning on mass produces a spurious relationship between temperature and appendage.

```{r}
#| label: fig-dag
#| fig-cap: "Three causal models relating temperature, body mass, and appendage length, differing in the assumed relationship between mass and appendage length. In each, temperature has a direct effect on both body size (Bergmann's rule) and appendage length (Allen's rule). (a) The model implied by multiple regression: mass has a direct causal effect on appendage length via allometric scaling, so conditioning on mass appropriately isolates the direct effect of temperature on appendage length. (b) An alternative model in which appendage length instead causally precedes mass; here mass is a collider, and conditioning on it induces a spurious association between temperature and appendage length. (c) A model reflecting genuine uncertainty about the direction of the mass-appendage relationship, represented as an unresolved (bidirected) association; under this model it is unclear whether conditioning on mass isolates the direct effect of temperature or instead induces collider bias."
#| fig-pos: 'H'
#| fig-width: 10
#| fig-height: 4

# Plot function -------------------------------------------------------------
# dagitty's y-axis runs top-to-bottom (positive = down), while ggplot's runs
# bottom-to-top; flip y here so the pos = "" coordinates above still read the
# same way (negative = higher on the page).

library(ggraph)

plot_dag <- function(dag, title, panel_label, node_labels = c(Age_sex = "Age / sex"), curvature = 0.3,
                      xlim = c(-0.45, 0.45), ylim = c(-0.42, 0.16)) {

  df <- tidy_dagitty(dag)$data
  df$y <- -df$y
  df$yend <- -df$yend
  df$label <- ifelse(df$name %in% names(node_labels), node_labels[df$name], df$name)
  df$label_end <- ifelse(df$to %in% names(node_labels), node_labels[df$to], df$to)

  # Fixed xlim/ylim (rather than automatic per-panel scale expansion) so that
  # node labels land at identical positions across panels regardless of the
  # extra vertical room the curved bidirected edge in model_unk needs.
  ggplot(df, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_dag_edges(
      aes(start_cap = label_rect(label, padding = margin(3, 3, 3, 3, "mm")),
          end_cap = label_rect(label_end, padding = margin(3, 3, 3, 3, "mm"))),
      edge_colour = "grey55", edge_width = 0.5, curvature = curvature
    ) +
    geom_dag_text(aes(label = label), colour = "black", size = 4.3, fontface = "plain") +
    ggtitle(paste0(panel_label, ") ", title)) +
    coord_equal(xlim = xlim, ylim = ylim, clip = "off", expand = FALSE) +
    theme_dag() +
    theme(
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
      plot.margin = margin(5, 8, 5, 8)
    )
}

# Model implied by multiple regression
model_mr <- dagitty('
dag {

  bb = "-.5,-.5,.5,.5"

  Temperature  [pos="0, 0.28"]
  Mass  [exposure, pos="-0.20, 0"]
  Appendage   [pos="0.20,0"]

  Mass -> Appendage
  Temperature -> Mass
  Temperature -> Appendage
}
')

# Model in which mass is a collider
model_collider <- dagitty('
dag {

  bb = "-.5,-.5,.5,.5"

  Temperature  [pos="0, 0.28"]
  Mass  [exposure, pos="-0.20, 0"]
  Appendage   [pos="0.20,0"]

  Mass <- Appendage
  Temperature -> Mass
  Temperature -> Appendage
}
')

# Model with unresolved directionality between mass and appendage
model_unk <- dagitty('
dag {

  bb = "-.5,-.5,.5,.5"

  Temperature  [pos="0, 0.28"]
  Mass  [exposure, pos="-0.20, 0"]
  Appendage   [pos="0.20,0"]

  Mass <-> Appendage
  Temperature -> Mass
  Temperature -> Appendage
}
')

ggarrange(
  plot_dag(model_mr,       "Implied by multiple regression", "a"),
  plot_dag(model_collider, "Mass as a collider",              "b"),
  plot_dag(model_unk,      "Unknown directionality",          "c", curvature = -0.3),
  nrow = 1, labels = NULL
)
```

The extent of bias produced by conditioning on the collider will depend on the relative strength of the two causal paths. The bias is greatest when the correlation between mass and appendage is low, as the model attributes less variance in absolute appendage length to mass and more to temperature (@fig-neg-bias-mvr). In species decreasing in size proportionally, the stronger the empirical temperature-morphology correlation, the greater the negative bias in $\beta_T$ (@fig-neg-bias-mvr, panel a). In species simulated to get relatively longer, the negative bias is greatest at more negative temperature-appendage correlations, and when there is a relatively large effect of temperature on mass (compared to the the effect of temperature on appendage; low 'Strengths' in @fig-neg-bias-mvr, panel b).

```{r}
#| label: fig-neg-bias-mvr
#| fig-cap: "Magnitude of the negative bias in multiple regression (mass-as-covariate) as a function of correlation strength. (a) In species simulated to decrease in appendage length proportionally with mass (Proportional), the negative bias grew more extreme as the temperature-morphology correlation became more negative and as the correlation between appendage and mass decreased; allometric scaling (point shape) did not affect the magnitude of bias. (b) In species simulated to become relatively longer, the negative bias was greatest at more negative temperature-appendage correlations. Point colour shows the difference between temperature's effect on wing and its effect on mass (Strength); point size shows the appendage-mass correlation, confirming that the bias is more extreme when this correlation is low."
#| fig-width: 10
#| fig-height: 4.5
#| fig-pos: 'H'

p_bias_prop <- Parms_tbl4 %>%
  filter(Temp_eff == "Proportional" & Model == "Ryding") %>%
  ggplot(aes(x = r_13, y = b_temp_inc, color = factor(r_12))) +
  geom_jitter(aes(shape = Scaling), alpha = .6, size = 3, width = 0.02) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, formula = y ~ poly(x, 2)) +
  labs(x = "Temperature-morphology correlation", y = expression(beta[T]),
       color = "Mass-\nappendage\ncorrelation") +
  guides(shape = "none")

p_bias_longer <- Parms_tbl4 %>%
  filter(Temp_eff == "Longer" & Model == "Ryding") %>%
  ggplot(aes(x = r_13, y = b_temp_inc)) +
  geom_point(aes(color = Strength, size = r_12), alpha = .6) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, formula = y ~ poly(x, 2)) +
  labs(x = "Temperature-appendage correlation", y = expression(beta[T]),
       size = "Mass-\nappendage\ncorrelation")

ggarrange(p_bias_prop, p_bias_longer, labels = "auto")
```

## Why SMA regression performs better

```{r}
#| label: fig-ols-sma
#| fig-cap: "The allometric relationship between log-mass and log-appendage for a simulated hyperallometric species in which temperature increases produce relatively longer appendages. The SMA regression line (blue) is steeper than the OLS line (red). Individual points are coloured by temperature. Sections labelled I and II highlight regions where OLS and SMA assign opposite signs to residuals: small-bodied individuals (Section I) are classified as having relatively short appendages by OLS but relatively long appendages by SMA; large-bodied individuals (Section II) are classified in the opposite direction. As temperature increases select for smaller body size (Bergmann's rule), individuals are displaced toward Section I — exactly where OLS overestimates negative residuals, producing a spurious negative $\\hat{\\beta}_T$."
#| fig-pos: 'H'
#| fig-height: 4.5

Ex_parms_hypo_hyper <- Parms_mat3 %>%
  filter(Scaling != "Isometry" & Temp_eff == "Longer" & r_12 == r_12[1]) %>%
  slice_max(n = 1, order_by = Strength, by = c(Temp_eff, Scaling), with_ties = FALSE)

Ex_df_hypo_hyper <- gen_ex_data(Ex_parms_hypo_hyper) %>%
  filter(Scaling == "Hyperallometry") %>%
  mutate(Temp_inc = Temp_inc / 3)

xy_lims <- Ex_df_hypo_hyper %>%
  summarize(max.append = max(Append_log), min.append = min(Append_log),
            max.mass   = max(Mass_log),   min.mass   = min(Mass_log))

ggplot(Ex_df_hypo_hyper, aes(x = Mass_log, y = Append_log)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ggpmisc::stat_ma_line(method = sma_or_ma, color = "blue") +
  geom_point(alpha = 0.4, aes(color = Temp_inc)) +
  annotate("text", x = 4.23, y = 5.13, label = "I",  size = 6, color = "black") +
  annotate("text", x = 4.52, y = 5.25, label = "II", size = 6, color = "black") +
  scale_color_viridis_c(option = "plasma") +
  lims(y = c(xy_lims$min.append, xy_lims$max.append),
       x = c(xy_lims$min.mass,   xy_lims$max.mass)) +
  labs(x = "Log(mass)", y = "Log(appendage)", color = "Temperature\nincrease")
```

SMA regression produces a steeper slope than OLS whenever $|r| < 1$, because it minimises residuals in both directions rather than only vertically. This steeper slope more accurately estimates the functional relationship between body size and appendage length when both are measured with error [@wartonBivariateLinefittingMethods2006; @smithUseMisuseReduced2009]. The consequence for residual-based shape indices is illustrated in @fig-ols-sma. Where the OLS line is too shallow, it overestimates negative residuals for small-bodied individuals (Section I) and overestimates positive residuals for large-bodied individuals (Section II). Under Bergmann's rule, increasing temperatures push individuals toward smaller body sizes — directly into Section I — so OLS-based indices tend to classify them as having relatively short appendages. The SMA-based SLI, reprojecting individuals along the steeper slope, assigns more positive shape values to these same individuals.

The SLI additionally inherits two useful properties from the SMI [@peigNewPerspectivesEstimating2009]. First, the resulting index retains the units of the original measurement (millimetres), which facilitates biological interpretation. Second, the reference body size $M_0$ enables comparison of individuals across populations and studies: by standardising to the same body size, SLI values from different populations are directly comparable, unlike OLS residuals or ratio values whose scale depends on the particular dataset.

Importantly, these results are conditional on the data-generating process used in the simulations. Traits were drawn from a multivariate normal distribution, which imposes symmetry while ensureing that no method family (e.g., OLS or SMA) is implicitly advantaged by the structure of the simulated data. However, it also means that results are not guaranteed to generalize to alternative generative models. For instance, @rydingMultipleRegressionNot2025 found that multiple regression was unbiased in their simulation study, which targeted a different causal structure. These findings are therefore not contradictory, but instead reflect sensitivity to differing assumptions about the underlying causal system. In empirical morphological datasets, the true directionality of relationships is often ambiguous, which complicates the choice of model structure. In this context, SLI approaches offer a complementary tool that relies on weaker structural assumptions than multiple regression, and may therefore be more robust when causal direction is uncertain.

## Estimating versus assigning SLI scaling exponents

The two SLI parameterisations differ in how the scaling exponent is specified, and this choice becomes most consequential in species with atypical allometric structure. SLI-isometry uses $b_{SLI} = 0.33$, corresponding to the expected isometric scaling of a linear appendage dimension against a volumetric proxy of body size (i.e., mass). This formulation is most appropriate when the goal is to test for deviations from isometric expectations. This sort of null model ($\text{Appendage length} \propto \text{Mass}^{0.33}$) is intuitively appealing for tests of Allen's rule, however there is little consensus regarding the implications of biological scaling for thermoregulation [@tabhLimitedEvidenceThat2025; @santoroAllometryEvaluateAllens2022; @rydingResponseAllometryEvaluate2022].

In contrast, SLI-estimated uses the empirically derived SMA slope, which is useful if the objective is to characterise deviations from the observed scaling relationship rather than from a theoretical null. For example, it removes systematic shifts in appendage length associated with hyperallometric or hypoallometric scaling, leaving residual variation in relative morphology. It may also improve comparability across structured groups such as sex or age classes, where SMA slopes can be estimated separately and incorporated into the standardisation, analogous to the scaled mass index used in body condition studies [@peigParadigmBodyCondition2010]. This approach provides flexibility, but may not be feasible in certain datasets, particularly if appendage–mass relationships are sufficiently weak (e.g., we used $r_{A,S} \geq 0.3$ as a threshold for applying SLI-estimated). In this sense, SLI-estimated standardises trait values relative to a species- or population-specific allometric trend, effectively expressing each individual’s position relative to expected appendage length at a given mass.

Data permitting, the choice between SLI parameterisations depends on whether the inferential target is deviation from a theoretical scaling law (SLI-isometry) or deviation from the observed species-specific scaling structure (SLI-estimated).

## Limitations of the SLI approaches

Despite their advantages, the SLI approaches are not without flaws. Unfortunately, SMA regression is inflexible in its ability to customize models because it can only model the two scaling variables $X$ and $Y$. This inflexibility can make it more challenging to meet SMA regression assumptions relative to OLS assumptions [@wartonBivariateLinefittingMethods2006]. For example, SMA regression assumes homoskedasticity, but missing covariates that influence appendage measurements (e.g., feather wear) cannot be included in SMA regression, then residual patterns may exhibit heteroskedasticity. Similarly, if there is still a non-linear relationship after log-log transforming the linear and volumetric measurement, the SMA assumption of linearity is violated. This non-linearity can be easily accomodated in classical regression [e.g., $\mathit{Body\ size} + \mathit{Body\ size^2}$, @baldwinComplementarityAllensBergmanns2023], but not in SMA regression. In these cases, classical regression approaches with additional covariates may offer improved inference.

## Recommendations for empirical studies

Based on my simulation results and case study, I offer the following practical recommendations for studies of Allen's rule or any investigation of how a linear morphological trait scales with body size across an environmental gradient.

1.  **Avoid ratio-based metrics** as primary indicators of relative appendage length [@nakagawaDivideConquerSize2017]. Ratios are systematically biased when Bergmann's rule co-occurs with Allen's rule, and they are often correlated with body size itself.

2.  **Interpret multiple regression coefficients with care.** The temperature coefficient in a model including body size as a covariate estimates the *direct* effect of temperature on absolute appendage length, not the effect on relative appendage length. This is appropriate for some research questions but misleading for others, and the mediating role of body size should be acknowledged.

3.  **Use SLI-isometry as a default.** It performed well across the widest range of conditions and requires only the mean body size of the species as a reference value. I provide R code for computing SLI in the Supplementary Material.

4.  **Prefer SLI-estimated when the scaling relationship is well-constrained and the allometric slope departs from isometry.** When the appendage–mass correlation is strong ($r_{A,S} \geq 0.4$) and sample sizes are sufficient ($n \geq 100$ per species or group), SLI-estimated captures the empirical allometric relationship and provides a more accurate body-size correction than the fixed isometric exponent.

5.  **Report multiple methods and examine sensitivity.** Given that the choice of method can reverse biological conclusions — as seen in the nightjar case study — I encourage authors to report results from at least two methods (e.g., SLI-isometry and multiple regression) and discuss how conclusions depend on methodological choice.

6.  **Report the correlation between appendage length and body size** ($r_{A,S}$ or $r_{\log A, \log S}$). This correlation determines the divergence between OLS and SMA slopes and is therefore directly informative about the likely magnitude of bias in classical regression approaches.

# Conclusions

The method used to quantify relative appendage length has large and predictable effects on biological conclusions in studies of Allen's rule. Classical approaches — ratios, residual regression, and multiple regression — are biased in opposite directions for reasons that are well-understood from the perspective of statistical theory: ratios are inflated by co-occurring Bergmann's rule effects; OLS-based residuals and multiple regression underestimate the true allometric slope and produce attenuated or reversed temperature effects. The Standardized Length Index, grounded in SMA regression, avoids these biases by distributing error to both variables and standardising each individual's appendage length to a common reference body size. Its performance in simulations spanning a realistic range of biological conditions, and the consistency of the rank-order pattern in empirical nightjar data, support its adoption as a default approach for studies examining body shape across environmental gradients. More broadly, the issues identified here apply to any study that uses OLS-based methods to quantify how a morphological trait scales relative to body size — a problem that extends well beyond Allen's rule to questions in comparative biomechanics, ecomorphology, and evolutionary biology.

# Data Availability

All analyses can be reproduced using the code available on [Github](https://github.com/LezzGitIt/SLI-allens-rule.git). The nightjar morphological data are available from @skinnerEnvironmentalGeographicConditions2025; Atlantic birds are available from @rodriguesATLANTICBIRDTRAITS2019; the Chicago field museum dataset is available from @weeksSharedMorphologicalConsequences2020.

# Acknowledgements

I thank Andy Green for helpful comments on the development of the Standardized Length Index and its relationship to the Standardized Mass Index, and David Warton for clarifying the assumptions and properties of SMA regression.

# EXTRAS - delete

These classical approaches share a critical assumption: that body size is measured without error. This is the operating assumption of ordinary least squares regression and is retained in most implementations of multiple regression, maximum likelihood, and Bayesian models unless the model is explicitly formulated otherwise. In practice, this assumption is nearly always violated. Mass — the most commonly used body size proxy — reflects structural body size but also transient biological variation such as gut fill, hydration state, and reproductive condition. Instrument and observer error introduce additional noise. When predictor variables are measured with error, OLS slope estimates are attenuated toward zero — a bias that grows as the correlation between the predictor and response weakens [@wartonBivariateLinefittingMethods2006; @greenMassLengthResiduals2001]. By ignoring this attenuation, classical regression approaches underestimate the allometric relationship between mass and appendage length, which propagates as bias into downstream estimates of temperature's effect on shape.

{{< pagebreak >}}

# References
