---
title: "Detecting shape shifting across environmental gradients -- simulations show promise in a new metric, the standardized length index"
number-sections: true
# If you do not want code to show... echo: false
execute: 
 echo: false
 #include: false
 message: false
 warning: false 
format: docx
include-in-header: 
      text: |
        \usepackage{float}
editor: visual
bibliography: ../Supp_files/EWPW.bib
link-citations: true
csl: ../Supp_files/methods-in-ecology-and-evolution.csl
knitr:
  opts_chunk: 
    comment: "#>"
    collapse: true
editor_options: 
  chunk_output_type: console
---

{{< pagebreak >}}

# Abstract

Due to its important implications for wildlife responses to climate change, studies of Allen's rule (i.e., that animals have relatively longer appendages in warmer climates to facilitate thermoregulation) have increased in recent decades. While there is general agreement that warm-blooded animals become longer in warmer climates, the methods used to quantify changes in relative appendage length in response to temperature gradients have varied widely. I conduct a literature review of studies examining Allen's rule since 2021 and find that, in addition to modeling absolute (i.e., not relative) appendage length, authors have taken three main approaches: (1) modeling simple ratios (e.g., appendage:mass) as a function of temperature, (2) modeling appendage length as a function of body size and temperature, and (3) conducting a two-step regression by (3a) modeling appendage length as a function of mass, and then (3b) modeling the residuals as a function of temperature.

Here, I test whether a new metric of relative appendage length, called the Standardized Length Index (SLI), is able to accurately capture changes in *relative* appendage length due to a simulated temperature increase. SLI leverages standardized major axis (SMA) regression, which has been advocated for in applications involving biological scaling. I compare SLI to the three approaches commonly used in the literature using simulated morphological data from 180 hypothetical species that vary in 1) their response to increasing temperatures, and 2) the strength of appendage length–mass covariation. The SLI approach accurately capture changes in *relative* appendage length in the vast majority of simulations, although it is positively biased in specific situations. I find that ratio approaches (approach 1) overestimate the increase in relative appendage length, while 'multivariate regression' (approach 2) and a two-step residual regression (approach 3) underestimate relative appendage length. Importantly, the 'multivariate regression' approach suffers from collider bias in several of the hypothetical species, a previously overlooked issue in studies of Allen's rule. I find qualitatively similar results in a case study examining relative appendage length across temperature gradients in three nightjar species from North America and Europe.

These results show that the methodological approach leveraged to examine changes in relative appendage length can have drastic effects on results. Approaches 1 - 3 make the assumption that error exists only in $Y$ (appendage length), which is biologically implausible in studies of allometric scaling, and particularly unrealistic when mass is used as an estimate of structural body size ($X$). Thus, approaches that allow for measurement error in both $X$ and $Y$ outperform the approaches that are regularly used in the literature. I highlight that alternative approaches where measurement error in $X$ and $Y$ is directly estimated from repeat morphological measurements are more flexible and will perform better in a greater variety of circumstances. However, the SLI approach presented here offer easily-implementable alternatives, and would be a great improvement over the methods regularly used to assess shape-shifting in wildlife at present.

# Introduction

Wildlife appear to be changing their size and shape in response to climate change. Temporal applications of Bergmann’s rule predict that body size will decrease with rising temperatures, while Allen’s rule posits that appendage length will increase. Because Allen’s rule concerns organismal shape, it is often interpreted in terms of appendage length relative to body size. This framing is important as absolute body size and appendage length are usually positively correlated, so isolating shape requires methods that account for this covariance.

In a concurrent literature review (Skinner, in prep), I identify three predominant approaches used to quantify relative appendage length in the context of Allen’s rule, where $\beta_T$ is the estimand of interest in all three cases.:

1)  Ratio approach: Using the ratio of appendage length to body size as the response variable\
    $\frac{Appendage}{\mathit{Body\ size}} \sim \beta_T * \mathit{Temperature\ increase}$

2)  'Two-step residual regression' approach: Using residuals from an ordinary least squares (or similar) regression of appendage on body size\
    Model 1: $Appendage \sim \beta_S * \mathit{Body\ size}$\
    Model 2: $Residuals_{OLS} \sim \beta_T * \mathit{Temperature\ increase}$

3)  'Multivariate regression' approach: Including body size as a covariate in models predicting appendage length\
    $Appendage \sim \beta_S * \mathit{Body\ size} + \beta_T * \mathit{Temperature\ increase}$

All three approaches attempt to control for body size to isolate changes in relative appendage length. These approaches share another key assumption – that the independent variable $X$ (body size in this case) is measured without error. This assumption is often untenable, particularly in field-based studies of morphology where both traits are prone to measurement error. Hereafter, I term any statistical approach making this assumption as 'classical regression'. Importantly, more sophisticated regression techniques such as generalized least squares, maximum likelihood, or Bayesian approaches make the same assumption regarding the error distribution between $X$ and $Y$ (appendage length in this case), unless models are explicitly formulated to the contrary.

Mass was by far the most commonly used proxy of body size in recent studies of Allen's rule (Skinner in prep), thus it is important to briefly mention an additional source of 'error'. Transient fluctuation 'error' is real biological variation in our *estimate* of body size that do not reflect changes in structural body size \[this has also been called 'biological error', CITE PELABON\]. Thus, when using mass as an estimate of body size, this could be variation due to gut fill, defecation, reproductive or hydration state, etc. that do not reflect relevant changes in structural body size, yet will influence mass measurements.

Each of these classical approaches has known limitations. Ratio-based metrics, despite widespread continued use [e.g., @jirinecMorphologicalConsequencesClimate2021; @dubinerWidespreadRecentChanges2022], are often correlated with body size and thus have limited utility as an index of shape [@greenMassLengthResiduals2001; @jakobEstimatingFitnessComparison1996, CITE Ryding 2025]. The 'two-step residual regression' and the 'multivariate regression' approaches are mathematically similar when there is no collinearity between predictor variables (e.g., temperature and body size), but the former becomes increasingly biased as collinearity increases [@freckletonMisuseResidualsEcology2002]. Importantly, both of these approaches will underestimate the true slope of the functional relationship between $X$ and $Y$ (i.e., attenuation towards 0) when there is unmodeled measurement or transient fluctuation error in $X$.

Despite the near universal application of classical regression approaches in studies of Allen's rule (Skinner in prep), alternative approaches that allow for measurement error in $X$, as well as $Y$, may be more appropriate. For example, standardized major axis (SMA) regression can more accurately estimate the true functional relationship between two variables as it acknowledges measurement error in both $X$ and $Y$ [@greenMassLengthResiduals2001; @wartonBivariateLinefittingMethods2006; @peigNewPerspectivesEstimating2009; @peigParadigmBodyCondition2010]. The key difference between SMA and classical regression is how they minimize residuals to generate a line of best fit. OLS regression minimizes the residuals on the $Y$ axis, whereas SMA regression minimizes the residuals in both $X$ and $Y$ directions (it is also called 'orthogonal regression'). This makes sense for situations where we don't think that $X$ -\> $Y$ (we wouldn't say that mass causes appendage length), but instead just want to understand the relationship between $X$ and $Y$ (CITE Smith, 2009). The statistical distinctions between OLS and SMA have received attention in several fields including statistics (CITE Smith 2009), evolutionary biology (CITE Hansen, Pelabon), ecology, genetics (CITE Ponzi), and even astrophysics (1996 paper), and thus will not be discussed further here.

Methodologically, the quantification of body condition is challenging in ways that parallel the difficulties of measuring relative appendage length. Both involve attempts to assess a biological trait (i.e., energy reserves or appendage length) *relative* to an organism’s structural body size. In the case of body condition, direct measurements of energy reserves are rarely available in wild animals, so researchers have relied on statistical methods to control for structural size. Common strategies include ratios, residual-based methods, and multivariate regression, all of which face similar limitations to those outlined above for relative appendage length.

Peig and Green recognized these challenges and developed the **Standardized Mass Index (SMI)**, now widely used as a size-corrected measure of body condition [@peigNewPerspectivesEstimating2009; @peigParadigmBodyCondition2010]. Their approach uses SMA regression to estimate the scaling exponent linking structural body size and mass, then adjusts each individual's mass to a common reference body size. This process effectively removes allometric effects, allowing for comparisons of mass across individuals with different structural sizes. The SMI has since become a foundational tool in ecological research, cited in thousands of studies.

Inspired by the logic of Peig and Green's SMI [@peigNewPerspectivesEstimating2009; @peigParadigmBodyCondition2010], I develop a related metric tailored to body shape rather than body condition: the **Standardized Length Index (SLI)**. Like the SMI, the SLI uses SMA regression to empirically derive an allometric scaling exponent (or can be set based on biological principles) and standardizes each individual's appendage length to a common body size within a group (e.g., species or population). In doing so, it isolates departures from expected scaling relationships and provides a biologically interpretable index of shape. The SLI is similar to two-step residual regression, it offers several key advantages. The SLI 1) maintains the units of the original variable (e.g., millimeters) which facilitates interpretation, 2) is grounded in allometric scaling theory (CITE Huxley), modeling power-law relationships rather than assuming additive error structures, and 3) is comparable across populations within the same species [@peigNewPerspectivesEstimating2009; @peigParadigmBodyCondition2010].

In this paper, I use simulations to evaluate how well the SLI captures changes in relative appendage length due to temperature increases, comparing its performance to the three classical regression approaches described earlier. These comparisons are made across a range of realistic conditions, including varying degrees of measurement and transient biological error, trait covariance, and allometric scaling relationships. While several authors have modeled *absolute* appendage length as a function of temperature (Skinner, in prep), I focus exclusively on approaches that attempt to isolate *relative* shape. Although Allen’s rule provides the motivating context, the insights from this study extend to many questions rooted in biological scaling theory that seek to quantify a trait *relative to body size*.

```{r}
#| label: Libraries
#| message: false

library(tidyverse)
library(readxl)
library(janitor)
library(png)
library(grid)
library(gridExtra)
library(gt)
library(ggpubr)
library(cowplot)
library(smatr)
library(MASS) 
library(ggpmisc) # Plots MA or SMA lines of best fit
ggplot2::theme_set(theme_cowplot())
#library(scales) 
#library(conflicted)
#conflicts_prefer(dplyr::select)
#conflicts_prefer(dplyr::filter)
```

# Methods

To evaluate the three classical regression approaches and the newly proposed SLI, I simulated increasing temperatures (in line with climate change), along with morphological data for 180 hypothetical species. Each species was simulated with \~1,700 individuals, and each method was evaluated by its ability to recover known (simulated) morphological responses within each species. These species varied in their morphological response to temperature, where some species got relatively longer while others got relatively fatter with increasing temperatures, as well as the strength of body size–appendage length covariation. The three classical regression approaches tested were ratios, two-step residual regression, and multivariate regression, which were compared against two parameterizations of the Standardized Length Index (SLI). Finally, I applied all methods to empirical data from three migratory nightjar species.

## Data simulation

I simulated three log-transformed variables: appendage length, body size (estimated from mass), and temperature increase. These were jointly drawn from a multivariate normal (MVN) distribution:

$$
\begin{bmatrix}
\log(\text{Appendage}) \\
\log(\text{Body size}) \\
\log(\text{Temperature increase})
\end{bmatrix}
\sim \mathrm{MVN}(\boldsymbol{\mu}, \Sigma)
$$ {#eq-mvn}

where $\boldsymbol{\mu} = [3,\ 3,\ 3]^\top$ is the mean vector, and $\Sigma$ is a $3 \times 3$ covariance matrix constructed to yield biologically plausible correlations among variables. This strategy avoids biasing results toward any method. For example, simulating from a directional model, e.g.,

$$
\log(\text{Appendage}) \sim \alpha + \beta_S * \log(\text{Body size}) + \beta_T * \log(\text{Temperature increase}) + \epsilon
$$

would advantage methods that explicitly include mass as a covariate. Likewise, simulating a relationship between temperature and residuals would favor residual-based approaches. Instead, all methods start from the same joint distribution, and none are privileged by design.

```{r}
#| label: Source-key-fns
#| echo: true

# Read in the most important functions
source("Scripts/Key_allometry_fns.R")
```

```{r}
#| label: Data-generating-function
#| echo: true

# Examine the data generating function
gen_data
```

## Variance-covariance matrix

Specification of the $3 \times 3$ covariance matrix determines the variance within variables and the relationships between variables. $\Sigma$ is defined as

$$
\Sigma = 
\begin{bmatrix}
\sigma_1^2 & \rho_{12} \sigma_1 \sigma_2 & \rho_{13} \sigma_1 \sigma_3 \\
\rho_{12} \sigma_1 \sigma_2 & \sigma_2^2 & \rho_{23} \sigma_2 \sigma_3 \\
\rho_{13} \sigma_1 \sigma_3 & \rho_{23} \sigma_2 \sigma_3 & \sigma_3^2
\end{bmatrix}
$$

where $\sigma_1$, $\sigma_2$, and $\sigma_3$ are the standard deviations of $\log(\text{Appendage})$, $\log(\text{Body size})$, and $\log(\text{Temperature increase})$, and the correlation coefficients $\rho_{12}$, $\rho_{13}$, and $\rho_{23}$ control the pairwise associations among these traits. Consistent with Bergmann's rule, all temperature-morphology correlations are \< 0 (except in species exhibiting inverse allometry, see @sec-selecting-parms), meaning that temperature increases $\rightarrow$ smaller body size. Some species were simulated to get relatively longer (i.e., the decrease in body size is greater than the decrease in appendage length), some to get fatter, while others were simulated to decrease in appendage length and body size proportionally (i.e., body shape doesn't change).

```{r}
#| label: Function-var-cov
#| echo: true

# Examine the function used to generate the variance-covariance matrix. This function is used in the gen_data() function above
gen_3var_cov

# See an example cov matrix using default parameters 
ex_cov_mat <- gen_3var_cov(b_sma_12 = 0.33) 
round(ex_cov_mat, 2)
```

## Biological realism

This simulation implies a symmetric, undirected dependence structure — no variable is treated as a causal parent or response. While this does not encode the directional hypothesis of Allen’s rule (i.e., temperature influencing relative appendage length), I have taken several steps to ensure biological realism: 1) simulating data from empirically-derived correlations and allometric slopes, 2) adding morphological noise, which is impossible to avoid, 3) conducting sensitivity analyses varying correlations and allometric slopes, and 4) conducting an empirical validation in a case study of nightjar species.

### Using empirically-derived parameter combinations {#sec-selecting-parms}

```{r}
#| label: Set-b_sma

## Select whether you'd like to constrain the b_sma values or not. This is a step to ensure that setting b_sma values is not biasing our findings -- constraining the SMA values will fix the OLS slopes (see the subsection Sensitivity analyses). 
b_sma <- c(.22, .33, .44) # NULL  
```

```{r}
#| label: Define-parameter-matrix

## In the parameter matrix we define the relationships we want to simulate — you can think of each row as a hypothetical species that responds differently to increasing temperatures, and in the strength of appendage length–mass covariation.

# Create combinations of parameters to run through the extract_coefs() function
Shape <- c("Wingyier", "Proportional", "Fatter")
Shape <- setNames(Shape, Shape) 

r_12 <- c(.15, .3, .45)
r_13 <- c(0, -.1, -.25, -.4) # Wing temp
r_23 <- r_13
Parms_mat <- expand_grid(
  b_sma_12 = b_sma,
  r_12 = r_12, # Wing mass
  r_13 = r_13, # Wing temp
  r_23 = r_23 # Mass temp
  # Effect of temperature on wingyness
) %>% mutate(Temp_eff = case_when( 
  r_13 > r_23 ~ Shape[1], 
  r_13 == r_23 ~ Shape[2], 
  r_13 < r_23 ~ Shape[3]), 
  Strength = abs(r_13 - r_23))
if(!is.null(b_sma)){
  Parms_mat <- Parms_mat %>% mutate(Scaling = case_when(
    b_sma_12 < 0.33 ~ "Hypoallometry", 
    b_sma_12 > 0.33 ~ "Hyperallometry",
    near(b_sma_12, 0.33) ~ "Isometry"
  ))
}
Parms_mat
```

```{r}
#| label: Neg-allometry

Parms_neg <- expand_grid(
  b_sma_12 = -0.22,
  r_12 = c(-.2, -.3, -.4),
  r_13 = c(0, .1, .2, .4),
  r_23 = c(-.1, -.2, -.4)
  # Effect of temperature on wingyness
) %>% mutate(Temp_eff = "Much wingyier", 
             Strength = abs(r_13 - r_23), 
             Scaling = "Negative") 
```

```{r}
#| label: Combine-format-parms-mat

# Combine & format tibble
Parms_mat2 <- bind_rows(Parms_mat, Parms_neg) %>%
  mutate(Temp_eff = factor(Temp_eff, levels = c("Fatter",  "Proportional", "Wingyier", "Much wingyier")), 
           Scaling = factor(Scaling, levels = c("Hypoallometry", "Isometry", "Hyperallometry", "Negative"))) %>% 
    relocate(Scaling, .before = b_sma_12)
if(is.null(b_sma)){
  Parms_mat2 <- Parms_mat2 %>% dplyr::select(-b_sma_12)
}
```

Data were simulated from correlation structures encompassing a broad but realistic range of appendage length–mass values \[`r min(r_12)` \< $\rho_{12}$ \< `r max(r_12)`, @greenMassLengthResiduals2001; @goslerFieldDeterminationBody1998\], and temperature-morphology values \[`r min(r_13)` \< $\rho_{13}$ and $\rho_{23}$ \< `r max(r_13)`, consistent with literature reviews in birds, @ashtonPatternsWithinSpeciesBody2002, and mammals CITE ashton 2000\]. I also deterministic set the intraspecific SMA slope to be either `r paste0(b_sma)`, reflecting biologically plausible scaling coefficients [e.g. @youngfleshAbioticConditionsShape2022]. To challenge the different methods with adversarial data sets [@lotterhosSimulationTestsMethods2022], hypothetical species with inverse allometric scaling ($\beta_{SMA}$ = -0.22; i.e., where mass and appendage length negatively covary), were simulated as rare but biologically possible [e.g., @weeksSharedMorphologicalConsequences2020].

It is important to note that, from the formula $\beta_{SMA} = \beta_{OLS} / \rho_{12}$, you can see that $\beta_{OLS}$ is the only free parameter in our simulation. Thus mass, wing, and temperature covary in a way that satisfy the specified correlations of $\rho_{12}$, $\rho_{13}$, and $\rho_{23}$, as well as $\beta_{SMA}$, and $\beta_{OLS}$ varies to accommodate the specified $\rho_{12}$ and $\beta_{SMA}$.

### Additional sources of error

As mentioned in the introduction, estimates of body size and appendage length are always influenced by measurement error, and some measurements are subject to additional types of error. In particular, when mass is used as the proxy for body size, transient fluctuation 'error' should not be ignored given that it is likely substantially larger than measurement error (CITE Pelabon). Thus, I include an equivalent amount of measurement error in both $X$ and $Y$ (2% of the standard deviation of $X$ and $Y$), as well as transient fluctuation 'error' in mass equal to 4% of the standard deviation of $X$.

### Sensitivity analyses

Simulations were repeated 1) without deterministically setting $\beta_{SMA}$, 2) with more realistic log-scale trait means as wing values (mm) are usually larger than mass (g), 3) varying the amount of measurement and transient fluctuation error, and 4) with lower correlations between temperature-morphology relationships. General conclusions changed negligibly in all four sensitivity analyses.

NOTE: I also ran using MA regression instead of SMA regression, and found that MA regression does change the results. However, several sources suggest that SMA regression is the approach that makes the most sense (Warton, 2006; Peig & Green, 2009) given that the units are not on the same scale, so I have ignored this for now. I can easily render this document using MA regression & send your way if that would be helpful.

```{r}
#| label: Define-sma-or-ma
#| echo: FALSE

# Set regression method to be SMA or MA
sma_or_ma <- "SMA" 
```

### Case study in nightjars

I applied all methodological approaches to data from three migratory nightjar species (n = 3,272), sampled across a (spatial) temperature gradient (Skinner et al., *accepted* *in the Journal of Biogeography*). These species follow Bergmann's rule and vary in the allometric scaling relationship between wing and mass. European nightjars exhibit isometric scaling ($\beta_{SMA}$ = 0.33), whereas nighthawks and whip-poor-wills exhibit hyperallometric scaling ($\beta_{SMA}$ = 0.44 and $\beta_{SMA}$ = 0.40, respectively). However, Skinner et al. (2025) did not examine how temperature influences relative wing length in nightjars. Thus, my goals were to 1) test method performance using real-world data, 2) evaluate the consistency with simulation results, and 3) assess biological conclusions under each methodological approach.

# Simulate full dataset

```{r}
#| label: Generate-data
#| echo: true
#| cache: true

# Simulate data 
Cols <- Parms_mat2 %>% 
  dplyr::select(starts_with(c("b_", "r_"))) %>% 
  # Define measurement error and transient error
  mutate(meas_error = 0.02,
         transient_error = 0.04)
df_morph_l <- pmap(Cols, gen_data)

# Average number of individuals per hypothetical species
#mean(map_dbl(df_morph_l, nrow))
```

## Simulation Checks

I conducted graphical checks to ensure that simulations behaved as expected across all hypothetical species. First, when simulating wingyier species, SMA intercepts should increase with temperature, and the opposite should be true when simulating fatter birds. Species decreasing in size proportionally should show no systematic pattern in the intercepts and the amount of temperature increase. See Figure 4 in @shingletonAllometryStudyBiological2010 for additional information regarding interpretation of these temperature-body shape plots (i.e., @fig-temp-body-shape-plot). I use 3 example species

```{r}
#| label: Example-parms

# Hold b_sma & the correlation between wing & mass constant so slopes are roughly the same 
Ex_parms <- Parms_mat2 %>% 
  filter(Scaling == "Isometry") %>%
  slice_max(n = 1, order_by = tibble(Strength), 
            by = Temp_eff, with_ties = FALSE) %>%  #desc(r_13)
  dplyr::select(-Strength)
Ex_parms

# Function to generate example data for plotting
gen_ex_data <- function(Parms_mat){
  Cols <- Parms_mat %>% dplyr::select(starts_with(c("b_", "r_")))
  Parms_mat %>%
    mutate(coefs = pmap(Cols, gen_data)) %>%
    unnest(coefs)
}
```

to produce the following plot:

```{r}
#| label: run-sma-and-formatting-functions

# Run SMA model with or without interaction
run_sma_mod <- function(df, interaction = FALSE){
  if(interaction == FALSE){
    sma_mod <- sma(Wing ~ Mass + Temp_bin, data = df, method = sma_or_ma)
  } else{
    sma_mod <-sma(Wing ~ Mass * Temp_bin, data = df, method = sma_or_ma)
  }
  return(sma_mod)
}

# Format 'temperature increase' & 'Temp label' columns
format_sma_parms <- function(sma_mod){
  coef(sma_mod) %>% 
    rownames_to_column("Temp_inc") %>% 
    mutate(
      Temp_inc = str_pad(Temp_inc, side = "left", width = 2, pad = "0"),
      Temp_inc = str_replace(Temp_inc, pattern = "^([0-9])([0-9])$", replacement = "\\1.\\2"),
      Temp_label = paste0(Temp_inc, "°C"), 
      Temp_inc = as.numeric(Temp_inc)) %>%
    tibble() 
}
```

```{r}
#| label: ex-parms-temperature-body-shape 

## Run the SMA models & format output tbl
ex_parms_temp_bs <- map(Shape, \(Sim_shp){
  Ex_df_wingy <- gen_ex_data(Ex_parms) %>% filter(Temp_eff == Sim_shp)
  # Allow slopes to vary or force to remain constant 
  mod_temp_bin <- run_sma_mod(df = Ex_df_wingy, interaction = FALSE) 
  mod_temp_bin_int <- run_sma_mod(df = Ex_df_wingy, interaction = TRUE)
  
  mod_parms <- format_sma_parms(mod_temp_bin) %>% 
    mutate(Slopes_vary = "No")
  mod_parms_int <- format_sma_parms(mod_temp_bin_int) %>% 
    mutate(Slopes_vary = "Yes")
  # Bind tbls together where slopes are held constant or vary
  bind_rows(mod_parms, mod_parms_int)
}) %>% list_rbind(names_to = "Temp_eff")
```

```{r}
#| label: fig-temp-body-shape-plot
#| fig-cap: The allometric scaling slope determines the rate of increase of appendage length with an increase in the body size. The intercept reflects the proportions of the appendage relative to body size, such that when slopes are constant, the group with the higher intercept has proportionally larger wings across all body sizes. As temperature increases, simulated birds get fatter (left panel), wingyier (right panel), or decrease in size proportionally (center panel). This plot confirms that the simulation is working as expected. 

ex_parms_temp_bs %>%
  filter(Slopes_vary == "Yes") %>% # & Temp_eff != "Proportional"
  mutate(x = 0) %>% 
  ggplot() +
  geom_abline(aes(intercept = elevation, slope = slope, color = Temp_inc)) +
  ggrepel::geom_text_repel(aes(x = x, y = elevation, label = Temp_label)) +
  labs(x = "Log mass", y = "Log wing") + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank()) + 
  guides(color = "none") + 
  facet_wrap(~Temp_eff, scales = "free_y")
```

Generally, the simulation has worked, but there are a few data sets where the simulated effect of temperature did not have the intended directional effect on body shape, e.g., points in the 'fatter' category that have a positive correlation (i.e., points above the dashed line in @fig-cor-allometry-values). These are datasets that had a minor difference between the simulated effect of temperature on wing and mass, and by random chance the directional effect on body shape was the opposite intended.

```{r}
#| label: extract-sma-intercepts-function

extract_sma_intercepts <- function(df){
  ## Fit SMA models that include temperature to ensure that temp had desired effect on allometry
  # Keep the slope fixed
  mod_temp_bin <- run_sma_mod(df = df, interaction = FALSE)
    # Allow the slope to vary with binned temp
  mod_temp_bin_int <- run_sma_mod(df = df, interaction = TRUE)

  # Format coefficients
  mod_parms <- format_sma_parms(mod_temp_bin)
  mod_parms_int <- format_sma_parms(mod_temp_bin_int)
  
  # Return tibble with correlation between Temp increase & SMA intercept
  tibble(
    cor_allometry = cor(mod_parms$Temp_inc, mod_parms$elevation),
    cor_allometry_int = cor(mod_parms_int$Temp_inc, mod_parms_int$elevation)
  )
}
```

```{r}
#| label: extract-sma-intercepts

sma_intercepts <- map(df_morph_l, \(df){
  extract_sma_intercepts(df)
}) %>% list_rbind()
sma_intercepts2 <- bind_cols(Parms_mat2, sma_intercepts)
```

```{r}
#| label: fig-cor-allometry-values
#| fig-cap: Despite generating datasets where birds should become fatter or wingyier with increasing temperature (x-axis), when the strength of the difference in the simulated relationships is relatively weak the opposite relationship is observed in rare cases.

# Manipulate tbl for plotting the correlations between the intercepts from SMA models (body shape) and temperature increase
Parms_temp_bs <- sma_intercepts2 %>% filter(Scaling != "Negative") %>% 
  pivot_longer(cols = c(cor_allometry, cor_allometry_int), 
               names_to = "SMA_mod", values_to = "Correlation") %>%
  mutate(SMA_mod = if_else(
    SMA_mod == "cor_allometry", "No interaction", "Interaction")
  ) 
Parms_temp_bs %>%
  ggplot(aes(x = Temp_eff, y = Correlation)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') +
  geom_boxplot(outliers = FALSE) + 
  geom_point(position = position_jitter(width = .3), alpha = .3, size = 4,
             aes(color = Strength, shape = SMA_mod)) + #
  labs(x = NULL, 
       y = "Correlation temp increase\n and SMA intercepts", 
       title = "Effect of temperature on wingyness")
```

```{r}
#| label: extract-coefficients-general-function

# Workhorse function to fit models, extract residuals, & extract the estimated effect of temperature increase on relative appendage length
extract_coefs <- function(Sim_df) {
  
  ### Compare approaches 
  # Allometric (SMA) residuals approach
  sma_mod <- sma(Wing ~ Mass, data = Sim_df, method = sma_or_ma)
  est_b_sma <- coef(sma_mod)["slope"]
  
  # two-step residual regression 
  Ols_mod <- lm(Wing ~ Mass, data = Sim_df)
  resid_ols <- residuals(Ols_mod)
  ols_resid <- lm(resid_ols ~ Temp_inc, data = Sim_df)
  
  # Ryding
  ryding_app <- lm(Wing ~ Mass + Temp_inc, data = Sim_df)
  
  # Ratio
  Sim_df <- Sim_df %>% mutate(Wing_mass = Wing / Mass, 
                              Wing2_mass = Wing^2 / Mass)
  ratio_app <- lm(Wing_mass ~ Temp_inc, data = Sim_df) # Ratio Wing
  ratio2_app <- lm(Wing2_mass ~ Temp_inc, data = Sim_df) # Ratio Wing^2
  
  # SLI
  Sim_df <- Sim_df %>% 
    calc_sli(b_sli = 0.33, rename_col = "sli_isometry") %>% 
    calc_sli(b_sli = est_b_sma, rename_col = "sli_estimated")
  sli_iso_app <- lm(sli_isometry ~ Temp_inc, data = Sim_df)
  sli_est_app <- lm(sli_estimated ~ Temp_inc, data = Sim_df)
  
  tibble(
    coef_sli_iso_app = coef(sli_iso_app)["Temp_inc"],
    coef_sli_est_app = coef(sli_est_app)["Temp_inc"],
    coef_ols_resid_app = coef(ols_resid)["Temp_inc"],
    coef_ryding_app = coef(ryding_app)["Temp_inc"],
    coef_ratio = coef(ratio_app)["Temp_inc"],
    coef_ratio2 = coef(ratio2_app)["Temp_inc"], 
    est_b_sma = est_b_sma
  )
}
```

```{r}
#| label: Format-parms-tbl

# Next, apply the function & combine with the parameters that generated the data. 
Parms_tbl <- map(df_morph_l, extract_coefs) %>% list_rbind()
#Parms_mat_pos <- Parms_mat2 %>% filter(Scaling != "Negative") 
Parms_tbl2 <- bind_cols(Parms_mat2, Parms_tbl)
```

```{r}
#| label: Remove-incorrect-corr

Sim_fail <- Parms_tbl2 %>%
  left_join(sma_intercepts2) %>%
  filter(Temp_eff == "Fatter" & cor_allometry_int > 0 | Temp_eff == "Fatter" & cor_allometry > 0 | Temp_eff == "Wingyier" & cor_allometry_int < 0 | Temp_eff == "Wingyier" & cor_allometry < 0)
Parms_tbl3 <- Parms_tbl2 %>% anti_join(Sim_fail)
```

I remove these `r nrow(Sim_fail)` hypothetical species so the simulated data always matches the desired effect of simulated temperature.

```{r}
#| label: format-for-plotting

# Format the tibble for plotting 
# Pivot_longer to allow comparison between Ryding & the residual model 
Parms_tbl4 <- Parms_tbl3 %>% 
  pivot_longer(
    cols = c(coef_sli_iso_app, coef_sli_est_app, coef_ols_resid_app, coef_ryding_app, coef_ratio, coef_ratio2), 
               names_to = "Model", values_to = "b_temp_inc") %>% 
  mutate(Model = str_remove_all(Model, "coef_|_app"), 
         Model = str_to_sentence(Model))
```

Second, I simulated allometric scaling slopes of `r b_sma`, and -0.22. This plot shows the estimated SMA slopes from each hypothetical species.

```{r}
#| label: fig-SMA-slopes-maintained
#| fig-cap: Ensure that the SMA slopes set during simulation were recovered.

# Ensure that the SMA scaling relationships were maintained during simulation
if(!is.null(b_sma)){
  Parms_tbl4 %>% ggplot(aes(x = Scaling, y = est_b_sma, color = Scaling)) + 
    geom_boxplot(outliers = FALSE) + 
    geom_point(position = position_jitter(width = .1), color = "black", 
               alpha = .5) +  
    theme(legend.position = "top") + 
    labs(x = NULL, y = "Estimated SMA slope")
}
```

## Assumptions

I assessed OLS and SMA assumptions (e.g., linearity, independence, homoskedasticity, normality) using residuals vs. fitted plots and QQ plots. I just show a single example of a residuals vs fitted plot & a qq plot.

### Homoskedasticity & normality

#### SMA

```{r}
#| label: Test-assumptions-run-sma-mod

# SMA models list
sma_mod_l <- map(df_morph_l, \(df){
  sma_mod <- sma(Wing ~ Mass, data = df, method = sma_or_ma)
}) 
```

```{r}
#| label: SMA-assumptions-examples

# SMA 
plot(sma_mod_l[[1]], which = "residual")
plot(sma_mod_l[[1]], which = "qq")
```

#### OLS

```{r}
#| label: Test-assumptions-run-ols-mod

# OLS models list
ols_mod_l <- map(df_morph_l, \(df){
  ols_mod <- lm(Wing ~ Mass + Temp_inc, data = df)
})
```

```{r}
#| label: OLS-assumptions-examples
# OLS
plot(ols_mod_l[[1]], which = 1) # residuals vs fitted
plot(ols_mod_l[[1]], which = 2) # qq plot
```

```{r}
#| label: Test-assumptions-pdfs
#| message: false
#| echo: false

## Plot & save SMA assumption plots in a PDF
# Custom function to generate PDFs with plots to test assumptions
if(FALSE){
pdf_plot_assumptions <- function(name, model, plot.type){
pdf(paste0("Assumptions_check/", name, ".pdf"), width = 9, height = 9)
par(mfrow = c(3, 3))  # Arrange 3x3 plots per page
  

walk(model, \(mod) {
  plot(mod, which = plot.type)
})

dev.off()
} 

## Run custom function to produce PDFs testing assumptions for 
# SMA
pdf_plot_assumptions(name = "sma_resid_v_fit_plots", mod = sma_mod_l, plot.type = "residual")
pdf_plot_assumptions(name = "sma_qq_plots", mod = sma_mod_l, plot.type = "qq")

# OLS
pdf_plot_assumptions(name = "ols_resid_v_fit_plots", mod = ols_mod_l, plot.type = 1)
pdf_plot_assumptions(name = "ols_qq_plots", mod = ols_mod_l, plot.type = 2)
}
```

## Key question 1:

NOTE:: After examining all species (see PDFs), it seems there is some heteroskedasticity, where residual variance *increases toward the center* of the fitted values and *decreases toward the extremes.* The PDFs labeled sma_residual/qq_plots contains the full list of plots for all 180 datasets. Please take a quick look at the 'Assumptions_check' folder to visualize residuals vs fitted and qq plots for all 180 hypothetical species. Do you think this heteroskedasticity is problematic?

## SMA-specific assumptions

See Table 5 in Warton et al. @wartonBivariateLinefittingMethods2006 for a full list of assumptions of SMA regression.

### Assumption of errors

SMA regression assumes that measurement error variances in $X$ and $Y$ are proportional to trait variances in $X$ and $Y$. Given that we can't estimate the true errors in $X$ and $Y$ without repeated measurements, we follow @wartonBivariateLinefittingMethods2006 and @arnoldAllometricRelationshipSize2007 in using the ratio of empirical coefficients of variation of $X$ and $Y$ (i.e., $\lambda$), as proxies. $\lambda$ can be responsible for a large amount of variation that authors attribute to allometric scaling — for example, in a review of allometric scaling studies, Arnold and Green [-@arnoldAllometricRelationshipSize2007] found that 84% of the variation in the allometric slope estimates could be explained by the estimated $\lambda$. Thus, I compare $\lambda$ in the simulated datasets to those observed in our empirical nightjar dataset to ensure biological plausability.

## SLI: Standardized length index

Peig & Green [-@peigNewPerspectivesEstimating2009]; [-@peigParadigmBodyCondition2010] revolutionized the way we estimate body condition, using allometric scaling theory to create the standardized mass index [see 3 step process outlined on pages 1886 & 1887 of @peigNewPerspectivesEstimating2009]. We employ a similar process but instead create the standardized length index (SLI), where (appendage) length is on the y-axis. To calculate the SLI for individual $i$ of species $j$, first take the mean of the body size trait in question (e.g., mass) of species $j$, which serves to standardize against the appendage length of individual $i$. We call this $L_0$ as in @peigNewPerspectivesEstimating2009, \[[-@peigParadigmBodyCondition2010]\].\
\
$L_0 = mean(log(\mathit{Body\ size}))$

$SLI_i = \log(\text{Appendage}_i) * \dfrac{L_0}{\log(\text{Body size}_i)^{\beta_{SLI}}}$

Finally, one can model the SLI as a function of temperature increase.

$SLI \sim \mathit{Temperature\ increase}$

I propose two similar parameterizations of this model, differing only in the scaling slope $B_{SLI}$. One can estimate the functional relationship between $X$ and $Y$ empirically, or set the slope $B_{SLI}$ to a value with biological significance. For example, setting $B_{SLI} = 0.33$ compares the appendage length of individual $i$ to a hypothetical species exhibiting isometric scaling with the same body size as individual $i$. Note that setting $B_{SLI} = 0.33$ is similar to the approach taken by @youngfleshAbioticConditionsShape2022 in creation of their 'wingyness' index (although they had empirical support for comparing to isometry as well), as they explicitly wanted to capture departures from isometry.

# Results

## Simulated species

### Ratio of coefficients of variation

```{r}
#| label: coefficient-variation-function
#| echo: true

# Inspect function
calc_lambda

# CVs from simulated hypothetical species
lambdas_sim <- map_dbl(df_morph_l, \(df){ 
  df %>% summarize(lambda = calc_lambda(x = Mass, y = Wing)) %>% 
    pull(lambda) %>% 
    round(2)}
  ) 

#range(lambdas_sim)
#mean(lambdas_sim) 
#sd(lambdas_sim)
```

In our hypothetical species, $\lambda$ ranged from `r min(lambdas_sim)` - `r max(lambdas_sim)`, with a mean of `r round(mean(lambdas_sim), 2)`.

### Comparing approaches

The 5 approaches (2 SLI approaches, 3 traditional approaches) are all interested in accurately estimating $\beta_T$ from @eq-beta1-5-approaches.

$$
Wingyness \sim \alpha + \beta_T * \mathit{Temperature\ increase}
$$ {#eq-beta1-5-approaches}

The difference in the 5 approaches (note that Wing / mass and Wing² / mass are both 'ratio approaches') is how wingyness is measured (as outlined in the introduction). Note that the 'multivariate regression' approach would also contain $\beta_S * \mathit{Body\ size}$ in @eq-beta1-5-approaches.

### Positive allometry

```{r}
#| label: plot-approaches-fun

x_labs <- c("Sli_est" = "SLI estimated", 
            "Sli_iso" = "SLI isometry",
            "Ols_resid" = "OLS residuals",
            "Ryding" = "Mass as covariate",
            "Ratio2" = "Wing² / mass",
            "Ratio" = "Wing / mass")

## Plot relationship of interest -- 
plot_approaches <- function(df, negative = FALSE, ylim = TRUE, x_txt_size = 9, legend.pos = "right"){
  df <- df %>% mutate(r_12 = as_factor(r_12), 
                      Model = str_replace_all(Model, x_labs))
  if(isFALSE(negative)){df <- df %>% filter(Scaling != "Negative")}
  if(negative){ df <- df %>% filter(Scaling == "Negative") }
  plot <- df %>% 
  ggplot(aes(x = Model, y = b_temp_inc)) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_boxplot(position = position_dodge(width = .75), outlier.shape = NA) + 
  geom_point(position = position_jitterdodge(jitter.width = .9), 
             alpha = .6, size = 3,
             aes(color = Strength, shape = Scaling, #size = r_12,
                 group = Model)) + 
  labs(x = NULL, y = expression(beta[T] ~ "on wingyness")) + 
  facet_wrap(vars(Temp_eff)) + #, labeller = as_labeller(Facet_labs))
    theme(
      axis.text.x = element_text(size = x_txt_size, vjust = .58, angle = 60), 
      legend.position = legend.pos
    )
  if(ylim == TRUE){
    plot <- plot + ylim(c(-.25, .25))
  }
  return(plot)
}
```

```{r}
#| label: fig-compare-approaches
#| fig-cap: The true simulated effect above each plot, and the impact of temperature increase on wingyness on the y-axis. A positive slope means that birds get wingyer as temperature increases, and a negative slope means that birds get fatter as temperature increases. Each point represents a hypothetical dataset from a hypothetical species. Points are colored by the strength of the true effect in the simulated species, where lighter shades of blue mean a given increase in temperature results in a proportionally fatter (left panel) or wingyier (right panel) bird. The bird wingyness figure was borrowed from Youngflesh et al (2022), but will be changed before submission of the manuscript. 
#| fig-width: 11
#| fig-height: 8
#| fig-pos: 'H' 

# Plot
Compare_plot <- plot_approaches(Parms_tbl4, x_txt_size = 12, negative = FALSE, legend.pos = "top") #ylim = FALSE

# Load PNG, convert to a rasterGrob, shift upwards
img <- readPNG("Figures/Wingyness_fatty_spectrum.png")
img_grob <- rasterGrob(img, width = .9, height = .85)
img_shifted <- ggdraw() + draw_grob(img_grob, y = 0.08)

# Combine wingyness spectrum image with plot
plot_grid(
  img_shifted,
  Compare_plot,  
  ncol = 2,
  rel_widths = c(.4, 2)
)

# Save
#ggsave("Figures/Mod_comparison.png", bg = "white")
```

### Inverse allometry

```{r}
#| label: fig-negative-allometry
#| fig-cap: When species exhibit inverse allometry, a temperature increase results in species increasing in wing and decreasing in mass. These species are clearly much wingyier, yet several of the approaches (importantly, the 'SLI estimated' approach) fail to detect the increase in wingyness. Note that the the y-limits are truncated at -0.25, 0.25 to improve visibility, but the ratio methods have some extreme estimates in both the positive and negative directions.
#| fig-pos: 'H'

Parms_tbl4 %>%
  plot_approaches(negative = TRUE, ylim = TRUE)
```

## Evaluation - simulation

The ratio approaches generally return positively biased estimates of $\beta_T$, the multivariate regression and the OLS residuals approach return negatively biased estimates of $\beta_T$, whereas the SLI approaches correctly identify the relationship between temperature increase and relative appendage length in most cases.

```{r}
#| label: prep-eval-figs

# Determine direction of parameter estimate, & whether the model estimated the effect of temperature increase correctly
Parms_tbl5 <- Parms_tbl4 %>% 
  mutate(b_dir = if_else(b_temp_inc < 0, "Neg", "Pos")) %>%
  mutate(Est_correct = case_when(
    Temp_eff == "Much wingyier" & b_dir == "Pos" ~ TRUE, 
    Temp_eff == "Wingyier" & b_dir == "Pos" ~ TRUE, 
    Temp_eff == "Fatter" & b_dir == "Neg" ~ TRUE, 
    Temp_eff == "Proportional" ~ NA, 
    .default = FALSE
), .by = Model)

# Determine the proportion of correct classifications for each approach
Eval_tbl <- Parms_tbl5 %>% 
  summarize(Prop_correct = sum(Est_correct, na.rm = TRUE) / n(), 
            .by = c(Model, Temp_eff)) %>% 
  mutate(Prop_correct = round(Prop_correct, 2)) %>% 
  filter(Temp_eff != "Proportional") %>% 
  arrange(Model)

# Determine the proportion of positive (wingy) & negative (fatter) classifications when data sets were simulated as proportional
Proportional_methods_eval <- Parms_tbl5 %>%
  filter(Temp_eff == "Proportional") %>%
  summarize(Prop_pos = sum(b_dir == "Pos") / n(), .by = c(Model, Temp_eff)) %>%
  mutate(Prop_neg = 1 - Prop_pos) %>%
  pivot_longer(cols = c(Prop_pos, Prop_neg), names_to = "Direction", values_to = "Proportion") %>%
  mutate(Direction = str_remove(Direction, "Prop_"),
         Model = str_replace_all(Model, x_labs))

# Of the hypothetical species that were simulated as 'proportional', what % of the simulated datasets ended up being wingyer (positive) vs fatter (negative)?
Proportional_true_eff_tbl <- Parms_temp_bs %>% 
  filter(Temp_eff == "Proportional") %>% 
  mutate(True_temp_eff = if_else(Correlation > 0, "Wingyer", "Fatter")) %>% 
  tabyl(True_temp_eff)
Prop_fatter <- Proportional_true_eff_tbl %>% 
  filter(True_temp_eff == "Fatter") %>% 
  mutate(percent = round(percent, 2)) %>%
  pull(percent)
Prop_wingyer <- 1 - Prop_fatter
```

```{r}
#| label: fig-eval
#| fig-cap: Evaluation of methodological approaches in estimating the true simulated effects of temperature on body shape. In a) the proportion of hypothetical species that were correctly classified, where optimal model performance is a score of 1. In b) when species were simulated to decrease in wing and mass proportionally, hypothetical species get either wingyer or fatter randomly, and thus optimal methodological performance is approximately 0.5 (the black dashed line is the true proportion of hypothetical species that respond to temperature by getting fatter). Note there is no single statistical approach that performs perfectly across all hypothetical species. 
#| fig-width: 10
#| fig-height: 7.5
#| fig-pos: 'H'

# Create bar plot with the percent correctly classified for each temperature effect, similar to what I had for EWPW home ranges
Prop_correct_p <- Eval_tbl %>%
  mutate(Model = str_replace_all(Model, x_labs)) %>%
  ggplot(aes(x = Model, y = Prop_correct, fill = Temp_eff)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = NULL, y = "Proportion correctly classified", 
       fill = "True effect") +
  theme(axis.text.x = element_text(size = 10, vjust = .58, angle = 60),
        legend.position = "top")

# Separate panel with the proportional Pos & negative. The ideal model would classify about 50% of these as positive and 50% as negative
Pos_neg_p <- Proportional_methods_eval %>% 
  ggplot(aes(x = Model, y = Proportion, fill = fct_rev(Direction))) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = Prop_fatter, linetype = "dashed") +
  labs(x = NULL, y = "Proportion") +
  theme(axis.text.x = element_text(size = 10, vjust = .58, angle = 60),
        legend.position = "top") +
  scale_fill_discrete(name = NULL, labels = c("Wingyer", "Fatter"))

ggarrange(Prop_correct_p, Pos_neg_p, labels = "auto") 
```

```{r}
#| label: percent-right-wrong

### Generate the values to insert into the written text 

## Function to pull the desired proportion & convert to a percent using the Eval_tbl
pull_percent <- function(Model, Temp_eff){
  Eval_tbl %>% filter(Model == {{ Model }} & Temp_eff == {{ Temp_eff }}) %>% 
    mutate(per_corr = Prop_correct * 100) %>%
    pull(per_corr) 
} 

# Naming scheme of created objects: Method, Temp_eff, right_wrong
Sli.iso_fatter_wrong <- 100 - pull_percent("Sli_iso", "Fatter")
Sli_fatter_wrong <- Sli.iso_fatter_wrong / 2
Sli.est_much.wingyier_wrong <- 100 - pull_percent("Sli_est", "Much wingyier")
Ryding_much.wingyier_wrong <- 100 - pull_percent("Ryding", "Much wingyier")
Ryding_much.wingyier_right <- pull_percent("Ryding", "Much wingyier")
Ryding_wingyier_right <- pull_percent("Ryding", "Wingyier")

## Function to pull the desired proportion & convert to a percent using the Eval_tbl
pull_percent_proportional <- function(Model, Direction){
    Proportional_methods_eval %>% 
    filter(Model == {{ Model }} & Direction == {{ Direction }}) %>% 
    mutate(per_corr = round(Proportion * 100), 2) %>%
    pull(per_corr) 
}

# Proportion classified as wingy
Sli.iso_proportional_pos <- pull_percent_proportional("Sli_iso", "pos")
Sli.est_proportional_pos <- pull_percent_proportional("Sli_est", "pos")
# Difference between SLI classification & reality 
Dif_sli.iso <- Sli.iso_proportional_pos - (Prop_wingyer * 100)
Dif_sli.est <- Sli.est_proportional_pos - (Prop_wingyer * 100)

Dif_ryding <- pull_percent_proportional("Mass as covariate", "neg") - (Prop_fatter * 100)
```

### SLI approaches

The SLI approaches are the least biased overall, estimating the correct direction of the effect in most situations (@fig-compare-approaches). Both SLI approaches correctly estimate the direction of the simulated effect when temperature increases relative appendage length (i.e., a positive $\beta_T$ in @fig-compare-approaches; 'Wingyier' in the @fig-eval), but incorrectly estimates the direction of the simulated effect in about `r Sli_fatter_wrong`% of hypothetical species when temperature decreases relative appendage length (i.e., 'fatter' species). Both SLI approaches overestimate the increase in relative appendage length in species that were simulated to decrease in size proportionally ('SLI isometry' = `r Dif_sli.iso`% overestimation, 'SLI estimated' = `r Dif_sli.est`% overestimation; note that in these 'proportional' species about \~ 50% of species will get wingyer, `r Prop_wingyer`% in this simulation got wingyer, @fig-eval panel b). The SLI isometry approach correctly classified all species with an exaggerated response to warming (i.e., those exhibiting inverse allometry), whereas 'SLI estimated' approach misidentifies the relationship `r Sli.est_much.wingyier_wrong`% of the time ('Much wingyier' in the @fig-eval).

### Multivariate regression

Multivariate regression only detects the appropriate change in birds that are getting wingyier a little more than half the time (`r Ryding_much.wingyier_right`% and `r Ryding_wingyier_right`% for hypothetical species simulated to get 'Much wingyier' and 'Wingyier', respectively, in @fig-eval panel a). It misidentifies birds that we simulated to decrease in mass and wing proportionally as getting fatter with increasing temperatures (a `r Dif_ryding`% overestimation; @fig-eval panel b).

## Case study: Wing shape in migratory nightjars {#sec-case-study}

### Coefficient of variation {#sec-coef-var}

In the empirical nightjar data, $\lambda$ ranged from 0.08 - 0.16 (@tbl-cv-nightjars). This suggests that the variance in nightjar wings was low compared to the variance in mass (relative to the mean values), which could be due to higher measurement error in mass, [precision of wing is high in birds, @subasingheRepeatabilityValidityPhenotypic2021], or transient fluctuation error in mass. Importantly, $\lambda$ was similar in the 180 hypothetical species and the 3 nightjars we examined, providing additional evidence that our simulated datasets are biologically plausible. The simulated datasets cover a broader range of $\lambda$, indicating that these results may be relevant to additional evidence of appendage-mass datasets.

```{r}
#| label: tbl-cv-nightjars
#| tbl-cap: The ratio of coefficients of variation (lambda) in wing & mass in 3 nightjar species. Empirical data from Skinner et al. (2025).

cv_tab <- structure(list(Species = c("Nightjar", "Whip-poor-will", "Nighthawk"
), N = c(2416L, 535L, 321L), lambda = c(0.08, 0.16, 0.13)), row.names = c(NA, 
-3L), class = c("tbl_df", "tbl", "data.frame"))
cv_tab %>% gt() %>% 
  tab_options(latex.use_longtable = TRUE)
```

### Key question 2:

Arnold and Green [-@arnoldAllometricRelationshipSize2007] second McArdle's (1988) recommendation, that we should use SMA when 1 \< the ratio of errors $\lambda$ \< 3. In this case, do you think it’s still appropriate to use SMA regression? 

```{r}
#| label: fig-nightjar-shape-shifting
#| fig-cap: The approach used to quantify body shape greatly influences our biological conclusions. Points represent the mean parameter estimates and error bars represent 95% confidence interval (using the standard error). A positive slope means that birds get relatively wingyer as temperature increases, and a negative slope means that birds get relatively fatter as temperature increases. The same qualitative pattern is observed in our simulation study. 
#| fig-pos: 'H'

Nj_shape_img <- readPNG("Figures/Nightjar_shape.png")
Nj_shape_grob <- rasterGrob(Nj_shape_img, width = 1, height = 1)

ggdraw(Nj_shape_grob)
```

### Comparing approaches

I observed the same rank-order pattern in nightjars as in our simulation study: the ‘multivariate regression’ approach consistently yielded more negative estimates than the SLI methods, while the ratio approaches produced more positive estimates across all three species. Biological conclusions vary greatly depending on the approach used – all three species are getting wingyier according to the wing / mass ratio, whereas all three species are getting relatively fatter if using the multivariate regression or OLS residuals approaches. Given that our simulation study showed that the 'SLI isometry' approach was least biased, I conclude that, with warmer temperatures, nighthawks get wingyer, whereas nightjars whip-poor-wills get relatively fatter. In contrast, the SLI isometry approach, which was least biased in my simulations, implies divergent responses: nighthawks are getting wingyer (although the 95% barely overlaps 0) with increasing temperature, whereas nightjars and whip-poor-wills are getting relatively fatter. Importantly, the Wing/Mass ratio was strongly correlated with mass in all three species ($r < -0.9$), which likely drives its positive bias as body size declines with increasing temperatures (i.e., Bergmann’s rule). Squaring wing length before dividing by mass (Wing²/Mass) reduced this correlation ($-0.81 < r < -0.66$), suggesting it may be a more effective, albeit still imperfect, proxy for shape.

# Discussion

The 'SLI isometry' approach appears the most flexible and generalizeable of all the approaches we examined, but is slightly positively biased in situations that are likely common in the real world (e.g., that species decrease in size proportionally with climate warming). The multivariate regression approach returned negatively biased estimates of temperature's impact on wingyness, whereas the ratio approach appeared positively biased. There are at least two things that I believe are causing the SLI approaches to outperform the classical regression approaches...

From our case study of nightjars, is not surprising that there are large positive effects of $\beta_T$ on wingyness (given that all three species follow Bergmann's rule). Wing^2^ / mass also has more theoretical support as a proxy for the surface area to volume ratio (a squared to a cubic ratio).

The 'multivariate regression' approach has been advocated for [@rydingResponseAllometryEvaluate2022, CITE Ryding et al 2025] by authors in recent years, and was the most commonly utilized approach to quantify relative appendage length in a concurrent literature review [e.g., @rydingLongShorttermResponses2024; @mcqueenThermalAdaptationBest2022; Skinner et al in prep]. However, I found that it returns negatively biased estimates of temperature's impact on wingyness ($\beta_T$ in @eq-beta1-5-approaches), to where:

## 1. Absolute appendage length

The 'multivariate regression' approach claims that by including a metric of body size as a predictor, we are now measuring the impact of temperature on the proportional change in size of the response variable (i.e., wing length in this case). But in reality we are still measuring the impact of body size and temperature on absolute (not relative) appendage length. Let's examine this more formally from a causal modeling framework. When both Bergmann's and Allen's rules apply, we expect temperature to have causal effects on both appendage length and body size. While most biologists would agree that body size does not *cause* appendage length, the model $Appendage \sim \beta_T * \mathit{Temperature\ increase} + \beta_S * Size$implicitly treats body size as a causal predictor of appendage length. This corresponds to the following directed acyclic graph (DAG):

```{r}
#| label: fig-DAG
#| fig-cap: The directed acyclic graph (DAG) implied by the 'multivariate regression' approach. 
#| fig-pos: 'H'

library(dagitty)
library(ggdag)

# Specify DAG
dag_allen <- dagify(
  Appendage ~ Size + Temperature,
  Size ~ Temperature,
  exposure = c("Temperature"), 
  outcome = "Appendage",
  labels = c("Temp" = "Temperature")
)

# More complex DAG acknowledging latent variable size
# Error terms (not included explicitly): measurement error, transient biological error
dag_allen2 <- dagify(
  Mass ~ Size + Temp,
  Appendage ~ Size + Temp,
  Size ~ Temp,
  latent = "Size",
  exposure = "Temp",
  outcome = "Appendage",
  labels = c("Temp" = "Temperature", "Size" = "Latent Body Size")
)

# Adjustment set 
# adjustmentSets(dag_allen, type = "minimal", effect = "direct") 

# Plot
tidy_dagitty(dag_allen, layout = "fr") %>% 
  ggdag_status(text_col = "black",
               text = TRUE, 
               edge_type = "link_arc",
               node_size = 20,
               text_size = 3, 
               stylized = TRUE) + 
  theme_dag() + 
  guides(fill = "none", color = "none")
```

If our goal is to estimate the effect of temperature on relative appendage length, we hope that $\beta_T$ captures that. But conditioning on size removes the indirect path from temperature to appendage length that operates via changes in body size. The result is that $\beta_T$ estimates the direct effect of temperature on absolute appendage length, not relative shape. So when both Bergmann’s and Allen’s rules apply, i.e. when increasing temperatures reduce both body size and appendage length, this model often yields a negative estimate of $\beta_T$, even in cases where appendages become relatively longer (species below the dashed line in the 'wingyier' panel in @fig-compare-approaches). The sign and magnitude of $\beta_T$ will depend on the relative strength of the two causal paths, $Temperature \rightarrow Appendage$ and $Temperature \rightarrow Size$ (‘Strengths’ in @fig-compare-approaches).

## 2. SMA \> classical regression (at least in this simulation)

The other thing that is likely occurring is that classical regression approaches are not the appropriate type of regression for these simulated datasets (which I believe are biologically realistic). I simulated equivalent amounts of measurement error in $X$ and $Y$, and transient fluctuation error in just $X$, given that length is unlikely to change over short time periods. Thus, there is more error in $X$ than in $Y$, yet the classical regression approaches assume all error is in $Y$.

```{r}
#| label: plot_allo_temp-function

plot_allo_temp <- function(df, label = NULL, alpha = .7, ...){
  p <- ggplot(data = df, aes(x = Mass, y = Wing)) + 
  geom_smooth(method = "lm", color = "red") +
  ggpmisc::stat_ma_line(method = sma_or_ma, color = "blue") +
  geom_point(alpha = alpha, aes(color = Temp_inc)) +
  labs(x = "Log (size)", y = "Log (appendage)", color = "Temperature\n increase", ...) + 
  lims(y = c(xy_lims$min.wing, xy_lims$max.wing), 
       x = c(xy_lims$min.mass, xy_lims$max.mass))
  
  if(!is.null(label)){
    # x_pos <- quartile(Mass, .05, 1) # relative position
    p <- p + annotate("text", x = .05, y = 2.4, label = "I", size = 6, color = "black") +
  annotate("text", x = 6.0, y = 3.8, label = "II", size = 6, color = "black")
  }
  
  return(p)
}
```

```{r}
#| label: Generate-ex-hypo-hyper

Ex_parms_hypo_hyper <- Parms_mat2 %>% 
  filter(Scaling != "Isometry" & Temp_eff == "Wingyier" & r_12 == .3) %>%
  slice_max(n = 1, order_by = Strength, by = c(Temp_eff, Scaling), 
            with_ties = FALSE)
Ex_df_hypo_hyper <- gen_ex_data(Ex_parms_hypo_hyper) %>% 
  mutate(Temp_inc = Temp_inc / 3)
Ex_df_hypo_hyper_l <- Ex_df_hypo_hyper %>% group_split(Scaling)
```

```{r}
#| label: fig-allometric-relationship-temp-wingy
#| fig-cap: The allometric relationships of a single hypothetical species (i.e., each point is an individual bird) that we simulated to get wingyier as temperature increases. The SMA line of best fit is shown in blue, and the OLS regression line in red, and the points are colored by how much temperature has increased for a given individual of the hypothetical species. 
#| fig-pos: 'H'

# Define common set of X & Y limits 
xy_lims <- Ex_df_hypo_hyper %>% 
  summarize(max.wing = max(Wing), 
            min.wing = min(Wing), 
            max.mass = max(Mass), 
            min.mass = min(Mass))

# Plot 
hypo_p <- plot_allo_temp(df = Ex_df_hypo_hyper_l[[1]],
                         title = "Hypoallometric")
hyper_p <- plot_allo_temp(df = Ex_df_hypo_hyper_l[[2]], label = TRUE) #, 
                          #title = "Hyperallometric")
#ggarrange(hypo_p, hyper_p, common.legend = TRUE)

hyper_p
#ggsave("Figures/fig-underestimation-ral.png", bg = "white")
```

While the approach advocated for by @rydingResponseAllometryEvaluate2022 does not leverage the residuals from the line of best fit from classical regression, it does use an OLS line-fitting approach where wing size is controlled for by mass. Thus, the above plot is useful in developing some intuition for why the SLI approach does a better job at detecting the true relationship between an increase in temperature & the change in proportional wing size. Classical regressionn always underestimates the slope of the true relationship between mass and wing. This leads to an overestimation of negative residuals in small individuals (i.e., points in the bottom left quadrant of the plot - labeled with a black "I"), and an overestimation of positive residuals in large individuals (top right quadrant of the plot - labeled with a black "II"). If we assume that the temporal analogs of Allen's and Bergmann's rule apply to most species, then temperature increases will push individuals towards the bottom left quadrant of the plot – exactly where classical regression overestimates the negative residuals. In other words, classical regression approaches suggest that the individuals in Section I are relatively fat (i.e., negative residuals), whereas SMA labels them as relatively wingy. We see the inverse bias in Section II of the plot.

# When is the SLI approach NOT the best choice?

There are cases when including covariates in an OLS model may be required, or may perform better than using the residuals from SMA regression as our metric of shape.

## Negative allometric scaling

As noted previously, when there is negative allometric scaling, i.e., wing length increases when mass decreases, the 'SLI estimated' approach often does not identify the increase in wingyness (@fig-negative-allometry). Surprisingly, **only the 'SLI isometry' approach returns the correct estimated direction of effect in all cases,** and the ratio approaches return the correct direction of the simulated effect about 85% of the time. Surprisingly, the multivariate regression and the OLS residuals don't perform very well either, as they return negatively biased estimates of $\beta_T$ in @eq-beta1-5-approaches in many of the 180 hypothetical species. We believe this is due to a 'supression' effect resulting from collider bias (CITE causal modeling literature). Collider bias results from conditioning on a negatively correlated mediator like body size in this case). In short, the positive effect of Temp_increase → Relative appendage is overwhelmed by the path Temp_increase → Body size → Relative appendage, where a temperature increase decreases body size, and body size and appendage length are negatively correlated (i.e., there is inverse allometry). inverse allometry is not particularly common in nature, although has been observed in rare cases [e.g. @weeksSharedMorphologicalConsequences2020]. This suppression effect was observed in about `r Ryding_much.wingyier_wrong`% of hypothetical species', so more research is needed to determine when multivariate regression may be appropriate.

```{r}
#| label: Parm-collider-bias
#| echo: false
#| eval: false

Parms_tbl5 %>% filter(Temp_eff == "Much wingyier" & Est_correct == FALSE) %>% 
  dplyr::select(b_sma_12, r_12, r_13, r_23, Strength) %>% 
  distinct() 
```

```{r}
#| label: fig-neg-allometric-temp_wingy
#| fig-cap: Hypothetical species exhibiting inverse allometry. The 'SLI estimated' approach does not capture the relationship between an increase in temperature and relative appendage length. 

Ex_parms_neg <- Parms_mat2 %>% 
  filter(Scaling == "Negative" & Strength == 0.8) %>%
  slice_max(n = 1, order_by = Strength, by = Temp_eff, with_ties = FALSE)
Ex_df_neg <- gen_ex_data(Ex_parms_neg)

plot_allo_temp(df = Ex_df_neg)
```

## Violation of assumptions in SMA regression

The SLI approaches are inflexible in their ability to customize models. This is because SLI approaches can only model the two scaling variables $X$ and $Y$ (at least in R packages that can conduct SMA regression, e.g. `smatr`).

### Non-linearity between X and Y

For example, if after log-log transforming the linear & volumetric measurement there is still not **a linear relationship**, the SMA assumption of linearity (that $Y$ & $X$ are linearly related) is violated, and it would be prudent to compare linear vs non-linear (e.g., $\mathit{Body\ size} + \mathit{Body\ size^2}$) models, as was done in Baldwin et al. [-@baldwinComplementarityAllensBergmanns2023]. In this case the 'covariate in OLS regression' approach may be preferred.

### Heteroskedasticity due to missing covariates

A second case where it may be advisable to use the 'covariate in OLS regression' approach is if your SMA model shows severe heteroskedasticity due to missing covariates. For example, let's say we are trying to understand shape-shifting in a bird species that moves through thick underbrush and experiences severe flight feather wear throughout the year after its annual molt. You may want to control for this flight feather wear by including the date of capture in our model examining wingyness, but this isn't possible in SMA regression, i.e. $Appendage \sim \mathit{Temperature\ increase} + Date + \mathit{Body\ size}$ is not a feasible model in SMA regression. Including date of capture (or feather wear directly) directly in the SLI models ($SLI \sim \mathit{Temperature\ increase} + Date$) is an option, but more study is needed.

# Final recommendations

As suggested by previous authors [@wartonBivariateLinefittingMethods2006; @greenMassLengthResiduals2001], SMA regression can improve our inference in studies involving biological scaling. SMA assumes that the error variance in $X$ and $Y$ are equivalent to the variance in $X$ and $Y$, which makes more biological sense than assuming all error is in the $Y$ variable, as in classical regression approaches. However, the assumption of equal error variance may not always be reasonable either, and can also lead to biased allometric slopes [CITE SMITH 2009, @arnoldAllometricRelationshipSize2007; @wartonBivariateLinefittingMethods2006]. Another option is to model true body size as a latent trait $X_{true}$ that is not directly observed, and is made of up of $X$ plus some error term $\sigma_x$. The error terms from both transient fluctuations and measurement error can be partitioned with repeat measurements at the appropriate temporal scale (Skinner in prep). This can ameliorate issues due to error in $X$ in classical regression approaches, and these models can be directly extended to include additional biological complexity (e.g., phylogenetic or random effects, spatial or temporal autocorrelation, etc.). Importantly, the multivariate regression approach may still be plagued by the 'supression' effect even when modeling error in $X$.

However, repeat morphological measurements are rarely collected in ecology, and modeling $X_{true}$ as latent may be beyond the statistical understanding or time constraints of many people. The SLI isometry approach, while returning estimates of $\beta_T$ that were slightly positively biased in our study, provides a generalizeable and easily-implementable metric of relative appendage length. In just a few lines of code, users can have a biologically principled and tested estimate of body shape.

The biological world is complex, and our statistical approaches to handle this complexity have improved greatly in recent decades. For example, in a concurrent literature review, I found that most studies in recent years examined Allen's rule across groups of species, and that models accounting for phylogenetic relationships (e.g., PGLMM) were the most frequently employed model. I highlight that the SLI isometry approach could easily be generated for any number of species, and the output could then be used as the response variable in a model accounting for phylogenetic relatedness.

# Acknowledgements

I thank Asad Hassan for his help in conceptualizing an unbiased simulation, and Andy Green for helpful comments on the SLI as well as earlier drafts of this article.

# Ignore

No need to continue on... Thanks for reading!

# Still to do

-   Understand how allometric scaling influences response in the different approaches. When correlation between mass & temp is \< -.4, hyperallometric individuals show the strongest response, but when correlation = -.7 it is much more variable – why?

    -   Returning to the faceted plot comparing a hyper- & hypo- allometric species (@fig-allometric-relationship-temp-wingy), we can also see that hyperallometric species have the strongest response, irrespective of whether an increase in temperature makes organisms fatter or wingyier. While it's true that a given change in wing or mass will produce a larger change in the other variable due to the steeper slope exhibited in hyperallometric species, I thought fitting the SMA line of best fit would largely remove the effect of scaling. Thus, I thought the SMA residuals approach would show similar results irrespective of the simulated scaling relationship.

        -   The pattern of stronger residual–temperature effects in hyperallometric species is likely a **consequence** of how covariances are structured in this simulation. I don't think that this reflects a true biological phenomena, but a *mathematical property* of the multivariate normal distribution under the simulated covariances.

        -   Think of each species’ cloud of points like a cigar. The SMA line goes along the long axis of the cigar. But if an increase in temperature “pushes” the whole cigar diagonally — and more so for hyperallometric species — then the **residuals (shortest distance to the** SMA line of best fit**)** will stretch out more along the temperature gradient. That’s why we see a stronger correlation between temp & the residuals (wingyness) in a hyperallometric species.

    -   Steeper SMA line -\> greater residuals? Correlation & b_ols slope..

-   Run with MA instead of SMA

    -   SMA is essentially like running scale(log_wing) \~ scale(log_mass) and then backtransforming

    -   MA regression does change the results, it seems negatively biased along with the Ryding approach, but several sources suggest that SMA regression is the approach that makes the most sense (Warton, 2006; Peig & Green, 2009). So don't worry about MA for now.

-   Case study

    -   Include 2 SMA lines of best fit in real data and compare to mass + sex (or age) + temp increase

    -   Create Temp & body shape plot

    -   Check residual vs fitted, qq plots to ensure we meet assumptions of SMA regression

-   How will we evaluate these different methods? Could AUC-ROC be viable? Note an AUC-ROC of 0.5 is expected by chance

## To consider:

There are certainly more complete & time-intensive approaches for measuring shape in ecology e.g.,

-   Landmark analysis (Bookstein, 2013).

-   Measurements of bones

-   Functional trait space where each point is an individual (I haven't seen this in the literature though)

How can we make a size index? How does this differ from the SMI? Do we want a shape index or a size index?

-   Compare this approach to the SMI approach... Are they perfectly correlated?

Box-cox to understand the parameter of the Box-Cox power transformation?

```{r}
#boxcox(lm(Wing ~ Mass, data = Ex_df_wingy))
```

# Extras

## Varied scaling and correlation hold all else constant

It is confusing to me that a species can be hyperallometric (wing decreases quickly with an decrease in mass) and simultaneously get wingyier with an increase in temperature (relative wing increases with an increase in temperature).

```{r}
#| label: fig-wingy-varied-scaling
#| fig-cap: In each row the SMA slope is constant and the OLS slope increases with increasing correlation. Increasing correlations result in tighter spreads around the lines of best fit, whereas increasing SMA slopes shifts the orientation of the cloud of points (hypoallometry is more horizontal, hyperallometry is more vertical). Temperature....
#| fig-width: 11
#| fig-height: 14

# Temp reduces mass more than wing
vary_scaling_parms <- Parms_mat2 %>% 
  filter(Temp_eff == "Wingyier") %>%
  slice_max(n = 1, order_by = tibble(Strength), 
            by = c(Scaling, r_12), with_ties = FALSE)
vary_scaling_parms

# Generate data
vary_scaling_tbl <- gen_ex_data(vary_scaling_parms) %>% 
  mutate(Temp_inc = Temp_inc / 3) 
# Separate into list
vary_scaling_l <- vary_scaling_tbl %>% group_split(Scaling, r_12)
# Name list
names(vary_scaling_l) <- vary_scaling_tbl %>% distinct(Scaling, r_12) %>% 
  mutate(name = paste(Scaling, r_12, sep = "_")) %>% 
  pull(name)

# Create plots
vary_scale_p <- imap(vary_scaling_l, \(df, name){
  plot_allo_temp(df = df, title = name, alpha = .5)
})

# 3x3 grid
ggarrange(
  plotlist = vary_scale_p[1:9],
  nrow = 3,
  ncol = 3,
  common.legend = TRUE
)
```

## Extra text

While some authors have used raw appendage size to test Allen's rule [e.g., @fanBergmannsRuleAllens2019], I would argue that most appendages this is a better test of Bergmann's rule in most cases [although organs tightly tied to thermoregulation may be an exception for some species and contexts - bird bills: @tattersallEvolutionAvianBill2017; @symondsGeographicalVariationBill2010; elephant ears: CITE]. In studies of Allen's rule, I argue that tying multiple metrics together at the level of the individual is critical.

Allen's rule is often interpreted as appendage length *relative* to body size, which acknowledges that overall body size and appendage length will generally show positive covariance due to principles of biological scaling [@shingletonAllometryStudyBiological2010]. In fact, interpreting Allen's rule as relative appendage length is the only way that Bergmann's and Allen's rules can be true simultaneously for a species exhibiting positive allometry (i.e., body size and appendage size positively covary). For the rest of this section, I will refer to body size as mass (although as noted above there are many metrics that can approximate body size), as this is the metric commonly used to represent body size in studies of Allen's rule. Positive allometry is very common in nature, as larger organisms tend to weigh more and require longer appendages to transport this increased weight. For example, lift in bird wings is FILL IN and thus the laws of physics tend to result in species with positive allometry [@raynerFormFunctionAvian1988].

To examine *relative* appendage length, authors must compare a given appendage relative to a metric of body size for each individual. This has been done in several ways, including (1) modeling simple ratios (e.g., appendage:mass) as a function of temperature, (2) modeling appendage length as a function of body size and temperature, and (3) conducting a two-step regression by (3a) modeling appendage length as a function of mass using ordinary least squares (OLS) regression, and then (3b) modeling the residuals as a function of temperature. In the next section we discuss these three approaches and make our case as to why we believe t

{{< pagebreak >}}

# References
